---
title: "RNA Sequencing - Oysters"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
---

**00 - File Check**

```{r}
# ============================================================
# Environmental Regression Analysis
# Oyster RNA-seq Project
#
# Purpose:
# Test whether environmental conditions are associated with
# WGCNA module eigengenes and selected gene-expression patterns.
#
# Analysis order:
# 1. Prepare sample-level environmental data
# 2. Explore environmental variables
# 3. Run module-level linear regressions
# 4. Identify the strongest module-environment associations
# 5. Examine selected genes using VST-normalized expression
# ============================================================


# Check the current project folder
getwd()

# View the files available in the current folder
list.files()
```

**01 - Environmental Data Matrix**

```{r}

#Define seven day window as
  # collection date minus 6 days through collection date

# ============================================================
# 2. Inspect environmental source files
# ============================================================

library(readxl)
library(dplyr)
library(lubridate)
library(tidyr)
library(stringr)
library(purrr)

# The Cleaned folder is at the root of the GitHub repository
cleaned_dir <- file.path("..", "..", "..", "Cleaned")

# Check that R can find the folder
dir.exists(cleaned_dir)
```

```{r}

# Environmental files expected for the 2024 RNA-seq samples
environmental_files <- c(
  salinity = "Salinity_2024.xlsx",
  dissolved_oxygen = "DOsensor_2024.xlsx",
  precipitation = "Precipdaily_2024.xlsx",
  pH = "pH_2024.xlsx"
)

# Construct full file paths
environmental_paths <- file.path(
  cleaned_dir,
  unname(environmental_files)
)

# Confirm that every file exists
file_check <- data.frame(
  dataset = names(environmental_files),
  filename = unname(environmental_files),
  exists = file.exists(environmental_paths)
)

file_check
```

```{r}
all(file_check$exists)
```

```{r}
# ============================================================
# 2B. Inspect sample metadata and workbook structures
# ============================================================

# Load the existing sample-level table.
# We are using this for sample IDs and collection information,
# not trusting its environmental summaries yet.
existing_sample_table <- read.csv(
  "environmental_traits_by_sample.csv",
  stringsAsFactors = FALSE,
  check.names = FALSE
)

# Number of rows and columns
dim(existing_sample_table)

# Exact column names
names(existing_sample_table)

# First six samples
head(existing_sample_table, 6)
```

```{r}
# Give the paths names again so the output is easy to interpret
names(environmental_paths) <- names(environmental_files)

workbook_sheets <- lapply(
  environmental_paths,
  readxl::excel_sheets
)

workbook_sheets
```

```{r}
workbook_previews <- lapply(
  environmental_paths,
  function(path) {
    first_sheet <- readxl::excel_sheets(path)[1]

    readxl::read_excel(
      path,
      sheet = first_sheet,
      n_max = 5
    )
  }
)

workbook_previews

```

```{r}
# ============================================================
# 2C. Correct the sample-to-station mapping
# ============================================================

sample_metadata <- existing_sample_table %>%
  transmute(
    sample_id,
    location,
    month,
    oyster,
    date_collected = as.Date(date_collected),

    # Correct station mapping
    station = case_when(
      location == "CMT" ~ "CMAST",
      location == "DAF" ~ "DUML",
      TRUE ~ NA_character_
    )
  )

sample_metadata
```

```{r}

#Mapping check and recreating the collection events

sample_metadata %>%
  distinct(location, station)

stopifnot(
  all(sample_metadata$location == "CMT" |
        sample_metadata$location == "DAF"),
  all(!is.na(sample_metadata$station))
)

collection_events <- sample_metadata %>%
  distinct(
    location,
    station,
    date_collected,
    month
  ) %>%
  arrange(date_collected)

collection_events
```

```{r}
# ============================================================
# 2D. Load raw environmental data and create daily summaries
# ============================================================

# Safely calculate a mean without producing NaN
safe_mean <- function(x) {
  if (all(is.na(x))) {
    return(NA_real_)
  }

  mean(x, na.rm = TRUE)
}

# Safely calculate a sum without turning an entirely missing day into zero
safe_sum <- function(x) {
  if (all(is.na(x))) {
    return(NA_real_)
  }

  sum(x, na.rm = TRUE)
}

# Handles DateTime columns whether readxl imports them as
# text or as an existing date-time object
parse_environment_datetime <- function(x) {

  if (inherits(x, "POSIXt")) {
    return(lubridate::with_tz(x, "America/New_York"))
  }

  lubridate::ymd_hms(
    as.character(x),
    tz = "America/New_York",
    quiet = TRUE
  )
}
```

```{r}
# Temperature, conductivity, and salinity
salinity_raw <- readxl::read_excel(
  environmental_paths["salinity"],
  sheet = "Sheet1"
) %>%
  mutate(
    DateTime = parse_environment_datetime(DateTime),
    date = as.Date(DateTime),
    Farm = stringr::str_to_upper(stringr::str_trim(Farm))
  )

# Dissolved oxygen
do_raw <- readxl::read_excel(
  environmental_paths["dissolved_oxygen"],
  sheet = "DOUpdated"
) %>%
  mutate(
    DateTime = parse_environment_datetime(DateTime),
    date = as.Date(DateTime),
    Farm = stringr::str_to_upper(stringr::str_trim(Farm))
  )

# Precipitation
precip_raw <- readxl::read_excel(
  environmental_paths["precipitation"],
  sheet = "Sheet1"
) %>%
  mutate(
    DateTime = parse_environment_datetime(DateTime),
    date = as.Date(DateTime)
  )

# pH
ph_raw <- readxl::read_excel(
  environmental_paths["pH"],
  sheet = "pHUpdated"
) %>%
  mutate(
    DateTime = parse_environment_datetime(DateTime),
    date = as.Date(DateTime),
    Farm = stringr::str_to_upper(stringr::str_trim(Farm))
  )
```

```{r}
datetime_check <- tibble::tibble(
  dataset = c(
    "salinity",
    "dissolved_oxygen",
    "precipitation",
    "pH"
  ),
  total_rows = c(
    nrow(salinity_raw),
    nrow(do_raw),
    nrow(precip_raw),
    nrow(ph_raw)
  ),
  missing_datetime = c(
    sum(is.na(salinity_raw$DateTime)),
    sum(is.na(do_raw$DateTime)),
    sum(is.na(precip_raw$DateTime)),
    sum(is.na(ph_raw$DateTime))
  )
)

datetime_check
```

```{r}
# Reload DO data while preserving the original DateTime values
do_datetime_check <- readxl::read_excel(
  environmental_paths["dissolved_oxygen"],
  sheet = "DOUpdated"
) %>%
  mutate(
    source_row = dplyr::row_number(),
    DateTime_original = as.character(DateTime),
    DateTime_parsed = parse_environment_datetime(DateTime)
  )

# Display every row whose timestamp could not be parsed
do_datetime_problems <- do_datetime_check %>%
  filter(is.na(DateTime_parsed))

print(
  do_datetime_problems,
  n = Inf,
  width = Inf
)
```

```{r}
do_datetime_problems %>%
  summarise(
    problem_rows = n(),
    rows_with_DO_values = sum(!is.na(DomgL)),
    rows_with_farm_values = sum(!is.na(Farm)),
    rows_with_temperature_values = sum(!is.na(Temp_C))
  )
```

```{r}
# Remove four empty spreadsheet artifact rows
do_raw <- do_raw %>%
  filter(!is.na(DateTime))
```

```{r}
# Four rows with invalid Excel placeholder dates and no measurements
# were removed before daily aggregation.

nrow(do_raw)
sum(is.na(do_raw$DateTime))
```

```{r}
do_daily <- do_raw %>%
  filter(Farm %in% c("CMAST", "DUML")) %>%
  group_by(Farm, date) %>%
  summarise(
    DO_daily_mean = safe_mean(DomgL),
    DO_observations = sum(!is.na(DomgL)),
    .groups = "drop"
  )
```

```{r}
# ============================================================
# 2E. Create daily environmental summaries
# ============================================================

# Temperature, conductivity, and salinity
salinity_daily <- salinity_raw %>%
  filter(
    Farm %in% c("CMAST", "DUML"),
    !is.na(date)
  ) %>%
  group_by(Farm, date) %>%
  summarise(
    temperature_daily_mean = safe_mean(Temp_C),
    conductivity_daily_mean = safe_mean(ConduScm),
    salinity_daily_mean = safe_mean(Sal_ppt),

    temperature_observations = sum(!is.na(Temp_C)),
    conductivity_observations = sum(!is.na(ConduScm)),
    salinity_observations = sum(!is.na(Sal_ppt)),

    .groups = "drop"
  ) %>%
  arrange(Farm, date)


# Dissolved oxygen
do_daily <- do_raw %>%
  filter(
    Farm %in% c("CMAST", "DUML"),
    !is.na(date)
  ) %>%
  group_by(Farm, date) %>%
  summarise(
    DO_daily_mean = safe_mean(DomgL),
    DO_observations = sum(!is.na(DomgL)),
    .groups = "drop"
  ) %>%
  arrange(Farm, date)


# pH
ph_daily <- ph_raw %>%
  filter(
    Farm %in% c("CMAST", "DUML"),
    !is.na(date)
  ) %>%
  group_by(Farm, date) %>%
  summarise(
    pH_daily_mean = safe_mean(pH),
    pH_observations = sum(!is.na(pH)),
    .groups = "drop"
  ) %>%
  arrange(Farm, date)


# Precipitation is already daily.
# Summing protects us in case a date appears more than once.
precip_daily <- precip_raw %>%
  filter(!is.na(date)) %>%
  group_by(date) %>%
  summarise(
    precipitation_daily_mm = safe_sum(Precip_mm),
    precipitation_observations = sum(!is.na(Precip_mm)),
    .groups = "drop"
  ) %>%
  arrange(date)
```

```{r}
daily_summary_check <- tibble::tibble(
  dataset = c(
    "temperature_salinity_conductivity",
    "dissolved_oxygen",
    "pH",
    "precipitation"
  ),
  rows = c(
    nrow(salinity_daily),
    nrow(do_daily),
    nrow(ph_daily),
    nrow(precip_daily)
  ),
  missing_summary_values = c(
    sum(
      is.na(salinity_daily$temperature_daily_mean) |
      is.na(salinity_daily$conductivity_daily_mean) |
      is.na(salinity_daily$salinity_daily_mean)
    ),
    sum(is.na(do_daily$DO_daily_mean)),
    sum(is.na(ph_daily$pH_daily_mean)),
    sum(is.na(precip_daily$precipitation_daily_mm))
  )
)

daily_summary_check
```

```{r}
salinity_daily %>%
  group_by(Farm) %>%
  summarise(
    first_date = min(date),
    last_date = max(date),
    days_available = n_distinct(date),
    .groups = "drop"
  )

do_daily %>%
  group_by(Farm) %>%
  summarise(
    first_date = min(date),
    last_date = max(date),
    days_available = n_distinct(date),
    .groups = "drop"
  )

ph_daily %>%
  group_by(Farm) %>%
  summarise(
    first_date = min(date),
    last_date = max(date),
    days_available = n_distinct(date),
    .groups = "drop"
  )

precip_daily %>%
  summarise(
    first_date = min(date),
    last_date = max(date),
    days_available = n_distinct(date)
  )
```

```{r}
# ============================================================
# 2F. Check environmental coverage for each 7-day window
# ============================================================

# Create one row for every required calendar day
# Collection day and the six preceding days = 7 total days
event_windows <- collection_events %>%
  mutate(
    window_start = date_collected - lubridate::days(6),
    window_end = date_collected
  ) %>%
  rowwise() %>%
  mutate(
    date = list(
      seq(
        from = window_start,
        to = window_end,
        by = "day"
      )
    )
  ) %>%
  ungroup() %>%
  tidyr::unnest(date)

event_windows
```

```{r}
dim(event_windows)
```

```{r}
# Rename Farm to station before joining
salinity_daily_join <- salinity_daily %>%
  rename(station = Farm)

do_daily_join <- do_daily %>%
  rename(station = Farm)

ph_daily_join <- ph_daily %>%
  rename(station = Farm)

# Join each required date to its environmental measurements
event_window_data <- event_windows %>%
  left_join(
    salinity_daily_join,
    by = c("station", "date")
  ) %>%
  left_join(
    do_daily_join,
    by = c("station", "date")
  ) %>%
  left_join(
    ph_daily_join,
    by = c("station", "date")
  ) %>%
  left_join(
    precip_daily,
    by = "date"
  )

event_window_data
```

```{r}
seven_day_coverage <- event_window_data %>%
  group_by(
    location,
    station,
    month,
    date_collected,
    window_start,
    window_end
  ) %>%
  summarise(
    expected_days = n_distinct(date),

    temperature_days =
      sum(!is.na(temperature_daily_mean)),

    conductivity_days =
      sum(!is.na(conductivity_daily_mean)),

    salinity_days =
      sum(!is.na(salinity_daily_mean)),

    DO_days =
      sum(!is.na(DO_daily_mean)),

    pH_days =
      sum(!is.na(pH_daily_mean)),

    precipitation_days =
      sum(!is.na(precipitation_daily_mm)),

    .groups = "drop"
  )

seven_day_coverage
```

```{r}
missing_window_values <- event_window_data %>%
  select(
    location,
    station,
    month,
    date_collected,
    date,
    temperature_daily_mean,
    conductivity_daily_mean,
    salinity_daily_mean,
    DO_daily_mean,
    pH_daily_mean,
    precipitation_daily_mm
  ) %>%
  pivot_longer(
    cols = c(
      temperature_daily_mean,
      conductivity_daily_mean,
      salinity_daily_mean,
      DO_daily_mean,
      pH_daily_mean,
      precipitation_daily_mm
    ),
    names_to = "environmental_variable",
    values_to = "value"
  ) %>%
  filter(is.na(value)) %>%
  select(
    location,
    station,
    month,
    date_collected,
    missing_date = date,
    environmental_variable
  ) %>%
  arrange(
    date_collected,
    environmental_variable,
    missing_date
  )

print(
  missing_window_values,
  n = Inf
)
```

```{r}
# ============================================================
# 2G. Investigate missing environmental dates
# ============================================================

salinity_missing_dates <- missing_window_values %>%
  filter(environmental_variable == "salinity_daily_mean") %>%
  distinct(missing_date) %>%
  pull(missing_date)

salinity_gap_check <- salinity_raw %>%
  filter(date %in% salinity_missing_dates) %>%
  count(
    date,
    Farm,
    Location,
    name = "raw_rows"
  ) %>%
  arrange(date, Farm, Location)

print(
  salinity_gap_check,
  n = Inf
)
```

```{r}
ph_missing_dates <- missing_window_values %>%
  filter(environmental_variable == "pH_daily_mean") %>%
  distinct(missing_date) %>%
  pull(missing_date)

ph_gap_check <- ph_raw %>%
  filter(date %in% ph_missing_dates) %>%
  count(
    date,
    Farm,
    Location,
    name = "raw_rows"
  ) %>%
  arrange(date, Farm, Location)

print(
  ph_gap_check,
  n = Inf
)
```

```{r}
# Create daily temperature from the dissolved-oxygen sensor
do_temperature_daily <- do_raw %>%
  filter(
    Farm %in% c("CMAST", "DUML"),
    !is.na(date)
  ) %>%
  group_by(Farm, date) %>%
  summarise(
    temperature_DO_daily_mean = safe_mean(Temp_C),
    temperature_DO_observations = sum(!is.na(Temp_C)),
    .groups = "drop"
  ) %>%
  rename(station = Farm)
```

```{r}
do_temperature_coverage <- event_windows %>%
  left_join(
    do_temperature_daily,
    by = c("station", "date")
  ) %>%
  group_by(
    location,
    station,
    month,
    date_collected
  ) %>%
  summarise(
    expected_days = n_distinct(date),
    temperature_DO_days =
      sum(!is.na(temperature_DO_daily_mean)),
    .groups = "drop"
  ) %>%
  arrange(date_collected)

do_temperature_coverage
```

```{r}
# Install only if zoo is not already available
#if (!requireNamespace("zoo", quietly = TRUE)) {
#  install.packages("zoo")
#}

#library(zoo)
```

```{r}
# ============================================================
# 2H. Interpolate short salinity and conductivity gaps
# ============================================================

salinity_daily_interpolated <- salinity_daily %>%
  group_by(Farm) %>%
  tidyr::complete(
    date = seq(
      min(date, na.rm = TRUE),
      max(date, na.rm = TRUE),
      by = "day"
    )
  ) %>%
  arrange(Farm, date) %>%
  mutate(
    # Record whether the original daily measurement existed
    salinity_observed = !is.na(salinity_daily_mean),
    conductivity_observed = !is.na(conductivity_daily_mean),

    # Interpolate only internal gaps of four days or fewer
    salinity_daily_filled = zoo::na.approx(
      salinity_daily_mean,
      x = as.numeric(date),
      maxgap = 4,
      na.rm = FALSE
    ),

    conductivity_daily_filled = zoo::na.approx(
      conductivity_daily_mean,
      x = as.numeric(date),
      maxgap = 4,
      na.rm = FALSE
    ),

    # Flag values generated by interpolation
    salinity_imputed =
      is.na(salinity_daily_mean) &
      !is.na(salinity_daily_filled),

    conductivity_imputed =
      is.na(conductivity_daily_mean) &
      !is.na(conductivity_daily_filled)
  ) %>%
  ungroup()
```

```{r}
# Interpolate only pH gaps of two consecutive days or fewer
ph_daily_interpolated <- ph_daily %>%
  group_by(Farm) %>%
  tidyr::complete(
    date = seq(
      min(date, na.rm = TRUE),
      max(date, na.rm = TRUE),
      by = "day"
    )
  ) %>%
  arrange(Farm, date) %>%
  mutate(
    pH_observed = !is.na(pH_daily_mean),

    pH_daily_filled = zoo::na.approx(
      pH_daily_mean,
      x = as.numeric(date),
      maxgap = 2,
      na.rm = FALSE
    ),

    pH_imputed =
      is.na(pH_daily_mean) &
      !is.na(pH_daily_filled)
  ) %>%
  ungroup()
```

```{r}
salinity_values_imputed <- salinity_daily_interpolated %>%
  filter(salinity_imputed | conductivity_imputed) %>%
  select(
    Farm,
    date,
    salinity_daily_mean,
    salinity_daily_filled,
    salinity_imputed,
    conductivity_daily_mean,
    conductivity_daily_filled,
    conductivity_imputed
  ) %>%
  arrange(Farm, date)

ph_values_imputed <- ph_daily_interpolated %>%
  filter(pH_imputed) %>%
  select(
    Farm,
    date,
    pH_daily_mean,
    pH_daily_filled,
    pH_imputed
  ) %>%
  arrange(Farm, date)

print(salinity_values_imputed, n = Inf)
print(ph_values_imputed, n = Inf)
```

```{r}
tibble::tibble(
  variable = c("salinity", "conductivity", "pH"),
  values_imputed = c(
    sum(salinity_daily_interpolated$salinity_imputed),
    sum(salinity_daily_interpolated$conductivity_imputed),
    sum(ph_daily_interpolated$pH_imputed)
  ),
  values_still_missing = c(
    sum(is.na(salinity_daily_interpolated$salinity_daily_filled)),
    sum(is.na(salinity_daily_interpolated$conductivity_daily_filled)),
    sum(is.na(ph_daily_interpolated$pH_daily_filled))
  )
)
```

```{r}
# NOTE FOR MARINE LAB TEAM:
# Sensor gaps were identified before 7-day exposure calculation.
# Salinity and conductivity gaps of <= 4 consecutive days were
# linearly interpolated within station.
# pH gaps of <= 2 consecutive days were interpolated.
# The complete CMAST pH gap for 2024-08-13 through 2024-08-19
# was not imputed and requires confirmation or backup data.
```

Do we have backup or uncleaned sensor files for DUML on June 15–17, both farms around July 20–23, and CMAST pH for August 13–19? Were the sensors removed or moved during those periods?

```{r}
# ============================================================
# 2I. Recheck target windows after short-gap interpolation
# ============================================================

# Prepare interpolated salinity and conductivity data
salinity_filled_join <- salinity_daily_interpolated %>%
  transmute(
    station = Farm,
    date,

    salinity_daily_value = salinity_daily_filled,
    conductivity_daily_value = conductivity_daily_filled,

    salinity_was_imputed = salinity_imputed,
    conductivity_was_imputed = conductivity_imputed
  )


# Prepare interpolated pH data
ph_filled_join <- ph_daily_interpolated %>%
  transmute(
    station = Farm,
    date,

    pH_daily_value = pH_daily_filled,
    pH_was_imputed = pH_imputed
  )


# Temperature from the DO sensor has complete coverage
temperature_filled_join <- do_temperature_daily %>%
  transmute(
    station,
    date,
    temperature_daily_value = temperature_DO_daily_mean
  )


# Prepare DO data
do_filled_join <- do_daily %>%
  transmute(
    station = Farm,
    date,
    DO_daily_value = DO_daily_mean
  )
```

```{r}
event_window_final_check <- event_windows %>%
  left_join(
    salinity_filled_join,
    by = c("station", "date")
  ) %>%
  left_join(
    ph_filled_join,
    by = c("station", "date")
  ) %>%
  left_join(
    temperature_filled_join,
    by = c("station", "date")
  ) %>%
  left_join(
    do_filled_join,
    by = c("station", "date")
  ) %>%
  left_join(
    precip_daily %>%
      transmute(
        date,
        precipitation_daily_value =
          precipitation_daily_mm
      ),
    by = "date"
  )
```

```{r}
final_window_coverage <- event_window_final_check %>%
  group_by(
    location,
    station,
    month,
    date_collected
  ) %>%
  summarise(
    expected_days = n_distinct(date),

    temperature_days =
      sum(!is.na(temperature_daily_value)),

    salinity_days =
      sum(!is.na(salinity_daily_value)),

    conductivity_days =
      sum(!is.na(conductivity_daily_value)),

    DO_days =
      sum(!is.na(DO_daily_value)),

    pH_days =
      sum(!is.na(pH_daily_value)),

    precipitation_days =
      sum(!is.na(precipitation_daily_value)),

    salinity_days_imputed =
      sum(salinity_was_imputed %in% TRUE),

    conductivity_days_imputed =
      sum(conductivity_was_imputed %in% TRUE),

    pH_days_imputed =
      sum(pH_was_imputed %in% TRUE),

    .groups = "drop"
  ) %>%
  arrange(date_collected)

final_window_coverage
```

```{r}
# ============================================================
# 2J. Create coverage-aware environmental summaries
# ============================================================

environmental_events_7d <- event_window_final_check %>%
  group_by(
    location,
    station,
    month,
    date_collected
  ) %>%
  summarise(
    window_start = min(date),
    window_end = max(date),
    expected_days = n_distinct(date),

    # Complete temperature source: temperature from DO sensor
    temperature_7d_mean =
      safe_mean(temperature_daily_value),

    temperature_days_used =
      sum(!is.na(temperature_daily_value)),

    # Salinity: mean of available days in the intended 7-day window
    salinity_7d_available_mean =
      safe_mean(salinity_daily_value),

    salinity_days_used =
      sum(!is.na(salinity_daily_value)),

    # Conductivity: mean of available days in the intended window
    conductivity_7d_available_mean =
      safe_mean(conductivity_daily_value),

    conductivity_days_used =
      sum(!is.na(conductivity_daily_value)),

    # Dissolved oxygen
    DO_7d_mean =
      safe_mean(DO_daily_value),

    DO_days_used =
      sum(!is.na(DO_daily_value)),

    # pH: will remain NA when all seven days are unavailable
    pH_7d_available_mean =
      safe_mean(pH_daily_value),

    pH_days_used =
      sum(!is.na(pH_daily_value)),

    # Precipitation should be accumulated across the seven days
    precipitation_7d_total_mm =
      safe_sum(precipitation_daily_value),

    precipitation_days_used =
      sum(!is.na(precipitation_daily_value)),

    .groups = "drop"
  ) %>%

  # Add explicit coverage measures
  mutate(
    temperature_coverage =
      temperature_days_used / expected_days,

    salinity_coverage =
      salinity_days_used / expected_days,

    conductivity_coverage =
      conductivity_days_used / expected_days,

    DO_coverage =
      DO_days_used / expected_days,

    pH_coverage =
      pH_days_used / expected_days,

    precipitation_coverage =
      precipitation_days_used / expected_days,

    salinity_partial =
      salinity_days_used < expected_days,

    conductivity_partial =
      conductivity_days_used < expected_days,

    pH_partial =
      pH_days_used < expected_days
  ) %>%
  arrange(date_collected)

environmental_events_7d
```

```{r}
environmental_events_7d %>%
  select(
    location,
    station,
    month,
    date_collected,

    temperature_7d_mean,

    salinity_7d_available_mean,
    salinity_days_used,
    salinity_coverage,

    conductivity_7d_available_mean,
    conductivity_days_used,
    conductivity_coverage,

    DO_7d_mean,

    pH_7d_available_mean,
    pH_days_used,
    pH_coverage,

    precipitation_7d_total_mm
  )
```

```{r}
# We should have exactly six unique environmental events
stopifnot(nrow(environmental_events_7d) == 6)

# These three variables should have complete coverage
stopifnot(
  all(environmental_events_7d$temperature_days_used == 7),
  all(environmental_events_7d$DO_days_used == 7),
  all(environmental_events_7d$precipitation_days_used == 7)
)

# Exactly one event should currently have no pH measurements
sum(environmental_events_7d$pH_days_used == 0)

# Show that event
environmental_events_7d %>%
  filter(pH_days_used == 0) %>%
  select(
    location,
    station,
    month,
    date_collected,
    pH_days_used,
    pH_7d_available_mean
  )
```

```{r}
# ============================================================
# 2K. Create and save validated environmental matrices
# ============================================================

environmental_events_model <- environmental_events_7d %>%
  mutate(
    event_id = paste(
      location,
      format(date_collected, "%Y%m%d"),
      sep = "_"
    ),

    site_name = case_when(
      location == "DAF" ~ "Duke Aqua Farm",
      location == "CMT" ~ "CMAST / NC State",
      TRUE ~ NA_character_
    ),

    # Human-readable data-quality labels
    salinity_quality = case_when(
      salinity_days_used == 7 ~ "complete_7_of_7",
      salinity_days_used >= 4 ~ "partial_at_least_half",
      TRUE ~ "partial_under_half"
    ),

    conductivity_quality = case_when(
      conductivity_days_used == 7 ~ "complete_7_of_7",
      conductivity_days_used >= 4 ~ "partial_at_least_half",
      TRUE ~ "partial_under_half"
    ),

    pH_quality = case_when(
      pH_days_used == 7 ~ "complete_7_of_7",
      pH_days_used >= 4 ~ "partial_at_least_half",
      pH_days_used > 0 ~ "partial_under_half",
      TRUE ~ "no_data"
    )
  ) %>%
  relocate(
    event_id,
    site_name,
    location,
    station,
    month,
    date_collected
  )

environmental_events_model
```

```{r}
environmental_traits_7day_by_sample <- sample_metadata %>%
  left_join(
    environmental_events_model,
    by = c(
      "location",
      "station",
      "month",
      "date_collected"
    )
  ) %>%
  arrange(date_collected, location, oyster)

environmental_traits_7day_by_sample
```

```{r}
# Exactly 18 RNA-seq samples
stopifnot(
  nrow(environmental_traits_7day_by_sample) == 18
)

# Every sample ID should remain unique
stopifnot(
  !anyDuplicated(
    environmental_traits_7day_by_sample$sample_id
  )
)

# Every sample should match one environmental event
stopifnot(
  all(
    !is.na(
      environmental_traits_7day_by_sample$event_id
    )
  )
)
```

```{r}
sample_event_check <- environmental_traits_7day_by_sample %>%
  count(
    event_id,
    location,
    station,
    month,
    date_collected,
    name = "RNAseq_samples"
  )

sample_event_check
```

```{r}
write.csv(
  environmental_events_model,
  "environmental_events_7day_validated.csv",
  row.names = FALSE
)

write.csv(
  environmental_traits_7day_by_sample,
  "environmental_traits_7day_by_sample.csv",
  row.names = FALSE
)
```

```{r}
environmental_quality_summary <- environmental_events_model %>%
  select(
    event_id,
    location,
    station,
    month,
    date_collected,

    temperature_days_used,

    salinity_days_used,
    salinity_coverage,
    salinity_quality,

    conductivity_days_used,
    conductivity_coverage,
    conductivity_quality,

    DO_days_used,

    pH_days_used,
    pH_coverage,
    pH_quality,

    precipitation_days_used
  )

write.csv(
  environmental_quality_summary,
  "environmental_7day_data_quality_summary.csv",
  row.names = FALSE
)

environmental_quality_summary
```

**02 - WGCNA**

```{r}
# Quick Check

# Read the old analysis script as text
old_analysis_code <- readLines(
  "rstudio_analysis.Rmd",
  warn = FALSE
)

# Find the important WGCNA steps
grep(
  pattern = paste(
    "blockwiseModules",
    "moduleEigengenes",
    "moduleTrait",
    "environmental_traits_by_sample",
    sep = "|"
  ),
  x = old_analysis_code,
  value = TRUE
)
```

```{r}
# ============================================================
# 3A. Load and inspect WGCNA inputs
# ============================================================

# Load the saved VST-normalized expression object
vst_object <- readRDS(
  "oyster_deseq2_vst.rds"
)

# Check what type of object was saved
class(vst_object)
```

```{r}
# Extract the VST expression matrix
if (inherits(vst_object, "SummarizedExperiment")) {

  vst_matrix <- SummarizedExperiment::assay(
    vst_object
  )

} else if (
  is.matrix(vst_object) ||
  is.data.frame(vst_object)
) {

  vst_matrix <- as.matrix(vst_object)

} else {

  stop(
    "The saved VST object is not a recognized ",
    "SummarizedExperiment, matrix, or data frame."
  )
}
```

```{r}
dim(vst_matrix)

head(
  rownames(vst_matrix)
)

head(
  colnames(vst_matrix)
)
```

```{r}
gene_modules <- read.csv(
  "WGCNA_gene_modules.csv",
  stringsAsFactors = FALSE,
  check.names = FALSE
)

dim(gene_modules)

names(gene_modules)

head(gene_modules)
```

```{r}
sample_name_check <- list(

  metadata_not_in_VST = setdiff(
    sample_metadata$sample_id,
    colnames(vst_matrix)
  ),

  VST_not_in_metadata = setdiff(
    colnames(vst_matrix),
    sample_metadata$sample_id
  )
)

sample_name_check
```

```{r}
# ============================================================
# 3B. Standardize VST sample names and verify gene overlap
# ============================================================

# Save a record of the original sequencing names
vst_sample_name_map <- tibble::tibble(
  original_sample_id = colnames(vst_matrix),

  # Remove sequencing suffixes such as _S191_L005
  sample_id = stringr::str_remove(
    original_sample_id,
    "_S[0-9]+_L[0-9]+$"
  )
)

vst_sample_name_map
```

```{r}
duplicated_cleaned_names <- vst_sample_name_map %>%
  dplyr::count(sample_id) %>%
  dplyr::filter(n > 1)

duplicated_cleaned_names
```

```{r}
colnames(vst_matrix) <- vst_sample_name_map$sample_id

sample_name_check_cleaned <- list(
  metadata_not_in_VST = setdiff(
    sample_metadata$sample_id,
    colnames(vst_matrix)
  ),

  VST_not_in_metadata = setdiff(
    colnames(vst_matrix),
    sample_metadata$sample_id
  )
)

sample_name_check_cleaned
```

```{r}
vst_matrix <- vst_matrix[
  ,
  sample_metadata$sample_id,
  drop = FALSE
]

identical(
  colnames(vst_matrix),
  sample_metadata$sample_id
)
```

```{r}
# Remove incomplete module-assignment rows
gene_modules_clean <- gene_modules %>%
  dplyr::filter(
    !is.na(gene_id),
    !is.na(module),
    gene_id != "",
    module != ""
  )

# Confirm each gene has only one module assignment
duplicate_module_genes <- gene_modules_clean %>%
  dplyr::count(
    gene_id,
    name = "number_of_assignments"
  ) %>%
  dplyr::filter(number_of_assignments > 1)

duplicate_module_genes
```

```{r}
# Genes found in both the VST matrix and WGCNA module table
overlap_genes <- intersect(
  rownames(vst_matrix),
  gene_modules_clean$gene_id
)

gene_overlap_summary <- tibble::tibble(
  VST_genes = nrow(vst_matrix),

  module_assignment_genes =
    dplyr::n_distinct(gene_modules_clean$gene_id),

  overlapping_genes =
    length(overlap_genes),

  module_genes_missing_from_VST =
    length(
      setdiff(
        gene_modules_clean$gene_id,
        rownames(vst_matrix)
      )
    )
)

gene_overlap_summary
```

```{r}
module_sizes <- gene_modules_clean %>%
  dplyr::filter(gene_id %in% overlap_genes) %>%
  dplyr::count(
    module,
    name = "number_of_genes"
  ) %>%
  dplyr::arrange(
    dplyr::desc(number_of_genes)
  )

module_sizes
```

```{r}
# ============================================================
# 3C. Recalculate WGCNA module eigengenes
# ============================================================

library(WGCNA)

# Keep only assigned biological modules.
# Grey represents genes that were not assigned to a module.
module_gene_table <- gene_modules_clean %>%
  dplyr::filter(
    module != "grey",
    gene_id %in% rownames(vst_matrix)
  ) %>%
  dplyr::distinct(
    gene_id,
    module
  )
```

```{r}
nrow(module_gene_table)

table(module_gene_table$module)
```

```{r}
#Transposing
datExpr_modules <- t(
  vst_matrix[
    module_gene_table$gene_id,
    ,
    drop = FALSE
  ]
)

# Preserve the gene order explicitly
colnames(datExpr_modules) <- module_gene_table$gene_id

# Samples should remain the row names
rownames(datExpr_modules) <- colnames(vst_matrix)

dim(datExpr_modules)
```

```{r}
module_colors <- module_gene_table$module

names(module_colors) <- module_gene_table$gene_id
```

```{r}
identical(
  colnames(datExpr_modules),
  names(module_colors)
)
```

```{r}
good_expression_check <- WGCNA::goodSamplesGenes(
  datExpr_modules,
  verbose = 3
)

good_expression_check$allOK
```

```{r}
module_eigengenes <- WGCNA::moduleEigengenes(
  expr = datExpr_modules,
  colors = module_colors
)$eigengenes

# Put related eigengenes in a consistent order
module_eigengenes <- WGCNA::orderMEs(
  module_eigengenes
)

dim(module_eigengenes)

colnames(module_eigengenes)

head(module_eigengenes)
```

```{r}
identical(
  rownames(module_eigengenes),
  sample_metadata$sample_id
)
```

```{r}
# ============================================================
# 3D. Average module eigengenes within collection events
# ============================================================

# Convert row names into a regular sample ID column
module_eigengene_samples <- as.data.frame(
  module_eigengenes
) %>%
  tibble::rownames_to_column("sample_id")

# Attach corrected environmental and collection information
eigengenes_with_traits <- module_eigengene_samples %>%
  dplyr::left_join(
    environmental_traits_7day_by_sample,
    by = "sample_id"
  )

# Basic checks
dim(eigengenes_with_traits)

sum(is.na(eigengenes_with_traits$event_id))
```

```{r}
eigengenes_with_traits %>%
  dplyr::count(
    event_id,
    location,
    station,
    month,
    date_collected
  )
```

```{r}
module_columns <- colnames(module_eigengenes)

event_module_eigengenes <- eigengenes_with_traits %>%
  dplyr::group_by(
    event_id,
    location,
    station,
    month,
    date_collected
  ) %>%
  dplyr::summarise(
    dplyr::across(
      dplyr::all_of(module_columns),
      ~ mean(.x, na.rm = TRUE)
    ),

    RNAseq_samples = dplyr::n(),

    .groups = "drop"
  ) %>%
  dplyr::arrange(date_collected)

event_module_eigengenes
```

```{r}
# ============================================================
# 3E. Primary corrected module-environment heatmap
# ============================================================

primary_environment_traits <- environmental_events_model %>%
  dplyr::select(
    event_id,
    temperature_7d_mean,
    DO_7d_mean,
    precipitation_7d_total_mm
  )

event_analysis_primary <- event_module_eigengenes %>%
  dplyr::left_join(
    primary_environment_traits,
    by = "event_id"
  ) %>%
  dplyr::arrange(date_collected)

event_analysis_primary
```

```{r}
module_event_matrix <- event_analysis_primary %>%
  dplyr::select(
    dplyr::all_of(module_columns)
  ) %>%
  as.matrix()

rownames(module_event_matrix) <-
  event_analysis_primary$event_id


primary_trait_columns <- c(
  "temperature_7d_mean",
  "DO_7d_mean",
  "precipitation_7d_total_mm"
)

trait_event_matrix <- event_analysis_primary %>%
  dplyr::select(
    dplyr::all_of(primary_trait_columns)
  ) %>%
  as.matrix()

rownames(trait_event_matrix) <-
  event_analysis_primary$event_id
```

```{r}
dim(module_event_matrix)
dim(trait_event_matrix)

anyNA(module_event_matrix)
anyNA(trait_event_matrix)

identical(
  rownames(module_event_matrix),
  rownames(trait_event_matrix)
)
```

```{r}
primary_module_trait_cor <- cor(
  module_event_matrix,
  trait_event_matrix,
  method = "pearson"
)

primary_module_trait_p <- WGCNA::corPvalueStudent(
  primary_module_trait_cor,
  nSamples = nrow(event_analysis_primary)
)
```

```{r}

#Calculate false discovery rate adjusted p-values

primary_module_trait_FDR <- matrix(
  p.adjust(
    as.vector(primary_module_trait_p),
    method = "BH"
  ),
  nrow = nrow(primary_module_trait_p),
  ncol = ncol(primary_module_trait_p),
  dimnames = dimnames(primary_module_trait_p)
)
```

```{r}
#Heatmap labels

primary_heatmap_text <- matrix(
  paste0(
    sprintf(
      "%.2f",
      primary_module_trait_cor
    ),
    "\n(p=",
    sprintf(
      "%.3f",
      primary_module_trait_p
    ),
    ")"
  ),
  nrow = nrow(primary_module_trait_cor),
  ncol = ncol(primary_module_trait_cor),
  dimnames = dimnames(primary_module_trait_cor)
)
```

```{r}
plot_primary_environment_heatmap <- function() {

  WGCNA::labeledHeatmap(
    Matrix = primary_module_trait_cor,

    xLabels = c(
      "Temperature\n7-day mean",
      "DO\n7-day mean",
      "Precipitation\n7-day total"
    ),

    yLabels = rownames(
      primary_module_trait_cor
    ),

    ySymbols = rownames(
      primary_module_trait_cor
    ),

    colors = WGCNA::blueWhiteRed(50),

    textMatrix = primary_heatmap_text,

    setStdMargins = FALSE,

    cex.text = 0.75,

    zlim = c(-1, 1),

    main = paste0(
      "Corrected module-environment correlations\n",
      "Six independent collection events"
    )
  )
}

# Display in RStudio
plot_primary_environment_heatmap()
```

```{r}
pdf(
  "WGCNA_corrected_primary_environment_heatmap.pdf",
  width = 9,
  height = 7
)

plot_primary_environment_heatmap()

dev.off()
```

```{r}
primary_module_trait_results <- as.data.frame(
  as.table(primary_module_trait_cor)
)

names(primary_module_trait_results) <- c(
  "module",
  "environmental_trait",
  "correlation"
)

primary_module_trait_results$p_value <-
  as.vector(primary_module_trait_p)

primary_module_trait_results$FDR <-
  as.vector(primary_module_trait_FDR)

primary_module_trait_results <- primary_module_trait_results %>%
  dplyr::arrange(
    p_value
  )

primary_module_trait_results
```

```{r}
write.csv(
  primary_module_trait_results,
  "WGCNA_corrected_primary_environment_results.csv",
  row.names = FALSE
)
```

```{r}
#The strongest current associations are:

#MEred vs DO: r=−0.82, raw p=0.047
#MEblack vs DO: r=−0.80, raw p=0.054
#MEgreen vs temperature: r=0.74, raw p=0.093
#MEblue vs precipitation: r=−0.69, raw p=0.129
```

MEred and MEblack showed strong negative exploratory associations with dissolved oxygen, while MEgreen showed a positive association with temperature and MEblue showed a negative association with precipitation. None remained significant after multiple-testing correction, likely reflecting the limited number of independent collection events.

```{r}
# ============================================================
# 3F. Spearman sensitivity analysis for complete traits
# ============================================================

spearman_module_trait_cor <- cor(
  module_event_matrix,
  trait_event_matrix,
  method = "spearman",
  use = "complete.obs"
)

# Calculate exact correlation-test p-values for each pair
spearman_results <- expand.grid(
  module = colnames(module_event_matrix),
  environmental_trait = colnames(trait_event_matrix),
  stringsAsFactors = FALSE
) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(
    correlation = cor(
      module_event_matrix[, module],
      trait_event_matrix[, environmental_trait],
      method = "spearman"
    ),

    p_value = cor.test(
      module_event_matrix[, module],
      trait_event_matrix[, environmental_trait],
      method = "spearman",
      exact = FALSE
    )$p.value
  ) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    FDR = p.adjust(
      p_value,
      method = "BH"
    )
  ) %>%
  dplyr::arrange(p_value)

spearman_results
```

```{r}
spearman_results %>%
  dplyr::filter(
    module %in% c(
      "MEred",
      "MEblack",
      "MEgreen",
      "MEblue"
    )
  ) %>%
  dplyr::arrange(p_value)
```

```{r}
#Our leading complete-data candidates are now:

#MEblack–precipitation
#MEblue–precipitation
#MEblack–DO

#The black module is especially interesting because it is positively associated with precipitation but negatively associated with dissolved oxygen. Before interpreting that biologically, we must check whether precipitation and DO themselves are correlated.
```

```{r}
# ============================================================
# 3G. Check correlations among complete environmental traits
# ============================================================

environment_trait_data <- event_analysis_primary %>%
  dplyr::select(
    event_id,
    location,
    month,
    temperature_7d_mean,
    DO_7d_mean,
    precipitation_7d_total_mm
  )

environment_trait_data
```

```{r}
environment_pearson_cor <- environment_trait_data %>%
  dplyr::select(
    temperature_7d_mean,
    DO_7d_mean,
    precipitation_7d_total_mm
  ) %>%
  cor(
    method = "pearson",
    use = "complete.obs"
  )

environment_spearman_cor <- environment_trait_data %>%
  dplyr::select(
    temperature_7d_mean,
    DO_7d_mean,
    precipitation_7d_total_mm
  ) %>%
  cor(
    method = "spearman",
    use = "complete.obs"
  )

environment_pearson_cor
environment_spearman_cor
```

```{r}
precip_DO_pearson <- cor.test(
  environment_trait_data$precipitation_7d_total_mm,
  environment_trait_data$DO_7d_mean,
  method = "pearson"
)

precip_DO_spearman <- cor.test(
  environment_trait_data$precipitation_7d_total_mm,
  environment_trait_data$DO_7d_mean,
  method = "spearman",
  exact = FALSE
)

precip_DO_pearson
precip_DO_spearman

#Main finding: precipitation and DO move in opposite directions

#Collection windows with more precipitation tended to have lower dissolved oxygen. This could be collinearity or confounding variables.

#MEblack is associated with an environmental gradient characterized by higher precipitation and lower dissolved oxygen
```

```{r}
#MEblack showed a strong association with an environmental pattern involving higher precipitation and lower dissolved oxygen.

#MEblue showed a consistent negative association with precipitation across Pearson and Spearman analyses.

#None of the associations survived multiple-testing correction, and the small number of independent collection events limits statistical certainty.
```

```{r}
# ============================================================
# 3H. Visualize precipitation and dissolved oxygen
# ============================================================

library(ggplot2)

ggplot(
  environment_trait_data,
  aes(
    x = precipitation_7d_total_mm,
    y = DO_7d_mean,
    shape = month,
    label = event_id
  )
) +
  geom_point(size = 3) +
  geom_smooth(
    method = "lm",
    se = FALSE
  ) +
  geom_text(
    vjust = -0.8,
    check_overlap = TRUE,
    size = 3
  ) +
  labs(
    title = "Seven-day precipitation versus dissolved oxygen",
    subtitle = "Six independent oyster collection events",
    x = "Seven-day precipitation total (mm)",
    y = "Seven-day mean dissolved oxygen"
  ) +
  theme_minimal()

#Precipitation and DO showed an overall negative association across the six events, but this pattern may be partly driven by seasonal differences and the unusually low dissolved oxygen observed at DAF in August.
```

```{r}
ggplot(
  environment_trait_data,
  aes(
    x = precipitation_7d_total_mm,
    y = DO_7d_mean
  )
) +
  geom_point(
    aes(shape = month),
    size = 3
  ) +
  geom_smooth(
    aes(group = 1),
    method = "lm",
    se = TRUE
  ) +
  geom_text(
    aes(label = event_id),
    vjust = -0.8,
    check_overlap = TRUE,
    size = 3
  ) +
  labs(
    title = "Seven-day precipitation versus dissolved oxygen",
    subtitle = "Six independent oyster collection events",
    x = "Seven-day precipitation total (mm)",
    y = "Seven-day mean dissolved oxygen",
    shape = "Month"
  ) +
  theme_minimal()
```

```{r}
# ============================================================
# 3I. Leave-one-event-out sensitivity analysis
# ============================================================

leave_one_out_precip_DO <- purrr::map_dfr(
  seq_len(nrow(environment_trait_data)),
  function(i) {

    reduced_data <- environment_trait_data[-i, ]

    tibble::tibble(
      removed_event =
        environment_trait_data$event_id[i],

      pearson_correlation = cor(
        reduced_data$precipitation_7d_total_mm,
        reduced_data$DO_7d_mean,
        method = "pearson"
      ),

      spearman_correlation = cor(
        reduced_data$precipitation_7d_total_mm,
        reduced_data$DO_7d_mean,
        method = "spearman"
      )
    )
  }
)

leave_one_out_precip_DO

#Across the six collection events, higher seven-day precipitation generally coincided with lower dissolved oxygen. The direction remains negative when any single event is removed, although the magnitude is sensitive to which event is excluded.
```

**04 - Linear Regression**

```{r}

# ============================================================
# 4A. Linear regression: MEblack versus precipitation
# ============================================================

MEblack_precip_data <- event_analysis_primary %>%
  dplyr::select(
    event_id,
    location,
    month,
    MEblack,
    precipitation_7d_total_mm
  ) %>%
  dplyr::mutate(
    # Express precipitation in 10-mm units for easier interpretation
    precipitation_10mm =
      precipitation_7d_total_mm / 10
  )

MEblack_precip_data
```

```{r}
MEblack_precip_model <- lm(
  MEblack ~ precipitation_10mm,
  data = MEblack_precip_data
)

summary(MEblack_precip_model)
```

```{r}
confint(
  MEblack_precip_model,
  level = 0.95
)
```

```{r}
MEblack_precip_diagnostics <- MEblack_precip_data %>%
  dplyr::mutate(
    fitted_value = fitted(MEblack_precip_model),
    residual = residuals(MEblack_precip_model),
    cooks_distance = cooks.distance(MEblack_precip_model)
  ) %>%
  dplyr::arrange(
    dplyr::desc(cooks_distance)
  )

MEblack_precip_diagnostics
```

```{r}
4 / nrow(MEblack_precip_data)

#For every additional 10 mm of precipitation over the seven-day window, MEblack is estimated to increase by about 0.076 eigengene units.

#MEblack tended to increase with seven-day precipitation, but the evidence was inconclusive given the small number of independent collection events.

#The raw R² suggests precipitation accounts for about 45% of the observed variation in MEblack across the six events.After adjusting for the very small sample size, that drops to about 31%. That is still potentially meaningful, but quite uncertain.
```

```{r}
#June: low precipitation and low MEblack
#July: moderate precipitation and low/slightly higher MEblack
#August: high precipitation and much higher MEblack
```

```{r}
# ============================================================
# 4B. Leave-one-event-out MEblack precipitation regression
# ============================================================

MEblack_precip_leave_one_out <- purrr::map_dfr(
  seq_len(nrow(MEblack_precip_data)),
  function(i) {

    reduced_data <- MEblack_precip_data[-i, ]

    reduced_model <- stats::lm(
      MEblack ~ precipitation_10mm,
      data = reduced_data
    )

    model_summary <- summary(reduced_model)

    tibble::tibble(
      removed_event = MEblack_precip_data$event_id[i],

      precipitation_coefficient =
        stats::coef(reduced_model)[["precipitation_10mm"]],

      p_value =
        model_summary$coefficients[
          "precipitation_10mm",
          "Pr(>|t|)"
        ],

      R_squared =
        model_summary$r.squared,

      adjusted_R_squared =
        model_summary$adj.r.squared
    )
  }
)

MEblack_precip_leave_one_out

#The positive association between precipitation and the black-module eigengene was directionally robust to removing individual collection events. However, the estimated magnitude varied, and statistical uncertainty remained high because only six independent collection events were available.
```

```{r}
# ============================================================
# 4C. Plot MEblack versus seven-day precipitation
# ============================================================

MEblack_precip_plot <- ggplot(
  MEblack_precip_data,
  aes(
    x = precipitation_7d_total_mm,
    y = MEblack
  )
) +
  geom_point(
    aes(shape = month),
    size = 3
  ) +
  geom_smooth(
    aes(group = 1),
    method = "lm",
    se = TRUE
  ) +
  geom_text(
    aes(label = event_id),
    vjust = -0.8,
    check_overlap = TRUE,
    size = 3
  ) +
  labs(
    title = "Black module eigengene versus precipitation",
    subtitle = "Six independent oyster collection events",
    x = "Seven-day precipitation total (mm)",
    y = "MEblack",
    shape = "Month"
  ) +
  theme_minimal()

MEblack_precip_plot
```

```{r}
ggsave(
  "MEblack_vs_precipitation_7day.png",
  plot = MEblack_precip_plot,
  width = 9,
  height = 6,
  dpi = 300
)

#MEblack increased along a seasonal gradient that also corresponded with increasing seven-day precipitation. The available data cannot clearly separate precipitation from month.
```

```{r}
# ============================================================
# 4D. Compare MEblack patterns across months
# ============================================================

MEblack_month_data <- MEblack_precip_data %>%
  dplyr::mutate(
    month = factor(
      month,
      levels = c("Jun", "Jul", "Aug")
    )
  )

# Summarize the two collection events in each month
MEblack_month_summary <- MEblack_month_data %>%
  dplyr::group_by(month) %>%
  dplyr::summarise(
    mean_MEblack = mean(MEblack),
    minimum_MEblack = min(MEblack),
    maximum_MEblack = max(MEblack),
    number_of_events = dplyr::n(),
    .groups = "drop"
  )

MEblack_month_summary
```

```{r}
MEblack_month_model <- lm(
  MEblack ~ month,
  data = MEblack_month_data
)

summary(MEblack_month_model)
```

```{r}
model_comparison <- tibble::tibble(
  model = c(
    "Precipitation",
    "Month"
  ),

  R_squared = c(
    summary(MEblack_precip_model)$r.squared,
    summary(MEblack_month_model)$r.squared
  ),

  adjusted_R_squared = c(
    summary(MEblack_precip_model)$adj.r.squared,
    summary(MEblack_month_model)$adj.r.squared
  ),

  AIC = c(
    AIC(MEblack_precip_model),
    AIC(MEblack_month_model)
  )
)

model_comparison

#MEblack was substantially higher in August than in June or July. Because seven-day precipitation also rose strongly from June through August, the apparent precipitation association likely reflects a broader seasonal pattern rather than an independently identifiable precipitation effect.
```

```{r}
MEblack_month_plot <- ggplot(
  MEblack_month_data,
  aes(
    x = month,
    y = MEblack,
    group = location,
    shape = location
  )
) +
  geom_line() +
  geom_point(size = 3) +
  geom_text(
    aes(label = event_id),
    vjust = -0.8,
    check_overlap = TRUE,
    size = 3
  ) +
  labs(
    title = "Black module eigengene across collection months",
    subtitle = "Lines connect collection events from the same location",
    x = "Collection month",
    y = "MEblack",
    shape = "Location"
  ) +
  theme_minimal()

MEblack_month_plot
```

```{r}
# ============================================================
# 4E. Compare paired DAF and CMT MEblack values by month
# ============================================================

MEblack_location_differences <- MEblack_month_data %>%
  dplyr::select(
    month,
    location,
    MEblack
  ) %>%
  tidyr::pivot_wider(
    names_from = location,
    values_from = MEblack
  ) %>%
  dplyr::mutate(
    DAF_minus_CMT = DAF - CMT
  )

MEblack_location_differences
```

```{r}
MEblack_month_plot <- ggplot(
  MEblack_month_data,
  aes(
    x = month,
    y = MEblack,
    group = location,
    shape = location
  )
) +
  geom_line() +
  geom_point(size = 3) +
  geom_text(
    aes(label = event_id),
    vjust = -0.8,
    size = 3,
    check_overlap = TRUE
  ) +
  labs(
    title = "Black module eigengene across collection months",
    subtitle = "Lines connect collection events from the same location",
    x = "Collection month",
    y = "MEblack",
    shape = "Location"
  ) +
  theme_minimal()

MEblack_month_plot

#The black module was relatively stable and similar between locations in June and July, then increased at both locations in August, with a much larger August increase at DAF.

#MEblack was associated with a seasonal gradient that included increasing precipitation, but the pronounced August increase at DAF suggests additional site-specific environmental or biological factors.
```

```{r}
# ============================================================
# 4F. Compare month-only and additive month + location models
# ============================================================

MEblack_month_location_data <- MEblack_month_data %>%
  dplyr::mutate(
    location = factor(
      location,
      levels = c("CMT", "DAF")
    )
  )

MEblack_month_location_model <- lm(
  MEblack ~ month + location,
  data = MEblack_month_location_data
)

summary(MEblack_month_location_model)
```

```{r}
model_comparison_extended <- tibble::tibble(
  model = c(
    "Precipitation",
    "Month",
    "Month + location"
  ),

  R_squared = c(
    summary(MEblack_precip_model)$r.squared,
    summary(MEblack_month_model)$r.squared,
    summary(MEblack_month_location_model)$r.squared
  ),

  adjusted_R_squared = c(
    summary(MEblack_precip_model)$adj.r.squared,
    summary(MEblack_month_model)$adj.r.squared,
    summary(MEblack_month_location_model)$adj.r.squared
  ),

  AIC = c(
    AIC(MEblack_precip_model),
    AIC(MEblack_month_model),
    AIC(MEblack_month_location_model)
  )
)

model_comparison_extended

#MEblack variation was explained primarily by collection month, with a strong increase in August. Adding an overall location effect did not materially improve the model.
```

```{r}
anova(
  MEblack_month_model,
  MEblack_month_location_model
)

#MEblack remained similar between June and July and increased strongly in August at both locations. The increase was substantially larger at DAF, but the dataset is too small to formally test a month-by-location interaction. Month described MEblack better than precipitation alone, and adding a constant location effect did not improve the model.
```

```{r}
# Check whether the interaction model (location * month) would be saturated
interaction_design <- model.matrix(
  ~ month * location,
  data = MEblack_month_location_data
)

tibble::tibble(
  observations = nrow(interaction_design),
  model_parameters = qr(interaction_design)$rank,
  residual_degrees_of_freedom =
    nrow(interaction_design) - qr(interaction_design)$rank
)
```

```{r}
# ============================================================
# 4G. Complete simple module-regression screening table
# ============================================================

run_simple_module_model <- function(
    module_name,
    predictor_name,
    analysis_data
) {

  model_formula <- stats::reformulate(
    predictor_name,
    response = module_name
  )

  fitted_model <- stats::lm(
    model_formula,
    data = analysis_data
  )

  model_summary <- summary(fitted_model)
  coefficient_table <- model_summary$coefficients

  confidence_interval <- stats::confint(
    fitted_model,
    level = 0.95
  )

  tibble::tibble(
    module = module_name,
    predictor = predictor_name,
    n_events = stats::nobs(fitted_model),

    coefficient =
      stats::coef(fitted_model)[[predictor_name]],

    standard_error =
      coefficient_table[
        predictor_name,
        "Std. Error"
      ],

    t_value =
      coefficient_table[
        predictor_name,
        "t value"
      ],

    p_value =
      coefficient_table[
        predictor_name,
        "Pr(>|t|)"
      ],

    confidence_low =
      confidence_interval[
        predictor_name,
        1
      ],

    confidence_high =
      confidence_interval[
        predictor_name,
        2
      ],

    R_squared =
      model_summary$r.squared,

    adjusted_R_squared =
      model_summary$adj.r.squared,

    AIC = stats::AIC(fitted_model)
  )
}
```

```{r}
#Run every module trait comparison

all_primary_module_regressions <- tidyr::crossing(
  module = module_columns,
  predictor = primary_trait_columns
) %>%
  purrr::pmap_dfr(
    function(module, predictor) {
      run_simple_module_model(
        module_name = module,
        predictor_name = predictor,
        analysis_data = event_analysis_primary
      )
    }
  ) %>%
  dplyr::mutate(
    FDR = stats::p.adjust(
      p_value,
      method = "BH"
    )
  ) %>%
  dplyr::arrange(p_value)

all_primary_module_regressions
```

```{r}
write.csv(
  all_primary_module_regressions,
  "WGCNA_corrected_primary_module_regressions.csv",
  row.names = FALSE
)
```

```{r}
saveRDS(
  module_eigengenes,
  "WGCNA_corrected_module_eigengenes.rds"
)

saveRDS(
  event_module_eigengenes,
  "WGCNA_event_level_module_eigengenes.rds"
)

saveRDS(
  list(
    MEblack_precipitation = MEblack_precip_model,
    MEblack_month = MEblack_month_model,
    MEblack_month_location =
      MEblack_month_location_model
  ),
  "WGCNA_candidate_module_models.rds"
)

writeLines(
  capture.output(sessionInfo()),
  "environmental_regression_sessionInfo.txt"
)
```

```{r}

#Questions I need answered before contiuing

#Are backup or uncleaned logger files available for the missing salinity, conductivity, and pH periods?

#Is temperature from the DO sensor an acceptable substitute for the incomplete salinity logger temperature?

#Should the exposure window include collection day plus the previous six days, or the preceding 168 hours before collection?

#Is precipitation one shared regional measurement, or should each site have its own precipitation source?

#For later gene-level modeling, should oyster identity be treated as a repeated-measures effect?

#Does the large DAF-specific MEblack increase in August align with any known farm event, stressor, handling difference, storm, mortality, or water-quality change?
```

```{r}
#corrected the reversed station mapping;

#rebuilt seven-day environmental exposures;

#documented missing sensor coverage;

#recalculated WGCNA eigengenes;

#rebuilt the environmental correlation heatmap;

#found that the old MEblue–DO result does not persist;

#identified MEblack as associated with a seasonal pattern involving precipitation and DO;

#found that month explains MEblack better than precipitation alone;

#observed a potentially DAF-specific August increase that cannot yet be formally tested.
```
