---
title: "Sophie Counts: DESeq2 VST and WGCNA Environmental Analysis"
author: "Arushi Singh"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 3
---

# Purpose

This file performs only the steps needed before module-level regression:

1.  Load Sophie's raw featureCounts matrix.
2.  Build a DESeq2 object and calculate variance-stabilized expression values.
3.  Select the most variable genes and construct a WGCNA network.
4.  Calculate sample-level and six-event-level module eigengenes.
5.  Join the final spline-filled environmental matrix.
6.  Generate the expanded module-trait correlation heatmap and save all regression-ready outputs.

No PCA, differential-expression testing, GO enrichment, or regression is run here.

# 0. One-time environmental export

Run the following **once in the R session containing your final spline-filled environmental object**. Confirm that the object contains six rows and the required variables before exporting it.

```{r}
environmental_events_7d <- readr::read_csv(
  environmental_output_path,
  show_col_types = FALSE
)
```

```{r export-environment-once, eval=FALSE}
environmental_output_path <- paste0(
  "/work/clh162/Data+/Arushi/2025-2026-team-code/",
  "Data+_team/Arushi/scripts/",
  "environmental_events_7day_spline_final.csv"
)

required_environment_columns <- c(
  "event_id",
  "location",
  "month",
  "temperature_7d_mean",
  "DO_7d_mean",
  "salinity_7d_spline_mean",
  "conductivity_7d_spline_mean",
  "pH_7d_available_mean",
  "precipitation_7d_total_mm"
)

stopifnot(
  nrow(environmental_events_7d) == 6,
  all(required_environment_columns %in% names(environmental_events_7d))
)

readr::write_csv(
  environmental_events_7d,
  environmental_output_path
)
```

# 1. Setup

```{r setup, message=FALSE, warning=FALSE}
library(DESeq2)
library(WGCNA)
library(tidyverse)
library(matrixStats)
library(pheatmap)

options(stringsAsFactors = FALSE)
allowWGCNAThreads()
set.seed(2026)
```

```{r paths}
sophie_count_file <- paste0(
  "/work/clh162/Data+/Sophie/",
  "2025-2026-team-code-sac165/",
  "Data+_team/Sophie/gene_counts/gene_counts.csv"
)

environment_file <- paste0(
  "/work/clh162/Data+/Arushi/2025-2026-team-code/",
  "Data+_team/Arushi/scripts/",
  "environmental_events_7day_spline_final.csv"
)

output_directory <- paste0(
  "/work/clh162/Data+/Arushi/2025-2026-team-code/",
  "Data+_team/Arushi/scripts/",
  "SophieCounts_WGCNA"
)

dir.create(output_directory, recursive = TRUE, showWarnings = FALSE)

stopifnot(
  file.exists(sophie_count_file),
  file.exists(environment_file)
)
```

# 2. Load and prepare Sophie's raw counts

```{r load-sophie-counts}
sophie_raw <- readr::read_csv(
  sophie_count_file,
  comment = "#",
  name_repair = "minimal",
  show_col_types = FALSE,
  trim_ws = TRUE
)

dim(sophie_raw)
names(sophie_raw)[1:8]
```

```{r identify-columns}
sophie_gene_candidates <- intersect(
  c("Geneid", "gene_id", "GeneID", "gene", "Gene"),
  names(sophie_raw)
)

if (length(sophie_gene_candidates) == 0) {
  stop("No gene-ID column was detected in Sophie's count file.")
}

sophie_gene_column <- sophie_gene_candidates[1]

sophie_sample_columns <- grep(
  "(CMT|DAF)_SJ[123]_",
  names(sophie_raw),
  value = TRUE
)

stopifnot(
  length(sophie_sample_columns) == 18,
  !anyDuplicated(sophie_sample_columns)
)

sophie_gene_column
length(sophie_sample_columns)
```

```{r clean-sample-names}
clean_sample_id <- function(sample_name) {
  sample_name %>%
    basename() %>%
    stringr::str_remove("\\.bam$") %>%
    stringr::str_remove("_sorted$") %>%
    stringr::str_remove("_S[0-9]+_L[0-9]+$") %>%
    stringr::str_remove("_S[0-9]+$") %>%
    stringr::str_remove("_Gill$")
}

sample_name_map <- tibble(
  original_sample_name = sophie_sample_columns,
  sample_id = clean_sample_id(sophie_sample_columns)
)

stopifnot(n_distinct(sample_name_map$sample_id) == 18)

sample_name_map
```

```{r build-count-matrix}
count_matrix_sophie <- as.matrix(
  sophie_raw[, sophie_sample_columns, drop = FALSE]
)

storage.mode(count_matrix_sophie) <- "numeric"

rownames(count_matrix_sophie) <- trimws(
  as.character(sophie_raw[[sophie_gene_column]])
)

colnames(count_matrix_sophie) <- sample_name_map$sample_id

stopifnot(
  nrow(count_matrix_sophie) == 28538,
  ncol(count_matrix_sophie) == 18,
  !anyNA(count_matrix_sophie),
  all(count_matrix_sophie >= 0),
  all(count_matrix_sophie %% 1 == 0),
  !anyDuplicated(rownames(count_matrix_sophie)),
  !anyDuplicated(colnames(count_matrix_sophie))
)

storage.mode(count_matrix_sophie) <- "integer"

dim(count_matrix_sophie)
```

# 3. Construct sample metadata

```{r sample-metadata}
sample_metadata_sophie <- tibble(
  sample_id = colnames(count_matrix_sophie)
) %>%
  tidyr::separate(
    sample_id,
    into = c("location", "oyster_id", "collection_code"),
    sep = "_",
    remove = FALSE
  ) %>%
  mutate(
    date_collected = lubridate::dmy(collection_code),
    month = factor(format(date_collected, "%b"), levels = c("Jun", "Jul", "Aug")),
    location = factor(location, levels = c("CMT", "DAF")),
    station = if_else(location == "CMT", "CMAST", "DUML"),
    event_id = paste0(as.character(location), "_", format(date_collected, "%Y%m%d"))
  ) %>%
  arrange(date_collected, location, oyster_id) %>%
  as.data.frame()

rownames(sample_metadata_sophie) <- sample_metadata_sophie$sample_id

count_matrix_sophie <- count_matrix_sophie[
  , sample_metadata_sophie$sample_id, drop = FALSE
]

stopifnot(
  identical(colnames(count_matrix_sophie), rownames(sample_metadata_sophie)),
  n_distinct(sample_metadata_sophie$event_id) == 6,
  all(table(sample_metadata_sophie$event_id) == 3)
)

sample_metadata_sophie
```

# 4. DESeq2 filtering and variance-stabilizing transformation

A full differential-expression model is not needed in this file. We create the DESeq2 object, remove very low-count genes, and calculate variance-stabilized expression values for WGCNA.

```{r deseq2-object}
dds_sophie <- DESeqDataSetFromMatrix(
  countData = count_matrix_sophie,
  colData = as.data.frame(sample_metadata_sophie),
  design = ~ location + month
)

keep_gene <- rowSums(counts(dds_sophie) >= 10) >= 3

dds_sophie <- dds_sophie[keep_gene, ]

c(
  genes_before_filtering = nrow(count_matrix_sophie),
  genes_after_filtering = nrow(dds_sophie)
)
```

```{r vst}
vst_sophie <- vst(dds_sophie, blind = TRUE)
vst_matrix_sophie <- assay(vst_sophie)

stopifnot(
  identical(colnames(vst_matrix_sophie), rownames(sample_metadata_sophie)),
  !anyNA(vst_matrix_sophie)
)

dim(vst_matrix_sophie)
```

```{r save-deseq-objects}
saveRDS(
  dds_sophie,
  file.path(output_directory, "SophieCounts_deseq2_dds_filtered.rds")
)

saveRDS(
  vst_sophie,
  file.path(output_directory, "SophieCounts_deseq2_vst.rds")
)
```

# 5. Select genes for WGCNA

Use the top 5% of filtered genes by variance, matching the scale of the earlier exploratory WGCNA.

```{r variable-gene-selection}
gene_variance <- matrixStats::rowVars(vst_matrix_sophie)
names(gene_variance) <- rownames(vst_matrix_sophie)

number_of_wgcna_genes <- ceiling(length(gene_variance) * 0.05)

selected_wgcna_genes <- names(
  sort(gene_variance, decreasing = TRUE)
)[seq_len(number_of_wgcna_genes)]

datExpr_sophie <- t(
  vst_matrix_sophie[selected_wgcna_genes, , drop = FALSE]
)

dim(datExpr_sophie)
```

```{r good-samples-genes}
good_expression_check <- goodSamplesGenes(datExpr_sophie, verbose = 3)

if (!good_expression_check$allOK) {
  datExpr_sophie <- datExpr_sophie[
    good_expression_check$goodSamples,
    good_expression_check$goodGenes,
    drop = FALSE
  ]
}

stopifnot(
  nrow(datExpr_sophie) == 18,
  identical(rownames(datExpr_sophie), rownames(sample_metadata_sophie)),
  !anyNA(datExpr_sophie)
)

dim(datExpr_sophie)
```

# 6. Select a soft-thresholding power

```{r soft-threshold}
candidate_powers <- c(1:10, seq(12, 30, by = 2))

soft_threshold_results <- pickSoftThreshold(
  datExpr_sophie,
  powerVector = candidate_powers,
  networkType = "signed",
  corFnc = "cor",
  verbose = 5
)

soft_fit_table <- as_tibble(
  soft_threshold_results$fitIndices
) %>%
  transmute(
    power = Power,
    signed_scale_free_R2 = -sign(slope) * SFT.R.sq,
    mean_connectivity = mean.k.
  )

eligible_power <- soft_fit_table %>%
  filter(signed_scale_free_R2 >= 0.80) %>%
  slice_min(power, n = 1, with_ties = FALSE)

if (nrow(eligible_power) == 1) {
  soft_power <- eligible_power$power
} else {
  soft_power <- soft_fit_table %>%
    slice_max(signed_scale_free_R2, n = 1, with_ties = FALSE) %>%
    pull(power)
}

soft_fit_table
soft_power
```

# 7. Construct the WGCNA network

```{r wgcna-network}
network_sophie <- blockwiseModules(
  datExpr_sophie,
  power = soft_power,
  networkType = "signed",
  TOMType = "signed",
  minModuleSize = 30,
  deepSplit = 2,
  mergeCutHeight = 0.25,
  pamRespectsDendro = FALSE,
  numericLabels = FALSE,
  reassignThreshold = 0,
  maxBlockSize = ncol(datExpr_sophie),
  saveTOMs = FALSE,
  verbose = 3
)

# Extract module assignments from the completed WGCNA network
module_colors_sophie <- network_sophie$colors

# Convert to a simple character vector if WGCNA returned a list-like object
if (is.list(module_colors_sophie)) {
  module_colors_sophie <- unlist(
    module_colors_sophie,
    recursive = TRUE,
    use.names = FALSE
  )
}

module_colors_sophie <- as.character(
  module_colors_sophie
)

# Each WGCNA gene should have exactly one module assignment
stopifnot(
  length(module_colors_sophie) == ncol(datExpr_sophie),
  !anyNA(module_colors_sophie)
)

names(module_colors_sophie) <- colnames(
  datExpr_sophie
)

# Create module-size summary
module_size_table <- tibble::tibble(
  module = module_colors_sophie
) %>%
  dplyr::group_by(module) %>%
  dplyr::summarise(
    number_of_genes = dplyr::n(),
    .groups = "drop"
  ) %>%
  dplyr::arrange(
    dplyr::desc(number_of_genes)
  )

module_size_table
```

```{r module-gene-table}
module_gene_table_sophie <- tibble(
  gene_id = names(module_colors_sophie),
  module = unname(module_colors_sophie)
)

readr::write_csv(
  module_gene_table_sophie,
  file.path(output_directory, "SophieCounts_WGCNA_gene_modules.csv")
)

readr::write_csv(
  module_size_table,
  file.path(output_directory, "SophieCounts_WGCNA_module_sizes.csv")
)
```

# 8. Calculate module eigengenes

The grey module is excluded because it contains unassigned genes.

```{r module-eigengenes}
non_grey_genes <- module_colors_sophie != "grey"

datExpr_modules_sophie <- datExpr_sophie[
  , non_grey_genes, drop = FALSE
]

module_colors_non_grey <- module_colors_sophie[non_grey_genes]

module_eigengenes_sophie <- moduleEigengenes(
  datExpr_modules_sophie,
  colors = module_colors_non_grey
)$eigengenes

module_eigengenes_sophie <- orderMEs(module_eigengenes_sophie)

stopifnot(
  identical(
    rownames(module_eigengenes_sophie),
    rownames(sample_metadata_sophie)
  )
)

dim(module_eigengenes_sophie)
names(module_eigengenes_sophie)
```

```{r sample-level-eigengenes}
sample_level_eigengenes_sophie <- bind_cols(
  sample_metadata_sophie %>%
    rownames_to_column(var = "metadata_rowname") %>%
    select(-metadata_rowname),
  as_tibble(module_eigengenes_sophie)
)

readr::write_csv(
  sample_level_eigengenes_sophie,
  file.path(output_directory, "SophieCounts_WGCNA_sample_level_eigengenes.csv")
)

saveRDS(
  module_eigengenes_sophie,
  file.path(output_directory, "SophieCounts_WGCNA_sample_level_eigengenes.rds")
)
```

# 9. Collapse eigengenes to six independent collection events

```{r event-level-eigengenes}
eigengene_columns <- grep(
  "^ME",
  names(sample_level_eigengenes_sophie),
  value = TRUE
)

event_level_eigengenes_sophie <- sample_level_eigengenes_sophie %>%
  group_by(event_id, location, station, month, date_collected) %>%
  summarise(
    across(all_of(eigengene_columns), mean),
    RNAseq_samples = n(),
    .groups = "drop"
  ) %>%
  arrange(date_collected, location)

stopifnot(
  nrow(event_level_eigengenes_sophie) == 6,
  all(event_level_eigengenes_sophie$RNAseq_samples == 3)
)

event_level_eigengenes_sophie
```

```{r save-event-eigengenes}
readr::write_csv(
  event_level_eigengenes_sophie,
  file.path(output_directory, "SophieCounts_WGCNA_event_level_eigengenes.csv")
)

saveRDS(
  event_level_eigengenes_sophie,
  file.path(output_directory, "SophieCounts_WGCNA_event_level_eigengenes.rds")
)
```

```{r}
ls(pattern = "environment")
```

# 10. Load and join the final environmental matrix

```{r}
# Confirm the source table has the expected columns
names(environmental_events_7d)

# Create the finalized event-level environmental table
environmental_events_final <-
  environmental_events_7d %>%
  dplyr::mutate(
    location = trimws(as.character(location)),
    date_collected = as.Date(date_collected),
    event_id = paste0(
      location,
      "_",
      format(date_collected, "%Y%m%d")
    )
  ) %>%
  dplyr::relocate(
    event_id,
    .before = location
  )

# Preview the result
environmental_events_final
```

```{r}
missing_required_columns <- setdiff(
  required_environment_columns,
  names(environmental_events_final)
)

available_environment_columns <- names(
  environmental_events_final
)

missing_required_columns
available_environment_columns
```

```{r}
grep(
  "salinity|conductivity",
  names(environmental_events_final),
  value = TRUE,
  ignore.case = TRUE
)
```

```{r}
# Inspect relevant environmental columns
grep(
  "salinity|conductivity",
  names(environmental_events_final),
  value = TRUE,
  ignore.case = TRUE
)

# Rename only when the older column names actually exist
if (
  "salinity_7d_available_mean" %in% names(environmental_events_final) &&
  !"salinity_7d_spline_mean" %in% names(environmental_events_final)
) {
  environmental_events_final <-
    environmental_events_final %>%
    dplyr::rename(
      salinity_7d_spline_mean = salinity_7d_available_mean
    )
}

if (
  "conductivity_7d_available_mean" %in% names(environmental_events_final) &&
  !"conductivity_7d_spline_mean" %in% names(environmental_events_final)
) {
  environmental_events_final <-
    environmental_events_final %>%
    dplyr::rename(
      conductivity_7d_spline_mean = conductivity_7d_available_mean
    )
}

# Confirm final column names
grep(
  "salinity|conductivity",
  names(environmental_events_final),
  value = TRUE,
  ignore.case = TRUE
)
```

```{r}
setdiff(
  required_environment_columns,
  names(environmental_events_final)
)
```

```{r}
readr::write_csv(
  environmental_events_final,
  environment_file
)
```

```{r join-environment}
wgcna_environment_sophie <- event_level_eigengenes_sophie %>%
  left_join(
    environmental_events_final %>%
      select(-any_of(c("location", "month", "station", "date_collected"))),
    by = "event_id"
  )

stopifnot(
  nrow(wgcna_environment_sophie) == 6,
  n_distinct(wgcna_environment_sophie$event_id) == 6
)

readr::write_csv(
  wgcna_environment_sophie,
  file.path(output_directory, "SophieCounts_WGCNA_environment_joined.csv")
)

wgcna_environment_sophie


```

# 11. Build the expanded trait matrix

```{r trait-matrix}
# Create the trait data table
trait_data_sophie <- wgcna_environment_sophie %>%
  dplyr::transmute(
    event_id = as.character(event_id),

    temperature_7d_mean,
    DO_7d_mean,
    salinity_7d_spline_mean,
    conductivity_7d_spline_mean,
    pH_7d_available_mean,
    precipitation_7d_total_mm,

    month_Jun = as.integer(month == "Jun"),
    month_Jul = as.integer(month == "Jul"),
    month_Aug = as.integer(month == "Aug"),

    location_DAF = as.integer(location == "DAF")
  )

# Convert only numeric trait columns to a matrix
trait_matrix_sophie <- trait_data_sophie %>%
  dplyr::select(-event_id) %>%
  as.matrix()

# Assign event IDs AFTER conversion to matrix
rownames(trait_matrix_sophie) <-
  trait_data_sophie$event_id
```

```{r}
eigengene_matrix_event_sophie <-
  wgcna_environment_sophie %>%
  dplyr::select(
    dplyr::all_of(eigengene_columns)
  ) %>%
  as.matrix()

rownames(eigengene_matrix_event_sophie) <-
  as.character(
    wgcna_environment_sophie$event_id
  )
```

```{r}
trait_matrix_sophie <-
  trait_matrix_sophie[
    rownames(eigengene_matrix_event_sophie),
    ,
    drop = FALSE
  ]
```

```{r}
rownames(eigengene_matrix_event_sophie)
rownames(trait_matrix_sophie)

identical(
  rownames(eigengene_matrix_event_sophie),
  rownames(trait_matrix_sophie)
)
```

```{r}
dim(eigengene_matrix_event_sophie)
dim(trait_matrix_sophie)

sum(is.na(eigengene_matrix_event_sophie))
colSums(is.na(trait_matrix_sophie))
```

# 12. Module-trait correlations, p-values, and FDR

```{r correlation-functions}
pairwise_correlation_results <- function(eigengene_matrix, trait_matrix) {
  output <- vector(
    "list",
    ncol(eigengene_matrix) * ncol(trait_matrix)
  )

  result_number <- 1

  for (module_name in colnames(eigengene_matrix)) {
    for (trait_name in colnames(trait_matrix)) {
      module_values <- eigengene_matrix[, module_name]
      trait_values <- trait_matrix[, trait_name]

      complete_rows <- complete.cases(module_values, trait_values)
      n_complete <- sum(complete_rows)

      if (
        n_complete >= 3 &&
        sd(module_values[complete_rows]) > 0 &&
        sd(trait_values[complete_rows]) > 0
      ) {
        test_result <- cor.test(
          module_values[complete_rows],
          trait_values[complete_rows],
          method = "pearson"
        )

        correlation_value <- unname(test_result$estimate)
        p_value <- test_result$p.value
      } else {
        correlation_value <- NA_real_
        p_value <- NA_real_
      }

      output[[result_number]] <- tibble(
        module = module_name,
        trait = trait_name,
        correlation = correlation_value,
        n_events = n_complete,
        p_value = p_value
      )

      result_number <- result_number + 1
    }
  }

  bind_rows(output) %>%
    mutate(
      FDR = p.adjust(p_value, method = "BH"),
      absolute_correlation = abs(correlation)
    )
}
```

```{r module-trait-results}
module_trait_results_sophie <- pairwise_correlation_results(
  eigengene_matrix_event_sophie,
  trait_matrix_sophie
) %>%
  arrange(FDR, p_value, desc(absolute_correlation))

readr::write_csv(
  module_trait_results_sophie,
  file.path(
    output_directory,
    "SophieCounts_WGCNA_expanded_module_trait_results.csv"
  )
)

module_trait_results_sophie %>%
  slice_head(n = 25)
```

# 13. Expanded WGCNA heatmap

```{r heatmap-labels}
module_order <- colnames(eigengene_matrix_event_sophie)
trait_order <- colnames(trait_matrix_sophie)

correlation_matrix <- matrix(
  NA_real_,
  nrow = length(module_order),
  ncol = length(trait_order),
  dimnames = list(module_order, trait_order)
)

heatmap_text_matrix <- matrix(
  "",
  nrow = length(module_order),
  ncol = length(trait_order),
  dimnames = list(module_order, trait_order)
)

for (row_number in seq_len(nrow(module_trait_results_sophie))) {
  result_row <- module_trait_results_sophie[row_number, ]

  module_name <- result_row$module
  trait_name <- result_row$trait

  correlation_matrix[module_name, trait_name] <- result_row$correlation

  significance_symbol <- case_when(
    !is.na(result_row$FDR) && result_row$FDR < 0.05 ~ "**",
    !is.na(result_row$p_value) && result_row$p_value < 0.05 ~ "*",
    TRUE ~ ""
  )

  p_text <- case_when(
    is.na(result_row$p_value) ~ "p=NA",
    result_row$p_value < 0.001 ~ "p=<0.001",
    TRUE ~ paste0(
      "p=",
      formatC(result_row$p_value, format = "f", digits = 3)
    )
  )

  heatmap_text_matrix[module_name, trait_name] <- paste0(
    formatC(result_row$correlation, format = "f", digits = 2),
    significance_symbol,
    "\n",
    p_text,
    "\n",
    "n=",
    result_row$n_events
  )
}
```

```{r save-heatmap, fig.width=14, fig.height=8}
trait_display_names <- c(
  temperature_7d_mean = "Temperature",
  DO_7d_mean = "Dissolved oxygen",
  salinity_7d_spline_mean = "Salinity\n(spline)",
  conductivity_7d_spline_mean = "Conductivity\n(spline)",
  pH_7d_available_mean = "pH\n(available)",
  precipitation_7d_total_mm = "Precipitation",
  month_Jun = "June",
  month_Jul = "July",
  month_Aug = "August",
  location_DAF = "Location\nDAF vs CMT"
)

colnames(correlation_matrix) <- trait_display_names[
  colnames(correlation_matrix)
]

colnames(heatmap_text_matrix) <- colnames(correlation_matrix)

heatmap_file <- file.path(
  output_directory,
  "SophieCounts_WGCNA_expanded_module_trait_heatmap.pdf"
)

pdf(
  heatmap_file,
  width = 14,
  height = max(7, 0.72 * nrow(correlation_matrix) + 2)
)

pheatmap(
  correlation_matrix,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  display_numbers = heatmap_text_matrix,
  number_color = "black",
  fontsize_number = 9,
  fontsize_row = 11,
  fontsize_col = 11,
  angle_col = 45,
  breaks = seq(-1, 1, length.out = 101),
  main = paste0(
    "Sophie-count WGCNA: expanded module-trait correlations\n",
    "* nominal p < 0.05; ** BH-FDR < 0.05"
  ),
  border_color = "grey80"
)

dev.off()

heatmap_file
```

# 14. Save core objects

```{r save-core-objects}
core_wgcna_objects_sophie <- list(
  count_source = sophie_count_file,
  environmental_source = environment_file,
  gene_filter = "Count >= 10 in at least 3 samples",
  variable_gene_fraction = 0.05,
  selected_soft_power = soft_power,
  selected_wgcna_genes = selected_wgcna_genes,
  network = network_sophie,
  module_colors = module_colors_sophie,
  sample_metadata = sample_metadata_sophie,
  sample_level_eigengenes = sample_level_eigengenes_sophie,
  event_level_eigengenes = event_level_eigengenes_sophie,
  joined_environment = wgcna_environment_sophie,
  module_trait_results = module_trait_results_sophie
)

saveRDS(
  core_wgcna_objects_sophie,
  file.path(output_directory, "SophieCounts_WGCNA_core_objects.rds")
)
```

# 15. Final checks

```{r final-checks}
final_output_summary <- tibble(
  item = c(
    "Raw genes",
    "Filtered genes",
    "WGCNA genes",
    "RNA-seq samples",
    "Independent events",
    "Non-grey modules",
    "Environmental traits",
    "Selected soft power"
  ),
  value = c(
    nrow(count_matrix_sophie),
    nrow(dds_sophie),
    ncol(datExpr_sophie),
    nrow(datExpr_sophie),
    nrow(event_level_eigengenes_sophie),
    ncol(module_eigengenes_sophie),
    ncol(trait_matrix_sophie),
    soft_power
  )
)

final_output_summary
```

```{r regression-ready-files}
list.files(output_directory, full.names = TRUE)
```

The files most important for tomorrow's regression work are:

-   `SophieCounts_WGCNA_event_level_eigengenes.csv`
-   `SophieCounts_WGCNA_environment_joined.csv`
-   `SophieCounts_WGCNA_expanded_module_trait_results.csv`
-   `SophieCounts_WGCNA_expanded_module_trait_heatmap.pdf`
-   `SophieCounts_WGCNA_gene_modules.csv`
-   `SophieCounts_WGCNA_core_objects.rds`

```{r session-info}
writeLines(
  capture.output(sessionInfo()),
  file.path(output_directory, "SophieCounts_WGCNA_sessionInfo.txt")
)
```

**NEW WGCNA GRAPH**

```{r}
library(tidyverse)
library(patchwork)

# ------------------------------------------------------------
# 1. Preserve the module and environmental-trait ordering
# ------------------------------------------------------------

module_order <- colnames(eigengene_matrix_event_sophie)
trait_order <- colnames(trait_matrix_sophie)

trait_display_names <- c(
  temperature_7d_mean = "Temperature",
  DO_7d_mean = "Dissolved oxygen",
  salinity_7d_spline_mean = "Salinity\n(spline)",
  conductivity_7d_spline_mean = "Conductivity\n(spline)",
  pH_7d_available_mean = "pH\n(available)",
  precipitation_7d_total_mm = "Precipitation",
  month_Jun = "June",
  month_Jul = "July",
  month_Aug = "August",
  location_DAF = "Location\nDAF vs CMT"
)

# ------------------------------------------------------------
# 2. Prepare module-trait results
# ------------------------------------------------------------

plot_data <- module_trait_results_sophie %>%
  mutate(
    # Reverse levels so the first module appears at the top
    module = factor(
      module,
      levels = rev(module_order)
    ),

    trait = factor(
      trait,
      levels = trait_order
    ),

    # Label only nominally significant relationships
    significant = !is.na(p_value) & p_value < 0.05,

    p_display = case_when(
      is.na(p_value) ~ "NA",
      p_value < 0.001 ~ "<0.001",
      TRUE ~ formatC(
        p_value,
        format = "f",
        digits = 3
      )
    ),

    cell_label = if_else(
      significant,
      paste0(
        formatC(
          correlation,
          format = "f",
          digits = 2
        ),
        "\n(p=",
        p_display,
        ")"
      ),
      ""
    )
  )
```

```{r}
module_count_plot_data <- module_size_table %>%
  filter(module != "grey") %>%
  mutate(
    module_ME = paste0("ME", module),

    module_ME = factor(
      module_ME,
      levels = rev(module_order)
    )
  ) %>%
  filter(!is.na(module_ME))
```

```{r}
correlation_heatmap <- ggplot(
  plot_data,
  aes(
    x = trait,
    y = module,
    fill = correlation
  )
) +
  geom_tile(
    color = "grey85",
    linewidth = 0.6
  ) +
  geom_text(
    aes(label = cell_label),
    size = 3.8,
    fontface = "bold",
    lineheight = 0.9
  ) +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#D73027",
    midpoint = 0,
    limits = c(-1, 1),
    name = "Correlation"
  ) +
  scale_x_discrete(
    labels = trait_display_names,
    drop = FALSE
  ) +
  scale_y_discrete(
    drop = FALSE
  ) +
  labs(
    title = "WGCNA module-trait correlations",
    subtitle = "Labels shown only for nominal p < 0.05",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),

    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1,
      size = 10
    ),

    axis.text.y = element_text(
      size = 11
    ),

    plot.title = element_text(
      face = "bold"
    ),

    legend.position = "left",

    plot.margin = margin(
      t = 5,
      r = 5,
      b = 5,
      l = 5
    )
  )
```

```{r}
gene_count_plot <- ggplot(
  module_count_plot_data,
  aes(
    x = number_of_genes,
    y = module_ME
  )
) +
  geom_col(
    aes(fill = module),
    width = 0.72,
    show.legend = FALSE
  ) +
  scale_fill_identity() +
  geom_text(
    aes(
      label = scales::comma(number_of_genes)
    ),
    hjust = -0.15,
    size = 4
  ) +
  scale_y_discrete(
    drop = FALSE
  ) +
  scale_x_continuous(
    limits = c(
      0,
      max(module_count_plot_data$number_of_genes) * 1.4
    ),
    expand = expansion(mult = c(0, 0))
  ) +
  labs(
    title = "Number of genes",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),

    plot.title = element_text(
      face = "bold",
      hjust = 0.5,
      size = 12
    ),

    axis.text = element_blank(),
    axis.ticks = element_blank(),

    plot.margin = margin(
      t = 5,
      r = 10,
      b = 5,
      l = 0
    )
  )
```

```{r}
final_wgcna_plot <-
  correlation_heatmap +
  gene_count_plot +
  plot_layout(
    widths = c(5.5, 1.4)
  )

final_wgcna_plot

ggsave(
  filename = file.path(
    output_directory,
    "SophieCounts_WGCNA_significant_correlations_with_gene_counts.png"
  ),
  plot = final_wgcna_plot,
  width = 16,
  height = 7,
  dpi = 300,
  bg = "white"
)
```

```{r}
library(tidyverse)
library(patchwork)

# ------------------------------------------------------------
# 1. Preserve original order
# ------------------------------------------------------------
module_order <- colnames(eigengene_matrix_event_sophie)
trait_order  <- colnames(trait_matrix_sophie)

trait_display_names <- c(
  temperature_7d_mean = "Temperature",
  DO_7d_mean = "Dissolved oxygen",
  salinity_7d_spline_mean = "Salinity\n(spline)",
  conductivity_7d_spline_mean = "Conductivity\n(spline)",
  pH_7d_available_mean = "pH\n(available)",
  precipitation_7d_total_mm = "Precipitation",
  month_Jun = "June",
  month_Jul = "July",
  month_Aug = "August",
  location_DAF = "Location\nDAF vs CMT"
)

# ------------------------------------------------------------
# 2. Keep only traits with at least one significant correlation
# ------------------------------------------------------------
significant_traits <- module_trait_results_sophie %>%
  group_by(trait) %>%
  summarise(
    has_sig = any(!is.na(p_value) & p_value < 0.05),
    .groups = "drop"
  ) %>%
  filter(has_sig) %>%
  pull(trait)

# preserve the original left-to-right order
significant_traits <- trait_order[trait_order %in% significant_traits]

# ------------------------------------------------------------
# 3. Prepare plotting data
# ------------------------------------------------------------
plot_data <- module_trait_results_sophie %>%
  filter(trait %in% significant_traits) %>%
  mutate(
    module = factor(module, levels = rev(module_order)),
    trait  = factor(trait, levels = significant_traits),
    significant = !is.na(p_value) & p_value < 0.05,

    p_display = case_when(
      is.na(p_value) ~ "NA",
      p_value < 0.001 ~ "<0.001",
      TRUE ~ formatC(p_value, format = "f", digits = 3)
    ),

    cell_label = if_else(
      significant,
      paste0(
        formatC(correlation, format = "f", digits = 2),
        "\n(p=", p_display, ")"
      ),
      ""
    )
  )
```

```{r}
module_count_plot_data <- module_size_table %>%
  filter(module != "grey") %>%
  mutate(
    module_ME = paste0("ME", module),
    module_ME = factor(module_ME, levels = rev(module_order))
  ) %>%
  filter(!is.na(module_ME))
```

```{r}
correlation_heatmap <- ggplot(
  plot_data,
  aes(
    x = trait,
    y = module,
    fill = correlation
  )
) +
  geom_tile(
    color = "grey85",
    linewidth = 0.6
  ) +
  geom_text(
    aes(label = cell_label),
    size = 3.8,
    fontface = "bold",
    lineheight = 0.9
  ) +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#D73027",
    midpoint = 0,
    limits = c(-1, 1),
    name = "Correlation"
  ) +
  scale_x_discrete(
    labels = trait_display_names[significant_traits],
    drop = FALSE
  ) +
  scale_y_discrete(drop = FALSE) +
  labs(
    title = "WGCNA module-trait correlations",
    subtitle = "Only traits with at least one nominally significant correlation shown",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1,
      size = 10
    ),
    axis.text.y = element_text(size = 11),
    plot.title = element_text(face = "bold"),
    legend.position = "left",
    plot.margin = margin(5, 5, 5, 5)
  )
```

```{r}
gene_count_plot <- ggplot(
  module_count_plot_data,
  aes(
    x = number_of_genes,
    y = module_ME
  )
) +
  geom_col(
    aes(fill = module),
    width = 0.72,
    show.legend = FALSE
  ) +
  scale_fill_identity() +
  geom_text(
    aes(label = scales::comma(number_of_genes)),
    hjust = -0.15,
    size = 4
  ) +
  scale_y_discrete(drop = FALSE) +
  scale_x_continuous(
    limits = c(0, max(module_count_plot_data$number_of_genes) * 1.4),
    expand = expansion(mult = c(0, 0))
  ) +
  labs(
    title = "Number of genes",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5, size = 12),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    plot.margin = margin(5, 10, 5, 0)
  )
```

```{r}
final_wgcna_plot <-
  correlation_heatmap +
  gene_count_plot +
  plot_layout(widths = c(5.5, 1.4))

final_wgcna_plot

ggsave(
  filename = file.path(
    output_directory,
    "SophieCounts_WGCNA_significant_traits_only.png"
  ),
  plot = final_wgcna_plot,
  width = 13,
  height = 7,
  dpi = 300,
  bg = "white"
)
```

```{r}
focus_modules <- c("MEblack", "MEyellow", "MEgreen")
```

```{r}
plot_data <- module_trait_results_sophie %>%
  filter(trait %in% significant_traits) %>%
  mutate(
    module = factor(module, levels = rev(module_order)),
    trait  = factor(trait, levels = significant_traits),

    highlight_module = module %in% focus_modules,

    significant = !is.na(p_value) & p_value < 0.05,

    p_display = case_when(
      is.na(p_value) ~ "NA",
      p_value < 0.001 ~ "<0.001",
      TRUE ~ formatC(p_value, format = "f", digits = 3)
    ),

    cell_label = if_else(
      significant & highlight_module,
      paste0(
        formatC(correlation, format = "f", digits = 2),
        "\n(p=", p_display, ")"
      ),
      ""
    ),

    # Only highlighted modules keep correlation colors;
    # all others become NA and will be drawn in grey
    fill_value = if_else(
      highlight_module,
      correlation,
      NA_real_
    )
  )
```

```{r}
correlation_heatmap <- ggplot(
  plot_data,
  aes(
    x = trait,
    y = module,
    fill = fill_value
  )
) +
  geom_tile(
    color = "grey80",
    linewidth = 0.6
  ) +
  geom_text(
    aes(label = cell_label),
    size = 3.8,
    fontface = "bold",
    lineheight = 0.9
  ) +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#D73027",
    midpoint = 0,
    limits = c(-1, 1),
    na.value = "grey85",
    name = "Correlation"
  ) +
  scale_x_discrete(
    labels = trait_display_names[significant_traits],
    drop = FALSE
  ) +
  scale_y_discrete(drop = FALSE) +
  labs(
    title = "WGCNA module-trait correlations",
    subtitle = "Highlighted modules: MEblack, MEyellow, and MEgreen",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1,
      size = 10
    ),
    axis.text.y = element_text(size = 11),
    plot.title = element_text(face = "bold"),
    legend.position = "left",
    plot.margin = margin(5, 5, 5, 5)
  )
```

```{r}
correlation_heatmap <- ggplot(
  plot_data,
  aes(
    x = trait,
    y = module,
    fill = fill_value
  )
) +
  geom_tile(
    color = "grey80",
    linewidth = 0.6
  ) +
  geom_text(
    aes(label = cell_label),
    size = 3.8,
    fontface = "bold",
    lineheight = 0.9
  ) +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#D73027",
    midpoint = 0,
    limits = c(-1, 1),
    na.value = "grey85",
    name = "Correlation"
  ) +
  scale_x_discrete(
    labels = trait_display_names[significant_traits],
    drop = FALSE
  ) +
  scale_y_discrete(drop = FALSE) +
  labs(
    title = "WGCNA module-trait correlations",
    subtitle = "Highlighted modules: MEblack, MEyellow, and MEgreen",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1,
      size = 10
    ),
    axis.text.y = element_text(size = 11),
    plot.title = element_text(face = "bold"),
    legend.position = "left",
    plot.margin = margin(5, 5, 5, 5)
  )
```

```{r}
module_count_plot_data <- module_size_table %>%
  filter(module != "grey") %>%
  mutate(
    module_ME = paste0("ME", module),
    module_ME = factor(module_ME, levels = rev(module_order)),
    bar_fill = if_else(
      module_ME %in% focus_modules,
      module,
      "grey80"
    )
  ) %>%
  filter(!is.na(module_ME))
```

```{r}
gene_count_plot <- ggplot(
  module_count_plot_data,
  aes(
    x = number_of_genes,
    y = module_ME
  )
) +
  geom_col(
    aes(fill = bar_fill),
    width = 0.72,
    show.legend = FALSE
  ) +
  scale_fill_identity() +
  geom_text(
    aes(label = scales::comma(number_of_genes)),
    hjust = -0.15,
    size = 4
  ) +
  scale_y_discrete(drop = FALSE) +
  scale_x_continuous(
    limits = c(0, max(module_count_plot_data$number_of_genes) * 1.4),
    expand = expansion(mult = c(0, 0))
  ) +
  labs(
    title = "Number of genes",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5, size = 12),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    plot.margin = margin(5, 10, 5, 0)
  )
```

```{r}
final_wgcna_plot <-
  correlation_heatmap +
  gene_count_plot +
  patchwork::plot_layout(
    widths = c(5.5, 1.4)
  )

print(final_wgcna_plot)
```
