---
title: "Linear Regression"
output: html_notebook
---

**LINEAR REGRESSION ON THE MODULE**

Step 1: Confirm the regression dataset

```{r}
# Load required libraries
library(tidyverse)
library(broom)

# Check that the event-level modeling table exists
exists("big_environment_traits")

# Inspect its size and structure
dim(big_environment_traits)
names(big_environment_traits)
dplyr::glimpse(big_environment_traits)
```

```{r}
# Inspect the variables needed for our planned regression models
big_environment_traits %>%
  dplyr::select(
    dplyr::any_of(
      c(
        "event_id",
        "location",
        "month",
        "MEblack",
        "salinity_7d_spline_mean",
        "conductivity_7d_spline_mean",
        "MEyellow",
        "DO_7d_mean",
        "MEgreen"
      )
    )
  )
```

Step 2: Locate the module eigengene table

```{r}
# Find objects containing module eigengene columns such as MEblack
object_names <- ls()

objects_with_ME <- object_names[
  vapply(
    object_names,
    function(object_name) {
      object <- get(object_name, envir = .GlobalEnv)
      column_names <- colnames(object)

      (is.data.frame(object) || is.matrix(object)) &&
        !is.null(column_names) &&
        any(grepl("^ME", column_names))
    },
    logical(1)
  )
]

# Show the matching object names
objects_with_ME

# Summarize their dimensions and eigengene columns
purrr::map_dfr(
  objects_with_ME,
  function(object_name) {
    object <- get(object_name, envir = .GlobalEnv)

    tibble::tibble(
      object = object_name,
      rows = nrow(object),
      columns = ncol(object),
      ME_columns = paste(
        grep("^ME", colnames(object), value = TRUE),
        collapse = ", "
      )
    )
  }
)
```

Step 3: Identify the correct six-event eigengene object

```{r}
candidate_objects <- c(
  "event_level_eigengenes_sophie",
  "eigengene_matrix_event_sophie",
  "event_module_eigengenes",
  "big_wgcna_event_data"
)

for (object_name in candidate_objects) {

  if (exists(object_name)) {

    object <- get(object_name)

    cat("\n-----------------------------------\n")
    cat("Object:", object_name, "\n")
    cat("Class:", paste(class(object), collapse = ", "), "\n")
    cat(
      "Dimensions:",
      paste(dim(object), collapse = " x "),
      "\n"
    )

    cat("Module eigengene columns:\n")
    print(
      colnames(object)[
        startsWith(colnames(object), "ME")
      ]
    )

    cat("Identifier columns:\n")
    print(
      intersect(
        c("event_id", "location", "month"),
        colnames(object)
      )
    )

    cat("First row names:\n")
    print(head(rownames(object)))
  }
}
```

Step 4: Build the regression-ready table

```{r}
# Create one clean table for the planned regression analyses
regression_data <- big_wgcna_event_data %>%
  dplyr::transmute(
    event_id,
    location = factor(
      location,
      levels = c("CMT", "DAF")
    ),
    month = factor(
      month,
      levels = c("Jun", "Jul", "Aug")
    ),

    # Primary MEblack analysis
    MEblack,
    salinity_7d_spline_mean,
    conductivity_7d_spline_mean,

    # Secondary analyses
    MEyellow,
    DO_7d_mean,
    MEgreen
  ) %>%
  dplyr::arrange(month, location)

# Display the complete regression dataset
regression_data
```

```{r}
# Confirm dimensions and event uniqueness
dim(regression_data)

regression_data %>%
  dplyr::summarise(
    rows = dplyr::n(),
    unique_events = dplyr::n_distinct(event_id),
    duplicate_events = dplyr::n() - dplyr::n_distinct(event_id)
  )

# Count missing values in every variable
regression_data %>%
  dplyr::summarise(
    dplyr::across(
      dplyr::everything(),
      ~ sum(is.na(.))
    )
  )
```

Step 5: Plot MEblack against salinity

```{r}
# Labeled six-event scatterplot:
# MEblack versus 7-day spline-estimated salinity

MEblack_salinity_plot <- regression_data %>%
  ggplot(
    aes(
      x = salinity_7d_spline_mean,
      y = MEblack
    )
  ) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = TRUE,
    linewidth = 0.8
  ) +
  geom_point(
    aes(
      shape = location,
      fill = month
    ),
    size = 4
  ) +
  geom_text(
    aes(label = event_id),
    nudge_y = 0.025,
    check_overlap = TRUE,
    size = 3
  ) +
  labs(
    title = "MEblack association with salinity",
    subtitle = "Six independent location–month collection events",
    x = "7-day mean salinity (spline-estimated)",
    y = "MEblack module eigengene",
    shape = "Location",
    fill = "Month"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )

MEblack_salinity_plot
```

lower salinity corresponds to higher `MEblack` eigengene values.

Step 6: Fit the primary salinity model

```{r}

# Primary regression model
MEblack_salinity_model <- lm(
  MEblack ~ salinity_7d_spline_mean,
  data = regression_data
)

# Standard model summary
summary(MEblack_salinity_model)
```

```{r}
# Coefficient estimate, p-value, and 95% confidence interval
MEblack_salinity_coefficients <- broom::tidy(
  MEblack_salinity_model,
  conf.int = TRUE,
  conf.level = 0.95
)

MEblack_salinity_coefficients
```

```{r}
# Model-level fit statistics
MEblack_salinity_fit <- broom::glance(
  MEblack_salinity_model
)

# Calculate AICc for the small sample size
n_model <- nobs(MEblack_salinity_model)
k_model <- attr(logLik(MEblack_salinity_model), "df")

MEblack_salinity_AICc <-
  AIC(MEblack_salinity_model) +
  (2 * k_model * (k_model + 1)) /
  (n_model - k_model - 1)

MEblack_salinity_fit %>%
  dplyr::transmute(
    n = nobs,
    R_squared = r.squared,
    adjusted_R_squared = adj.r.squared,
    residual_standard_error = sigma,
    F_statistic = statistic,
    model_p_value = p.value,
    AIC = AIC,
    AICc = MEblack_salinity_AICc
  )
```

So, within these six collection events, a one-unit increase in the 7-day mean salinity is associated with an estimated **0.0747-unit decrease in the MEblack eigengene value**.

Salinity explains approximately **99.2% of the observed variation in MEblack across these six events**.

Step 7: Check influence and outliers

```{r}
# Influence diagnostics for the MEblack–salinity model

n_events <- nobs(MEblack_salinity_model)
n_parameters <- length(coef(MEblack_salinity_model))

cook_cutoff <- 4 / n_events
leverage_cutoff <- 2 * n_parameters / n_events

MEblack_salinity_diagnostics <- regression_data %>%
  dplyr::mutate(
    fitted_value = fitted(MEblack_salinity_model),
    residual = resid(MEblack_salinity_model),
    standardized_residual = rstandard(MEblack_salinity_model),
    studentized_residual = rstudent(MEblack_salinity_model),
    leverage = hatvalues(MEblack_salinity_model),
    cooks_distance = cooks.distance(MEblack_salinity_model)
  ) %>%
  dplyr::mutate(
    large_residual = abs(studentized_residual) > 2,
    high_leverage = leverage > leverage_cutoff,
    high_cooks_distance = cooks_distance > cook_cutoff
  ) %>%
  dplyr::select(
    event_id,
    location,
    month,
    salinity_7d_spline_mean,
    MEblack,
    fitted_value,
    residual,
    studentized_residual,
    leverage,
    cooks_distance,
    large_residual,
    high_leverage,
    high_cooks_distance
  ) %>%
  dplyr::arrange(dplyr::desc(cooks_distance))

MEblack_salinity_diagnostics

cat("Cook's distance reference cutoff:", cook_cutoff, "\n")
cat("Leverage reference cutoff:", leverage_cutoff, "\n")
```

Step 8: Leave one event out

```{r}
# Full-model slope for comparison
full_salinity_slope <- unname(
  coef(MEblack_salinity_model)[["salinity_7d_spline_mean"]]
)

# Refit the model six times, omitting one event each time
MEblack_salinity_LOEO <- purrr::map_dfr(
  regression_data$event_id,
  function(event_removed) {

    loo_data <- regression_data %>%
      dplyr::filter(event_id != event_removed)

    loo_model <- lm(
      MEblack ~ salinity_7d_spline_mean,
      data = loo_data
    )

    coefficient_results <- broom::tidy(
      loo_model,
      conf.int = TRUE,
      conf.level = 0.95
    ) %>%
      dplyr::filter(
        term == "salinity_7d_spline_mean"
      )

    fit_results <- broom::glance(loo_model)

    n_loo <- nobs(loo_model)
    k_loo <- attr(logLik(loo_model), "df")

    AICc_loo <-
      AIC(loo_model) +
      (2 * k_loo * (k_loo + 1)) /
      (n_loo - k_loo - 1)

    tibble::tibble(
      omitted_event = event_removed,
      n = n_loo,
      salinity_slope = coefficient_results$estimate,
      conf_low = coefficient_results$conf.low,
      conf_high = coefficient_results$conf.high,
      p_value = coefficient_results$p.value,
      R_squared = fit_results$r.squared,
      adjusted_R_squared = fit_results$adj.r.squared,
      AICc = AICc_loo,
      absolute_slope_change_percent =
        100 * abs(
          (coefficient_results$estimate - full_salinity_slope) /
            full_salinity_slope
        ),
      slope_remains_negative =
        coefficient_results$estimate < 0,
      CI_excludes_zero =
        coefficient_results$conf.high < 0
    )
  }
)

MEblack_salinity_LOEO
```

August DAF event strengthens the pattern, but it is **not creating the relationship by itself**.

Step 9: Visualize MEblack across months

```{r}
MEblack_month_plot <- regression_data %>%
  ggplot(
    aes(
      x = month,
      y = MEblack
    )
  ) +
  geom_line(
    aes(
      group = location,
      linetype = location
    ),
    linewidth = 0.8
  ) +
  geom_point(
    aes(
      shape = location
    ),
    size = 4
  ) +
  geom_text(
    aes(label = event_id),
    nudge_y = 0.025,
    check_overlap = TRUE,
    size = 3
  ) +
  stat_summary(
    aes(group = 1),
    fun = mean,
    geom = "point",
    shape = 18,
    size = 5
  ) +
  labs(
    title = "MEblack eigengene expression across months",
    subtitle = "Points represent six independent location–month events; diamonds show monthly means",
    x = "Collection month",
    y = "MEblack module eigengene",
    shape = "Location",
    linetype = "Location"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )

MEblack_month_plot
```

Step 10: Fit MEblack \~ month

```{r}
# Temporal comparison model
MEblack_month_model <- lm(
  MEblack ~ month,
  data = regression_data
)

summary(MEblack_month_model)
```

```{r}
# Month coefficients, p-values, and 95% confidence intervals
MEblack_month_coefficients <- broom::tidy(
  MEblack_month_model,
  conf.int = TRUE,
  conf.level = 0.95
)

MEblack_month_coefficients
```

```{r}
# Overall month-model fit
MEblack_month_fit <- broom::glance(
  MEblack_month_model
)

# Small-sample corrected AIC
n_month <- nobs(MEblack_month_model)
k_month <- attr(logLik(MEblack_month_model), "df")

MEblack_month_AICc <-
  AIC(MEblack_month_model) +
  (2 * k_month * (k_month + 1)) /
  (n_month - k_month - 1)

MEblack_month_fit %>%
  dplyr::transmute(
    n = nobs,
    R_squared = r.squared,
    adjusted_R_squared = adj.r.squared,
    residual_standard_error = sigma,
    F_statistic = statistic,
    model_p_value = p.value,
    AIC = AIC,
    AICc = MEblack_month_AICc
  )
```

Step 11: Directly compare salinity and month

```{r}
# Compare the two candidate explanations for MEblack
MEblack_model_comparison <- tibble::tibble(
  model = c(
    "Salinity",
    "Month"
  ),
  formula = c(
    "MEblack ~ salinity_7d_spline_mean",
    "MEblack ~ month"
  ),
  n = c(
    nobs(MEblack_salinity_model),
    nobs(MEblack_month_model)
  ),
  estimated_parameters = c(
    attr(logLik(MEblack_salinity_model), "df"),
    attr(logLik(MEblack_month_model), "df")
  ),
  R_squared = c(
    summary(MEblack_salinity_model)$r.squared,
    summary(MEblack_month_model)$r.squared
  ),
  adjusted_R_squared = c(
    summary(MEblack_salinity_model)$adj.r.squared,
    summary(MEblack_month_model)$adj.r.squared
  ),
  residual_standard_error = c(
    sigma(MEblack_salinity_model),
    sigma(MEblack_month_model)
  ),
  model_p_value = c(
    broom::glance(MEblack_salinity_model)$p.value,
    broom::glance(MEblack_month_model)$p.value
  ),
  AIC = c(
    AIC(MEblack_salinity_model),
    AIC(MEblack_month_model)
  ),
  AICc = c(
    MEblack_salinity_AICc,
    MEblack_month_AICc
  )
) %>%
  dplyr::mutate(
    delta_AICc = AICc - min(AICc)
  ) %>%
  dplyr::arrange(delta_AICc)

MEblack_model_comparison
```

Step 12: Check conductivity against salinity

```{r}
# Quantify the relationship between salinity and conductivity
salinity_conductivity_correlation <- regression_data %>%
  dplyr::summarise(
    n = sum(
      complete.cases(
        salinity_7d_spline_mean,
        conductivity_7d_spline_mean
      )
    ),
    Pearson_r = cor(
      salinity_7d_spline_mean,
      conductivity_7d_spline_mean,
      use = "complete.obs",
      method = "pearson"
    ),
    Spearman_r = cor(
      salinity_7d_spline_mean,
      conductivity_7d_spline_mean,
      use = "complete.obs",
      method = "spearman"
    )
  )

salinity_conductivity_correlation
```

```{r}
salinity_conductivity_plot <- regression_data %>%
  ggplot(
    aes(
      x = salinity_7d_spline_mean,
      y = conductivity_7d_spline_mean
    )
  ) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = TRUE,
    linewidth = 0.8
  ) +
  geom_point(
    aes(shape = location),
    size = 4
  ) +
  geom_text(
    aes(label = event_id),
    nudge_y = 250,
    check_overlap = TRUE,
    size = 3
  ) +
  labs(
    title = "Relationship between salinity and conductivity",
    subtitle = "Six independent location–month collection events",
    x = "7-day mean salinity (spline-estimated)",
    y = "7-day mean conductivity (spline-estimated)",
    shape = "Location"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )

salinity_conductivity_plot
```

Step 13: Fit the conductivity sensitivity model

```{r}
# Rescale conductivity for an interpretable coefficient
regression_data <- regression_data %>%
  dplyr::mutate(
    conductivity_7d_spline_per_1000 =
      conductivity_7d_spline_mean / 1000
  )

# Conductivity sensitivity model
MEblack_conductivity_model <- lm(
  MEblack ~ conductivity_7d_spline_per_1000,
  data = regression_data
)

summary(MEblack_conductivity_model)
```

```{r}
MEblack_conductivity_coefficients <- broom::tidy(
  MEblack_conductivity_model,
  conf.int = TRUE,
  conf.level = 0.95
)

MEblack_conductivity_coefficients
```

```{r}
MEblack_conductivity_fit <- broom::glance(
  MEblack_conductivity_model
)

n_conductivity <- nobs(MEblack_conductivity_model)
k_conductivity <- attr(
  logLik(MEblack_conductivity_model),
  "df"
)

MEblack_conductivity_AICc <-
  AIC(MEblack_conductivity_model) +
  (2 * k_conductivity * (k_conductivity + 1)) /
  (n_conductivity - k_conductivity - 1)

MEblack_conductivity_fit %>%
  dplyr::transmute(
    n = nobs,
    R_squared = r.squared,
    adjusted_R_squared = adj.r.squared,
    residual_standard_error = sigma,
    F_statistic = statistic,
    model_p_value = p.value,
    AIC = AIC,
    AICc = MEblack_conductivity_AICc
  )
```

Step 14: Fit salinity plus location

```{r}

# Salinity model adjusted for location
MEblack_salinity_location_model <- lm(
  MEblack ~ salinity_7d_spline_mean + location,
  data = regression_data
)

summary(MEblack_salinity_location_model)
```

```{r}
MEblack_salinity_location_coefficients <- broom::tidy(
  MEblack_salinity_location_model,
  conf.int = TRUE,
  conf.level = 0.95
)

MEblack_salinity_location_coefficients
```

After accounting for salinity, there was no detectable additional association between location and MEblack.

Step 15: Formally compare the nested models

```{r}
# Does adding location significantly improve the salinity-only model?
MEblack_salinity_location_test <- anova(
  MEblack_salinity_model,
  MEblack_salinity_location_model
)

MEblack_salinity_location_test
```

```{r}
# Fit statistics for the salinity + location model
MEblack_salinity_location_fit <- broom::glance(
  MEblack_salinity_location_model
)

n_sal_location <- nobs(MEblack_salinity_location_model)

k_sal_location <- attr(
  logLik(MEblack_salinity_location_model),
  "df"
)

MEblack_salinity_location_AICc <-
  AIC(MEblack_salinity_location_model) +
  (
    2 * k_sal_location * (k_sal_location + 1)
  ) /
  (
    n_sal_location - k_sal_location - 1
  )

# Compare the simple and adjusted models
MEblack_salinity_location_comparison <- tibble::tibble(
  model = c(
    "Salinity only",
    "Salinity + location"
  ),
  R_squared = c(
    summary(MEblack_salinity_model)$r.squared,
    summary(MEblack_salinity_location_model)$r.squared
  ),
  adjusted_R_squared = c(
    summary(MEblack_salinity_model)$adj.r.squared,
    summary(MEblack_salinity_location_model)$adj.r.squared
  ),
  residual_standard_error = c(
    sigma(MEblack_salinity_model),
    sigma(MEblack_salinity_location_model)
  ),
  AIC = c(
    AIC(MEblack_salinity_model),
    AIC(MEblack_salinity_location_model)
  ),
  AICc = c(
    MEblack_salinity_AICc,
    MEblack_salinity_location_AICc
  )
) %>%
  dplyr::mutate(
    delta_AICc = AICc - min(AICc)
  )

MEblack_salinity_location_comparison
```

Step 16: Plot MEyello against dissolved oxygen

```{r}
# Labeled six-event scatterplot:
# MEyellow versus 7-day mean dissolved oxygen

MEyellow_DO_plot <- regression_data %>%
  ggplot(
    aes(
      x = DO_7d_mean,
      y = MEyellow
    )
  ) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = TRUE,
    linewidth = 0.8
  ) +
  geom_point(
    aes(
      shape = location,
      fill = month
    ),
    size = 4
  ) +
  geom_text(
    aes(label = event_id),
    nudge_y = 0.025,
    check_overlap = TRUE,
    size = 3
  ) +
  labs(
    title = "MEyellow association with dissolved oxygen",
    subtitle = "Six independent location–month collection events",
    x = "7-day mean dissolved oxygen",
    y = "MEyellow module eigengene",
    shape = "Location",
    fill = "Month"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )

MEyellow_DO_plot
```

Step 17: Plot MEgreen across months

```{r}
# Examine the temporal pattern in MEgreen

MEgreen_month_plot <- regression_data %>%
  ggplot(
    aes(
      x = month,
      y = MEgreen
    )
  ) +
  geom_line(
    aes(
      group = location,
      linetype = location
    ),
    linewidth = 0.8
  ) +
  geom_point(
    aes(shape = location),
    size = 4
  ) +
  geom_text(
    aes(label = event_id),
    nudge_y = 0.025,
    check_overlap = TRUE,
    size = 3
  ) +
  stat_summary(
    aes(group = 1),
    fun = mean,
    geom = "point",
    shape = 18,
    size = 5
  ) +
  labs(
    title = "MEgreen eigengene expression across months",
    subtitle = paste(
      "Points represent six independent location–month events;",
      "diamonds show monthly means"
    ),
    x = "Collection month",
    y = "MEgreen module eigengene",
    shape = "Location",
    linetype = "Location"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )

MEgreen_month_plot
```

```{r}
MEgreen_month_model <- lm(
  MEgreen ~ month,
  data = regression_data
)

summary(MEgreen_month_model)
```

```{r}
MEgreen_month_coefficients <- broom::tidy(
  MEgreen_month_model,
  conf.int = TRUE,
  conf.level = 0.95
)

MEgreen_month_coefficients
```

```{r}
MEgreen_month_fit <- broom::glance(
  MEgreen_month_model
)

n_green <- nobs(MEgreen_month_model)

k_green <- attr(
  logLik(MEgreen_month_model),
  "df"
)

MEgreen_month_AICc <-
  AIC(MEgreen_month_model) +
  (
    2 * k_green * (k_green + 1)
  ) /
  (
    n_green - k_green - 1
  )

MEgreen_month_fit %>%
  dplyr::transmute(
    n = nobs,
    R_squared = r.squared,
    adjusted_R_squared = adj.r.squared,
    residual_standard_error = sigma,
    F_statistic = statistic,
    model_p_value = p.value,
    AIC = AIC,
    AICc = MEgreen_month_AICc
  )
```

Step 18: MEturquoise with pH

```{r}
# Build the complete-case dataset for MEturquoise and pH
MEturquoise_pH_data <- big_wgcna_event_data %>%
  dplyr::select(
    event_id,
    location,
    month,
    MEturquoise,
    pH_7d_available_mean
  ) %>%
  tidyr::drop_na(
    MEturquoise,
    pH_7d_available_mean
  )

MEturquoise_pH_data

# Confirm which event is missing pH
big_wgcna_event_data %>%
  dplyr::filter(is.na(pH_7d_available_mean)) %>%
  dplyr::select(
    event_id,
    location,
    month,
    pH_7d_available_mean
  )
```

```{r}
MEturquoise_pH_plot <- MEturquoise_pH_data %>%
  ggplot(
    aes(
      x = pH_7d_available_mean,
      y = MEturquoise
    )
  ) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = TRUE,
    linewidth = 0.8
  ) +
  geom_point(
    aes(
      shape = location
    ),
    size = 4
  ) +
  geom_text(
    aes(label = event_id),
    nudge_y = 0.018,
    check_overlap = TRUE,
    size = 3
  ) +
  labs(
    title = "MEturquoise association with pH",
    subtitle = "Five complete location–month events; CMT August is missing pH",
    x = "7-day mean pH",
    y = "MEturquoise module eigengene",
    shape = "Location"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )

MEturquoise_pH_plot
```

**GO ENRICHMENT ON MEBlack**

Step 1: A few checks

```{r}
# Find likely gene-to-module assignment objects
grep(
  pattern = "module|color|gene",
  x = ls(),
  ignore.case = TRUE,
  value = TRUE
)
```

```{r}
# Check whether common WGCNA module-label objects exist
c(
  module_colors = exists("module_colors"),
  moduleColors = exists("moduleColors"),
  mergedColors = exists("mergedColors"),
  dynamicColors = exists("dynamicColors"),
  gene_modules = exists("gene_modules"),
  module_assignments = exists("module_assignments")
)
```

```{r}
# Inspect the existing gene-to-module assignment object
class(gene_modules)
dim(gene_modules)
names(gene_modules)

head(gene_modules, 10)

str(gene_modules)
```

```{r}
# Inspect the module-color object
class(module_colors)
length(module_colors)

head(module_colors, 10)
head(names(module_colors), 10)

table(module_colors, useNA = "ifany")
```

```{r}
library(tidyverse)
library(goseq)

gene2cat_path <- paste0(
  "/work/clh162/Data+/Sophie/",
  "2025-2026-team-code-sac165/",
  "Data+_team/Sophie/gene2cat.rds"
)

# Confirm that the file exists
file.exists(gene2cat_path)

# Load the gene-to-GO mapping
gene2cat <- readRDS(gene2cat_path)

# Inspect its structure
class(gene2cat)
dim(gene2cat)
names(gene2cat)
head(gene2cat, 10)
str(gene2cat)
```

```{r}
# Convert both mapping columns to character
gene2cat <- gene2cat %>%
  dplyr::mutate(
    dplyr::across(
      dplyr::everything(),
      as.character
    )
  )

# GOseq expects the first column to contain gene IDs
annotation_gene_ids <- unique(gene2cat[[1]])
annotation_GO_ids <- unique(gene2cat[[2]])

gene2cat_coverage <- tibble::tibble(
  total_gene_GO_mappings = nrow(gene2cat),
  unique_annotated_genes = length(annotation_gene_ids),
  unique_GO_terms = length(annotation_GO_ids),

  WGCNA_genes_with_GO = sum(
    module_gene_table_final$gene_id %in%
      annotation_gene_ids
  ),

  total_WGCNA_genes = nrow(
    module_gene_table_final
  ),

  MEblack_genes_with_GO = sum(
    MEblack_genes$gene_id %in%
      annotation_gene_ids
  ),

  total_MEblack_genes = nrow(
    MEblack_genes
  )
)

gene2cat_coverage
```

```{r}
diamond_path <- paste0(
  "/work/clh162/Data+/Sophie/",
  "2025-2026-team-code-sac165/",
  "Data+_team/Sophie/diamond_result_matches.tsv"
)

file.exists(diamond_path)

diamond_raw <- readr::read_tsv(
  diamond_path,
  col_names = FALSE,
  show_col_types = FALSE
)

class(diamond_raw)
dim(diamond_raw)
head(diamond_raw, 10)
```

```{r}
# Check which DIAMOND column contains the current evm.TU gene IDs
diamond_column_overlap <- purrr::map_dfr(
  names(diamond_raw),
  function(column_name) {
    values <- as.character(diamond_raw[[column_name]])

    tibble::tibble(
      column = column_name,
      WGCNA_gene_matches = sum(
        unique(module_gene_table_final$gene_id) %in% values
      ),
      MEblack_gene_matches = sum(
        unique(MEblack_genes$gene_id) %in% values
      ),
      GO_annotation_gene_matches = sum(
        unique(values) %in% annotation_gene_ids
      ),
      example_value = dplyr::first(values)
    )
  }
)

diamond_column_overlap
```

Step 2: Extract MEBlack Genes

```{r}
# Convert the final named WGCNA module vector into a clean table
module_gene_table_final <- tibble::tibble(
  gene_id = names(module_colors),
  module = as.character(module_colors)
)

# Check that every gene appears once
module_gene_table_final %>%
  dplyr::summarise(
    total_rows = dplyr::n(),
    unique_gene_ids = dplyr::n_distinct(gene_id),
    duplicate_gene_ids =
      dplyr::n() - dplyr::n_distinct(gene_id)
  )

# Confirm module sizes
module_gene_table_final %>%
  dplyr::count(module, sort = TRUE)

# Extract the black-module genes
MEblack_genes <- module_gene_table_final %>%
  dplyr::filter(module == "black") %>%
  dplyr::arrange(gene_id)

# Confirm and preview
nrow(MEblack_genes)
head(MEblack_genes, 15)
```

Step 3: Create a bridge between Diamond IDs to WGCNA IDs

```{r}
# Create a candidate bridge from the DIAMOND IDs to our WGCNA IDs
diamond_bridge <- diamond_raw %>%
  dplyr::transmute(
    diamond_gene_id = X1,

    # Test whether only the prefix differs
    gene_id = stringr::str_replace(
      X1,
      "^evm\\.model\\.",
      "evm.TU."
    ),

    protein_accession = X2,
    percent_identity = X3,
    alignment_length = X4,
    e_value = X11,
    bit_score = X12
  )

# Check overlap after changing evm.model.* to evm.TU.*
diamond_prefix_check <- tibble::tibble(
  diamond_rows = nrow(diamond_bridge),

  direct_WGCNA_matches = sum(
    diamond_bridge$diamond_gene_id %in%
      module_gene_table_final$gene_id
  ),

  normalized_WGCNA_matches = sum(
    diamond_bridge$gene_id %in%
      module_gene_table_final$gene_id
  ),

  normalized_MEblack_matches = sum(
    diamond_bridge$gene_id %in%
      MEblack_genes$gene_id
  ),

  total_WGCNA_genes = nrow(module_gene_table_final),
  total_MEblack_genes = nrow(MEblack_genes)
)

diamond_prefix_check
```

```{r}
# Preview MEblack genes that successfully match DIAMOND proteins
MEblack_diamond_matches <- MEblack_genes %>%
  dplyr::left_join(
    diamond_bridge,
    by = "gene_id"
  )

MEblack_diamond_matches %>%
  dplyr::select(
    gene_id,
    protein_accession,
    percent_identity,
    e_value,
    bit_score
  ) %>%
  head(15)

# Count matched and unmatched MEblack genes
MEblack_diamond_matches %>%
  dplyr::summarise(
    total_MEblack_genes = dplyr::n(),
    matched_to_protein =
      sum(!is.na(protein_accession)),
    unmatched =
      sum(is.na(protein_accession))
  )
```

```{r}
protein_gene_path <- paste0(
  "/work/clh162/Data+/Sophie/",
  "2025-2026-team-code-sac165/",
  "Data+_team/Sophie/protein_genesymbol.tsv"
)

file.exists(protein_gene_path)

protein_and_gene <- readr::read_tsv(
  protein_gene_path,
  col_names = c(
    "protein_id",
    "gene_symbol"
  ),
  show_col_types = FALSE
)

class(protein_and_gene)
dim(protein_and_gene)
head(protein_and_gene, 10)
str(protein_and_gene)
```

```{r}
protein_gene_coverage <- tibble::tibble(
  total_rows = nrow(protein_and_gene),

  unique_proteins =
    dplyr::n_distinct(protein_and_gene$protein_id),

  unique_gene_symbols =
    dplyr::n_distinct(protein_and_gene$gene_symbol),

  diamond_proteins_matched = sum(
    unique(diamond_bridge$protein_accession) %in%
      protein_and_gene$protein_id,
    na.rm = TRUE
  ),

  gene_symbols_in_GO_table = sum(
    unique(protein_and_gene$gene_symbol) %in%
      annotation_gene_ids,
    na.rm = TRUE
  )
)

protein_gene_coverage
```

```{r}
MEblack_protein_gene_preview <- MEblack_diamond_matches %>%
  dplyr::left_join(
    protein_and_gene,
    by = c(
      "protein_accession" = "protein_id"
    )
  ) %>%
  dplyr::select(
    gene_id,
    protein_accession,
    gene_symbol,
    percent_identity,
    e_value
  )

head(MEblack_protein_gene_preview, 15)
```

```{r}
# Build the WGCNA gene → protein → LOC gene annotation bridge
wgcna_gene_annotation <- module_gene_table_final %>%
  dplyr::left_join(
    diamond_bridge %>%
      dplyr::select(
        gene_id,
        protein_accession,
        percent_identity,
        alignment_length,
        e_value,
        bit_score
      ),
    by = "gene_id"
  ) %>%
  dplyr::left_join(
    protein_and_gene %>%
      dplyr::distinct(
        protein_id,
        .keep_all = TRUE
      ),
    by = c(
      "protein_accession" = "protein_id"
    )
  )

# Convert LOC-based GO mappings back to the evm.TU WGCNA IDs
wgcna_gene2cat <- wgcna_gene_annotation %>%
  dplyr::filter(
    !is.na(gene_symbol)
  ) %>%
  dplyr::select(
    gene_id,
    gene_symbol
  ) %>%
  dplyr::distinct() %>%
  dplyr::inner_join(
    gene2cat,
    by = c(
      "gene_symbol" = "ensembl_gene_id"
    )
  ) %>%
  dplyr::transmute(
    gene_id,
    go_id
  ) %>%
  dplyr::filter(
    !is.na(go_id),
    go_id != ""
  ) %>%
  dplyr::distinct()

# Preview the final mapping
head(wgcna_gene2cat, 15)
```

```{r}
# Genes with at least one GO annotation
wgcna_genes_with_GO <- unique(
  wgcna_gene2cat$gene_id
)

MEblack_gene_ids <- MEblack_genes$gene_id

GO_coverage_summary <- tibble::tibble(
  gene_set = c(
    "All WGCNA genes",
    "MEblack genes"
  ),

  total_genes = c(
    nrow(module_gene_table_final),
    length(MEblack_gene_ids)
  ),

  protein_matches = c(
    sum(
      !is.na(
        wgcna_gene_annotation$protein_accession
      )
    ),
    sum(
      !is.na(
        wgcna_gene_annotation$protein_accession[
          wgcna_gene_annotation$module == "black"
        ]
      )
    )
  ),

  LOC_matches = c(
    sum(
      !is.na(
        wgcna_gene_annotation$gene_symbol
      )
    ),
    sum(
      !is.na(
        wgcna_gene_annotation$gene_symbol[
          wgcna_gene_annotation$module == "black"
        ]
      )
    )
  ),

  genes_with_GO = c(
    sum(
      module_gene_table_final$gene_id %in%
        wgcna_genes_with_GO
    ),
    sum(
      MEblack_gene_ids %in%
        wgcna_genes_with_GO
    )
  )
) %>%
  dplyr::mutate(
    GO_coverage_percent =
      100 * genes_with_GO / total_genes
  )

GO_coverage_summary

# Additional mapping totals
tibble::tibble(
  total_gene_GO_pairs = nrow(wgcna_gene2cat),
  unique_annotated_WGCNA_genes =
    dplyr::n_distinct(wgcna_gene2cat$gene_id),
  unique_GO_terms =
    dplyr::n_distinct(wgcna_gene2cat$go_id)
)
```

```{r}
MEblack_annotation_status <- wgcna_gene_annotation %>%
  dplyr::filter(
    module == "black"
  ) %>%
  dplyr::mutate(
    has_DIAMOND_match =
      !is.na(protein_accession),

    has_LOC_match =
      !is.na(gene_symbol),

    has_GO_annotation =
      gene_id %in% wgcna_genes_with_GO
  ) %>%
  dplyr::select(
    gene_id,
    protein_accession,
    gene_symbol,
    percent_identity,
    e_value,
    has_DIAMOND_match,
    has_LOC_match,
    has_GO_annotation
  )

MEblack_annotation_status %>%
  dplyr::filter(
    !has_GO_annotation
  )
```

```{r}
wgcna_gene2cat <- wgcna_gene_annotation %>%
  dplyr::filter(!is.na(gene_symbol)) %>%
  dplyr::select(
    gene_id,
    gene_symbol
  ) %>%
  dplyr::distinct() %>%
  dplyr::inner_join(
    gene2cat %>%
      dplyr::distinct(
        ensembl_gene_id,
        go_id
      ),
    by = c(
      "gene_symbol" = "ensembl_gene_id"
    ),
    relationship = "many-to-many"
  ) %>%
  dplyr::transmute(
    gene_id,
    go_id
  ) %>%
  dplyr::filter(
    !is.na(go_id),
    go_id != ""
  ) %>%
  dplyr::distinct()

# Confirm that every gene–GO combination is unique
wgcna_gene2cat %>%
  dplyr::count(
    gene_id,
    go_id
  ) %>%
  dplyr::filter(n > 1)
```

```{r}
# Extract one featureCounts length per gene
wgcna_gene_lengths <- sophie_raw %>%
  dplyr::transmute(
    gene_id = as.character(Geneid),
    gene_length = as.numeric(Length)
  ) %>%
  dplyr::group_by(gene_id) %>%
  dplyr::summarise(
    gene_length = dplyr::first(gene_length),
    .groups = "drop"
  )

# Build the complete 838-gene GOseq input table
MEblack_goseq_input <- module_gene_table_final %>%
  dplyr::transmute(
    gene_id,
    module,
    selected_MEblack = as.integer(module == "black")
  ) %>%
  dplyr::left_join(
    wgcna_gene_lengths,
    by = "gene_id"
  ) %>%
  dplyr::arrange(gene_id)

# Named binary selection vector
MEblack_gene_vector <- MEblack_goseq_input$selected_MEblack
names(MEblack_gene_vector) <- MEblack_goseq_input$gene_id

# Named gene-length vector
MEblack_length_vector <- MEblack_goseq_input$gene_length
names(MEblack_length_vector) <- MEblack_goseq_input$gene_id

# Validation checks
dim(MEblack_goseq_input)

table(MEblack_gene_vector)

sum(is.na(MEblack_length_vector))
sum(MEblack_length_vector <= 0, na.rm = TRUE)

identical(
  names(MEblack_gene_vector),
  names(MEblack_length_vector)
)

summary(MEblack_length_vector)
```

```{r}
# Estimate the probability weighting function for gene-length bias
MEblack_pwf <- goseq::nullp(
  DEgenes = MEblack_gene_vector,
  bias.data = MEblack_length_vector,
  plot.fit = TRUE
)

# Inspect the resulting weighting table
dim(MEblack_pwf)
names(MEblack_pwf)
head(MEblack_pwf, 10)
```

```{r}
# Confirm that the selected-gene indicators remained correct
table(MEblack_pwf$DEgenes)

# Summarize the estimated selection probabilities
summary(MEblack_pwf$pwf)

# Confirm that the weighting function contains no missing values
sum(is.na(MEblack_pwf$pwf))
```

Step 4: Run GoEnrich

```{r}
# Confirm the two required inputs exist
exists("MEblack_pwf")
exists("wgcna_gene2cat")

dim(MEblack_pwf)
dim(wgcna_gene2cat)

# Use a plain two-column data.frame for GOseq
MEblack_gene2cat_input <- wgcna_gene2cat %>%
  dplyr::select(
    gene_id,
    go_id
  ) %>%
  dplyr::distinct() %>%
  as.data.frame()

# Run MEblack GO enrichment
MEblack_goseq_results <- goseq::goseq(
  pwf = MEblack_pwf,
  gene2cat = MEblack_gene2cat_input,
  method = "Wallenius",
  use_genes_without_cat = TRUE
)

# Confirm that the result was created
exists("MEblack_goseq_results")
dim(MEblack_goseq_results)
head(MEblack_goseq_results, 10)
```

```{r}
# Run GO enrichment for the MEblack module
MEblack_goseq_results <- goseq::goseq(
  pwf = MEblack_pwf,
  gene2cat = wgcna_gene2cat,
  method = "Wallenius",
  use_genes_without_cat = TRUE
)

# Add BH false-discovery-rate correction
MEblack_goseq_results <- MEblack_goseq_results %>%
  dplyr::mutate(
    padj = p.adjust(
      over_represented_pvalue,
      method = "BH"
    )
  ) %>%
  dplyr::arrange(
    over_represented_pvalue
  )

# Inspect the most enriched GO categories
head(MEblack_goseq_results, 20)
```

```{r}
tibble::tibble(
  total_GO_terms_tested =
    nrow(MEblack_goseq_results),

  nominally_enriched_p_less_0_05 =
    sum(
      MEblack_goseq_results$over_represented_pvalue < 0.05,
      na.rm = TRUE
    ),

  FDR_significant_padj_less_0_05 =
    sum(
      MEblack_goseq_results$padj < 0.05,
      na.rm = TRUE
    ),

  minimum_raw_p =
    min(
      MEblack_goseq_results$over_represented_pvalue,
      na.rm = TRUE
    ),

  minimum_adjusted_p =
    min(
      MEblack_goseq_results$padj,
      na.rm = TRUE
    )
)
```

Several RNA-related GO categories showed nominal enrichment in the MEblack module, but none remained significant after correction for testing 578 GO terms.

We should **not** call these definitively enriched pathways. They are exploratory biological signals that may help us understand the module.

Because MEblack increases as salinity decreases, one exploratory biological hypothesis is that **lower-salinity events are associated with coordinated changes in genes involved in RNA regulation or gene-silencing processes**.

```{r}
# Keep the nominally enriched MEblack GO terms
MEblack_top_terms <- MEblack_goseq_results %>%
  dplyr::filter(
    over_represented_pvalue < 0.05
  ) %>%
  dplyr::select(
    category,
    term,
    ontology,
    over_represented_pvalue,
    padj,
    numDEInCat,
    numInCat
  ) %>%
  dplyr::arrange(
    over_represented_pvalue
  )

MEblack_top_terms
```

```{r}
# Identify the MEblack genes contributing to each top GO term
MEblack_top_term_genes <- wgcna_gene2cat %>%
  dplyr::filter(
    gene_id %in% MEblack_genes$gene_id
  ) %>%
  dplyr::inner_join(
    MEblack_top_terms,
    by = c(
      "go_id" = "category"
    ),
    relationship = "many-to-many"
  ) %>%
  dplyr::left_join(
    wgcna_gene_annotation %>%
      dplyr::select(
        gene_id,
        protein_accession,
        gene_symbol,
        percent_identity,
        e_value,
        bit_score
      ) %>%
      dplyr::distinct(
        gene_id,
        .keep_all = TRUE
      ),
    by = "gene_id"
  ) %>%
  dplyr::distinct(
    go_id,
    gene_id,
    .keep_all = TRUE
  ) %>%
  dplyr::arrange(
    over_represented_pvalue,
    term,
    gene_id
  )

MEblack_top_term_genes %>%
  dplyr::select(
    go_id,
    term,
    ontology,
    over_represented_pvalue,
    padj,
    gene_id,
    protein_accession,
    gene_symbol,
    percent_identity,
    e_value
  )
```

```{r}
# Summarize the supporting genes for each GO term
MEblack_top_term_gene_summary <- MEblack_top_term_genes %>%
  dplyr::group_by(
    go_id,
    term,
    ontology,
    over_represented_pvalue,
    padj,
    numDEInCat,
    numInCat
  ) %>%
  dplyr::summarise(
    observed_MEblack_genes =
      dplyr::n_distinct(gene_id),

    supporting_gene_ids =
      paste(
        sort(unique(gene_id)),
        collapse = ", "
      ),

    matched_LOC_ids =
      paste(
        sort(unique(stats::na.omit(gene_symbol))),
        collapse = ", "
      ),

    .groups = "drop"
  ) %>%
  dplyr::arrange(
    over_represented_pvalue
  )

MEblack_top_term_gene_summary
```

```{r}
# Genes contributing to multiple nominally enriched GO terms
MEblack_recurrent_GO_genes <- MEblack_top_term_genes %>%
  dplyr::group_by(
    gene_id,
    protein_accession,
    gene_symbol
  ) %>%
  dplyr::summarise(
    number_of_top_GO_terms =
      dplyr::n_distinct(go_id),

    GO_terms =
      paste(
        sort(unique(term)),
        collapse = " | "
      ),

    .groups = "drop"
  ) %>%
  dplyr::arrange(
    dplyr::desc(number_of_top_GO_terms),
    gene_id
  )

MEblack_recurrent_GO_genes
```

This tells us something really useful: the nine nominal GO terms are **not nine independent biological signals**. Most of the RNA-related terms are being driven by the **same two MEblack genes**.

**INDIVIDUAL GENE REGRESSION**

Step 1: Inspect individual gene expression data

```{r}
# Primary individual-gene candidate
candidate_gene <- "evm.TU.Chr09.2202"

# Inspect the normalized expression matrix used for WGCNA
class(datExpr_modules_sophie)
dim(datExpr_modules_sophie)

# Confirm whether genes are stored in columns or rows
candidate_gene %in% colnames(datExpr_modules_sophie)
candidate_gene %in% rownames(datExpr_modules_sophie)

# Preview sample and gene identifiers
head(rownames(datExpr_modules_sophie), 10)
head(colnames(datExpr_modules_sophie), 10)

# Preview the candidate gene's normalized expression values
if (candidate_gene %in% colnames(datExpr_modules_sophie)) {
  candidate_gene_expression <- as.numeric(
    datExpr_modules_sophie[, candidate_gene]
  )

  names(candidate_gene_expression) <-
    rownames(datExpr_modules_sophie)

  candidate_gene_expression
  summary(candidate_gene_expression)
}
```

```{r}
# Inspect the sample-level table that should connect samples to events
class(sample_level_eigengenes_sophie)
dim(sample_level_eigengenes_sophie)
names(sample_level_eigengenes_sophie)

sample_level_eigengenes_sophie %>%
  dplyr::select(
    dplyr::any_of(
      c(
        "sample",
        "sample_id",
        "Sample",
        "event_id",
        "location",
        "month",
        "OysterID"
      )
    )
  ) %>%
  head(18)

```

Step 2: Locate the candidate gene's expression values

```{r}
candidate_gene <- "evm.TU.Chr09.2202"

# Search matrix/data-frame objects for the candidate gene
candidate_gene_locations <- purrr::map_dfr(
  ls(),
  function(object_name) {

    object <- get(
      object_name,
      envir = .GlobalEnv
    )

    if (!(is.matrix(object) || is.data.frame(object))) {
      return(NULL)
    }

    row_names <- rownames(object)
    column_names <- colnames(object)

    exact_row_match <-
      !is.null(row_names) &&
      candidate_gene %in% row_names

    exact_column_match <-
      !is.null(column_names) &&
      candidate_gene %in% column_names

    gene_id_column_match <-
      is.data.frame(object) &&
      "gene_id" %in% names(object) &&
      candidate_gene %in% as.character(object$gene_id)

    partial_row_matches <- if (!is.null(row_names)) {
      paste(
        grep(
          "Chr09\\.2202",
          row_names,
          value = TRUE
        ),
        collapse = ", "
      )
    } else {
      ""
    }

    partial_column_matches <- if (!is.null(column_names)) {
      paste(
        grep(
          "Chr09\\.2202",
          column_names,
          value = TRUE
        ),
        collapse = ", "
      )
    } else {
      ""
    }

    tibble::tibble(
      object = object_name,
      class = paste(class(object), collapse = ", "),
      rows = nrow(object),
      columns = ncol(object),
      exact_row_match,
      exact_column_match,
      gene_id_column_match,
      partial_row_matches,
      partial_column_matches
    )
  }
) %>%
  dplyr::filter(
    exact_row_match |
      exact_column_match |
      gene_id_column_match |
      partial_row_matches != "" |
      partial_column_matches != ""
  )

candidate_gene_locations
```

```{r}
missing_from_datExpr_sophie <- module_gene_table_final %>%
  dplyr::filter(
    !gene_id %in% colnames(datExpr_modules_sophie)
  )

nrow(missing_from_datExpr_sophie)

missing_from_datExpr_sophie %>%
  dplyr::arrange(module, gene_id)
```

```{r}
tibble::tibble(
  gene_id = c(
    "evm.TU.Chr08.2401",
    "evm.TU.Chr09.2202"
  ),
  present_in_datExpr_modules_sophie = gene_id %in%
    colnames(datExpr_modules_sophie)
)
```
