---
title: "RNA Sequencing - Oyesters"
output:
  html_document:
    df_print: paged
---

Load the needed packages

```{r}
suppressPackageStartupMessages({
    library(AnnotationDbi)
    library(org.Mm.eg.db)
    library(hgu95av2.db)
    library(SummarizedExperiment)
})
```

Set Working Directory

```{r}
setwd("/work/clh162/Data+/Arushi/2025-2026-team-code/Data+_team/Arushi")
```

Check Files within the directory

```{r}
list.files("../multiqc_report_raw/multiqc_data")
list.files("../multiqc_report_trimmed/multiqc_data")
```

Open these files

```{r}
browseURL("../multiqc_report_trimmed/multiqc_report.html")
```

Read Data

```{r}
raw_fastqc <- read.delim(
  "../multiqc_report_raw/multiqc_data/multiqc_fastqc.txt",
  check.names = FALSE
)

trimmed_fastqc <- read.delim(
  "../multiqc_report_trimmed/multiqc_data/multiqc_fastqc.txt",
  check.names = FALSE
)

dim(raw_fastqc)
dim(trimmed_fastqc)

head(raw_fastqc)
head(trimmed_fastqc)

```

Count Check Matrix

```{r}
count_dir <- "/work/clh162/OysterRNA24/hisat2_align/featureCounts_counts"

files <- list.files(
  count_dir,
  pattern = "_gene_counts\\.txt$",
  full.names = TRUE
)

length(files)
basename(files)[1:5]

test <- read.delim(files[1], comment.char = "#", check.names = FALSE)

dim(test)
names(test)
head(test)
```

```{r}
count_dir <- "/work/clh162/OysterRNA24/hisat2_align/featureCounts_counts"

files <- list.files(
  count_dir,
  pattern = "_gene_counts\\.txt$",
  full.names = TRUE
)

read_featurecounts <- function(file) {
  df <- read.delim(
    file,
    comment.char = "#",
    check.names = FALSE
  )
  
  sample_name <- basename(file)
  sample_name <- sub("_gene_counts\\.txt$", "", sample_name)
  
  gene_col <- 1
  count_col <- ncol(df)
  
  counts <- df[, c(gene_col, count_col)]
  names(counts) <- c("Geneid", sample_name)
  
  return(counts)
}

count_list <- lapply(files, read_featurecounts)

counts_df <- Reduce(
  function(x, y) merge(x, y, by = "Geneid", all = TRUE),
  count_list
)

rownames(counts_df) <- counts_df$Geneid
counts <- counts_df[, -1]

dim(counts)
head(counts[, 1:5])
```

```{r}
# Check for missing values
sum(is.na(counts))

# Check data type
str(counts[, 1:3])

# Summary of library sizes
summary(colSums(counts))

# How many genes have zero counts across all samples?
sum(rowSums(counts) == 0)

# How many genes have at least 10 counts total?
sum(rowSums(counts) >= 10)
```

Save Count Matrix for later

```{r}
write.csv(
  counts,
  "combined_featureCounts_gene_counts.csv",
  quote = FALSE
)
```

Sample Annotations

```{r}
library(readxl)

coldata <- read_excel("../xRNA for BC Summer 2024.xlsx")

dim(coldata)
head(coldata)
names(coldata)
```

Gene Annotation

```{r}
gtf_file <- "/work/clh162/OysterRNA24/hisat2_align/Cv_genome_RU_2025_shared/Cvi_RU25.gtf"

gtf <- read.delim(
  gtf_file,
  comment.char = "#",
  header = FALSE,
  sep = "\t",
  quote = ""
)

dim(gtf)
head(gtf)
```

```{r}
names(gtf) <- c(
  "seqname", "source", "feature", "start", "end",
  "score", "strand", "frame", "attribute"
)

head(gtf)
table(gtf$feature)
```

```{r}
table(gtf$feature)
head(gtf$attribute, 10)
head(rownames(counts), 10)
```

```{r}
gtf$gene_id <- sub('.*gene_id "([^"]+)".*', '\\1', gtf$attribute)

# If gene_id was not found, this will leave the whole attribute string unchanged.
# So keep only rows where the regex actually worked:
gtf_gene_annot <- gtf[gtf$gene_id != gtf$attribute, ]

dim(gtf_gene_annot)
head(gtf_gene_annot[, c("gene_id", "seqname", "start", "end", "strand", "feature", "attribute")])
```

```{r}
rowranges <- gtf_gene_annot[!duplicated(gtf_gene_annot$gene_id), 
                            c("gene_id", "seqname", "start", "end", "strand", "feature", "attribute")]

rownames(rowranges) <- rowranges$gene_id

sum(rownames(counts) %in% rownames(rowranges))
length(rownames(counts))

rowranges <- rowranges[rownames(counts), ]

dim(rowranges)
head(rowranges)
all(rownames(rowranges) == rownames(counts))
```

```{r}
all(rownames(rowranges) == rownames(counts))

head(rownames(counts))
head(rownames(rowranges))
```

Quick Checks

```{r}
dim(rowranges)
# 28538 7
```

```{r}
nrow(counts)
# 28538
```

```{r}
all(rownames(rowranges) == rownames(counts))
```

Summarized Experiment

```{r}
library("SummarizedExperiment")
```

1.  Make sure count matrix is accurate and not in df form

```{r}
count_matrix <- as.matrix(counts)

storage.mode(count_matrix) <- "integer"

dim(count_matrix)
head(count_matrix[, 1:5])
```

2.  Clean sample annotations

```{r}

# Step 2: Clean sample annotations

sample_names <- colnames(count_matrix)

# Split sample names like:
# CMT_SJ1_19Aug24_Gill_S203_L005
sample_parts <- do.call(rbind, strsplit(sample_names, "_"))

coldata_clean <- data.frame(
  sample_id = sample_names,
  treatment = sample_parts[, 1],
  site = sample_parts[, 2],
  collection_date_raw = sample_parts[, 3],
  tissue = sample_parts[, 4],
  sample_code = sample_parts[, 5],
  lane = sample_parts[, 6],
  stringsAsFactors = FALSE
)

# Convert date if possible
coldata_clean$collection_date <- as.Date(
  coldata_clean$collection_date_raw,
  format = "%d%b%y"
)

# Make rownames match the count matrix columns
rownames(coldata_clean) <- coldata_clean$sample_id

# Check
dim(coldata_clean)
head(coldata_clean)
all(rownames(coldata_clean) == colnames(count_matrix))
```

```{r}
coldata <- coldata_clean

dim(count_matrix)
dim(coldata)

colnames(count_matrix)
rownames(coldata)

all(rownames(coldata) == colnames(count_matrix))
```

3.  Create usable sample metadata from sample names

```{r}

sample_names <- colnames(count_matrix)

coldata_from_counts <- data.frame(
  sample_id = sample_names,
  group = sub("_.*", "", sample_names),
  site = sub("^[^_]+_([^_]+)_.*", "\\1", sample_names),
  date = sub("^[^_]+_[^_]+_([^_]+)_.*", "\\1", sample_names),
  tissue = sub("^[^_]+_[^_]+_[^_]+_([^_]+)_.*", "\\1", sample_names),
  sample_code = sub(".*_(S[0-9]+)_.*", "\\1", sample_names),
  lane = sub(".*_(L[0-9]+)$", "\\1", sample_names),
  stringsAsFactors = FALSE
)

rownames(coldata_from_counts) <- coldata_from_counts$sample_id

coldata_from_counts
all(rownames(coldata_from_counts) == colnames(count_matrix))
```

4.  Prep gene annotations

```{r}

dim(rowranges)
nrow(count_matrix)

all(rownames(rowranges) == rownames(count_matrix))
```

5.  Create the Summarized Experiment

```{r}

library(SummarizedExperiment)

stopifnot(all(rownames(rowranges) == rownames(count_matrix)))
stopifnot(all(rownames(coldata_from_counts) == colnames(count_matrix)))

se <- SummarizedExperiment(
  assays = list(counts = count_matrix),
  rowData = rowranges,
  colData = coldata_from_counts
)

se
```

```{r}
assay(se)[1:5, 1:5]
head(rowData(se))
colData(se)
```

6.  Saving the data

```{r}

saveRDS(se, file = "oyster_summarized_experiment.rds")
```

7.  tidySummarizedExperiment

```{r}

#BiocManager::install("tidySummarizedExperiment")
library(tidySummarizedExperiment)
library(dplyr)
library(ggplot2)
```

```{r}

se_tidy <- se |>
  tidySummarizedExperiment::tidy()

library_sizes <- se_tidy |>
  group_by(sample_id, group, site, date) |>
  summarise(total_counts = sum(counts), .groups = "drop")

ggplot(library_sizes, aes(x = sample_id, y = total_counts, fill = group)) +
  geom_col() +
  theme_bw() +
  coord_flip() +
  labs(
    title = "Library sizes by sample",
    x = "Sample",
    y = "Total assigned gene counts"
  )
```

Finally, Plotting the top expressed genes

```{r}

library(tidySummarizedExperiment)
library(dplyr)
library(ggplot2)

top_genes <- se_tidy |>
  group_by(gene_id) |>
  summarise(mean_count = mean(counts), .groups = "drop") |>
  arrange(desc(mean_count)) |>
  slice_head(n = 10)

group_gene_summary <- se_tidy |>
  filter(gene_id %in% top_genes$gene_id) |>
  group_by(gene_id, group) |>
  summarise(mean_count = mean(counts), .groups = "drop")

ggplot(
  group_gene_summary,
  aes(x = gene_id, y = mean_count + 1, fill = group)
) +
  geom_col(position = "dodge") +
  scale_y_log10() +
  coord_flip() +
  theme_bw() +
  labs(
    title = "Mean counts for top expressed genes by group",
    x = "Gene",
    y = "Mean gene counts + 1, log10 scale",
    fill = "Group"
  )
```
