---
title: "RNA Sequencing - Oyesters"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
---

Load the needed packages

```{r}
suppressPackageStartupMessages({
    library(AnnotationDbi)
    library(org.Mm.eg.db)
    library(hgu95av2.db)
    library(SummarizedExperiment)
})
```

Set Working Directory

```{r}
setwd("/work/clh162/Data+/Arushi/2025-2026-team-code/Data+_team/Arushi")
```

Check Files within the directory

```{r}
list.files("../multiqc_report_raw/multiqc_data")
list.files("../multiqc_report_trimmed/multiqc_data")
```

Open these files

```{r}
browseURL("../multiqc_report_trimmed/multiqc_report.html")
```

Read Data

```{r}
raw_fastqc <- read.delim(
  "../multiqc_report_raw/multiqc_data/multiqc_fastqc.txt",
  check.names = FALSE
)

trimmed_fastqc <- read.delim(
  "../multiqc_report_trimmed/multiqc_data/multiqc_fastqc.txt",
  check.names = FALSE
)

dim(raw_fastqc)
dim(trimmed_fastqc)

head(raw_fastqc)
head(trimmed_fastqc)

```

Count Check Matrix

```{r}
count_file <- "/work/clh162/Data+/Arushi/2025-2026-team-code/Data+_team/Arushi/featureCounts_counts/combined_featureCounts_gene_counts.txt"

fc <- read.delim(
  count_file,
  comment.char = "#",
  check.names = FALSE
)

dim(fc)
names(fc)
head(fc)
```

```{r}
# The first 6 columns are annotation columns from featureCounts
# Count columns start at column 7
counts <- fc[, 7:ncol(fc)]

# Clean sample names from BAM paths
sample_names <- basename(colnames(counts))
sample_names <- sub("_sorted\\.bam$", "", sample_names)

colnames(counts) <- sample_names
rownames(counts) <- fc$Geneid

count_matrix <- as.matrix(counts)
storage.mode(count_matrix) <- "integer"

dim(count_matrix)
head(count_matrix[, 1:5])
summary(colSums(count_matrix))
```

```{r}
# Check for missing values
sum(is.na(counts))

# Check data type
str(counts[, 1:3])

# Summary of library sizes
summary(colSums(counts))

# How many genes have zero counts across all samples?
sum(rowSums(counts) == 0)

# How many genes have at least 10 counts total?
sum(rowSums(counts) >= 10)
```

Save Count Matrix for later

```{r}
write.csv(
  counts,
  "combined_featureCounts_gene_counts.csv",
  quote = FALSE
)
```

Sample Annotations

```{r}
library(readxl)

coldata <- read_excel("../xRNA for BC Summer 2024.xlsx")

dim(coldata)
head(coldata)
names(coldata)
```

Gene Annotation

```{r}
gtf_file <- "/work/clh162/OysterRNA24/hisat2_align/Cv_genome_RU_2025_shared/Cvi_RU25.gtf"

gtf <- read.delim(
  gtf_file,
  comment.char = "#",
  header = FALSE,
  sep = "\t",
  quote = ""
)

dim(gtf)
head(gtf)
```

```{r}
names(gtf) <- c(
  "seqname", "source", "feature", "start", "end",
  "score", "strand", "frame", "attribute"
)

head(gtf)
table(gtf$feature)
```

```{r}
table(gtf$feature)
head(gtf$attribute, 10)
head(rownames(counts), 10)
```

```{r}
gtf$gene_id <- sub('.*gene_id "([^"]+)".*', '\\1', gtf$attribute)

# If gene_id was not found, this will leave the whole attribute string unchanged.
# So keep only rows where the regex actually worked:
gtf_gene_annot <- gtf[gtf$gene_id != gtf$attribute, ]

dim(gtf_gene_annot)
head(gtf_gene_annot[, c("gene_id", "seqname", "start", "end", "strand", "feature", "attribute")])
```

```{r}
rowranges <- gtf_gene_annot[!duplicated(gtf_gene_annot$gene_id), 
                            c("gene_id", "seqname", "start", "end", "strand", "feature", "attribute")]

rownames(rowranges) <- rowranges$gene_id

sum(rownames(counts) %in% rownames(rowranges))
length(rownames(counts))

rowranges <- rowranges[rownames(counts), ]

dim(rowranges)
head(rowranges)
all(rownames(rowranges) == rownames(counts))
```

```{r}
all(rownames(rowranges) == rownames(counts))

head(rownames(counts))
head(rownames(rowranges))
```

Quick Checks

```{r}
dim(rowranges)
# 28538 7
```

```{r}
nrow(counts)
# 28538
```

```{r}
all(rownames(rowranges) == rownames(counts))
```

Summarized Experiment

```{r}
library("SummarizedExperiment")
```

1.  Make sure count matrix is accurate and not in df form

```{r}
count_matrix <- as.matrix(counts)

storage.mode(count_matrix) <- "integer"

dim(count_matrix)
head(count_matrix[, 1:5])
```

2.  Clean sample annotations

```{r}

sample_names <- colnames(count_matrix)

sample_parts <- do.call(rbind, strsplit(sample_names, "_"))

coldata <- data.frame(
  sample_id = sample_names,
  location = sample_parts[, 1],
  oyster_id = sample_parts[, 2],
  timepoint_raw = sample_parts[, 3],
  tissue = sample_parts[, 4],
  sequencing_sample = sample_parts[, 5],
  lane = sample_parts[, 6],
  stringsAsFactors = FALSE
)

coldata$collection_date <- as.Date(
  coldata$timepoint_raw,
  format = "%d%b%y"
)

coldata$month <- factor(
  format(coldata$collection_date, "%b"),
  levels = c("Jun", "Jul", "Aug")
)

coldata$location <- factor(coldata$location)
coldata$oyster_id <- factor(coldata$oyster_id)
coldata$tissue <- factor(coldata$tissue)

rownames(coldata) <- coldata$sample_id

all(rownames(coldata) == colnames(count_matrix))
```

```{r}
dim(count_matrix)
dim(coldata)

colnames(count_matrix)
rownames(coldata)

all(rownames(coldata) == colnames(count_matrix))
```

4.  Prep gene annotations

```{r}

dim(rowranges)
nrow(count_matrix)

all(rownames(rowranges) == rownames(count_matrix))
```

5.  Create the Summarized Experiment

```{r}

library(SummarizedExperiment)

stopifnot(all(rownames(rowranges) == rownames(count_matrix)))
stopifnot(all(rownames(coldata) == colnames(count_matrix)))

se <- SummarizedExperiment(
  assays = list(counts = count_matrix),
  colData = coldata,
  rowData = rowranges
)
se
```

```{r}
assay(se)[1:5, 1:5]
head(rowData(se))
colData(se)
```

6.  Saving the data

```{r}

saveRDS(se, file = "oyster_summarized_experiment.rds")
```

7.  tidySummarizedExperiment

```{r}

#BiocManager::install("tidySummarizedExperiment")
library(tidySummarizedExperiment)
library(dplyr)
library(ggplot2)
```

```{r}

library(tidySummarizedExperiment)
library(dplyr)
library(ggplot2)

se_tidy <- as_tibble(se)

library_sizes <- se_tidy |>
  group_by(sample_id, location, oyster_id, month, collection_date) |>
  summarise(total_counts = sum(counts), .groups = "drop")

ggplot(library_sizes, aes(x = sample_id, y = total_counts, fill = location)) +
  geom_col() +
  theme_bw() +
  coord_flip() +
  labs(
    title = "Library sizes by sample",
    x = "Sample",
    y = "Total assigned gene counts",
    fill = "Location"
  )
```

Finally, Plotting the top expressed genes

```{r}

library(tidySummarizedExperiment)
library(dplyr)
library(ggplot2)

se_tidy <- as_tibble(se)

top_genes <- se_tidy |>
  group_by(gene_id) |>
  summarise(
    mean_count = mean(counts),
    .groups = "drop"
  ) |>
  arrange(desc(mean_count)) |>
  slice_head(n = 5)

top_gene_sample_data <- se_tidy |>
  filter(gene_id %in% top_genes$gene_id)

ggplot(
  top_gene_sample_data,
  aes(x = sample_id, y = counts + 1, color = location)
) +
  geom_point(size = 2, alpha = 0.8) +
  scale_y_log10() +
  facet_wrap(~ gene_id, scales = "free_y") +
  coord_flip() +
  theme_bw() +
  labs(
    title = "Top expressed genes across samples",
    x = "Sample",
    y = "Gene counts + 1, log10 scale",
    color = "Location"
  )
```

```{r}
library(readxl)

metadata_file <- "../xRNA for BC Summer 2024.xlsx"

excel_sheets(metadata_file)
```

```{r}
meta <- read_excel(metadata_file, sheet = 1)

dim(meta)
names(meta)
View(meta)
```

```{r}
se <- readRDS("oyster_summarized_experiment.rds")
se
```

```{r}
assay(se)          # the count matrix / expression data
colData(se)        # sample metadata
rowData(se)        # gene/feature annotations
metadata(se)       # extra metadata, if any
```

DESeq2

1.  Load Packages

```{r}

suppressPackageStartupMessages({
  library(DESeq2)
  library(SummarizedExperiment)
  library(dplyr)
  library(ggplot2)
})
```

```{r}
dim(count_matrix)
dim(coldata)

head(coldata)

all(rownames(coldata) == colnames(count_matrix))
```

2.  Make sure variables are factors

```{r}
coldata$location <- factor(coldata$location)
coldata$month <- factor(coldata$month, levels = c("Jun", "Jul", "Aug"))
coldata$oyster_id <- factor(coldata$oyster_id)
```

```{r}
coldata$location <- relevel(coldata$location, ref = "CMT")
```

3.  Create the DESeq2 object and filter low count genes

```{r}
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = coldata,
  design = ~ month + location
)

keep <- rowSums(counts(dds) >= 10) >= 3
dds <- dds[keep, ]

dds <- DESeq(dds)

plotDispEsts(dds, main = "Dispersion estimates using Arushi featureCounts")
```

5.  Run DESeq2

```{r}
dds <- DESeq(dds)
```

```{r}
resultsNames(dds)
```

6.  Extract location results

```{r}

res_location <- results(dds, name = "location_DAF_vs_CMT")

summary(res_location)
```

```{r}
res_location_ordered <- res_location[order(res_location$padj), ]

head(res_location_ordered)
```

```{r}
res_location_df <- as.data.frame(res_location_ordered)

res_location_df$gene_id <- rownames(res_location_df)

head(res_location_df)
```

7.  Add gene annotations

```{r}
rowranges_df <- as.data.frame(rowranges)
rowranges_df$gene_id <- rownames(rowranges_df)

res_annotated <- left_join(
  res_location_df,
  rowranges_df,
  by = "gene_id"
)

head(res_annotated)
```

```{r}
write.csv(
  res_annotated,
  "DESeq2_location_DAF_vs_CMT_results.csv",
  row.names = FALSE,
  quote = FALSE
)
```

8.  Count significant genes

```{r}
sig_location_lfc <- res_annotated |>
  filter(!is.na(padj), padj < 0.05, abs(log2FoldChange) >= 1)

nrow(sig_location_lfc)
```

padj \< 0.05 = statistically significant after multiple testing correction abs(log2FoldChange) \>= 1 = at least 2-fold difference between locations

9.  MA Plot

```{r}
plotMA(res_location, ylim = c(-5, 5))
```

x-axis = average expression\
y-axis = log2 fold change between DAF and CMT

Most genes cluster around `log2 fold change = 0`, which means most genes are not strongly different between DAF and CMT. Genes with higher average expression tend to have more stable fold changes.

10. Volcano plot

```{r}
volcano_df <- res_annotated |>
  mutate(
    significant = ifelse(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) >= 1,
                         "Significant", "Not significant")
  )

ggplot(volcano_df, aes(x = log2FoldChange, y = -log10(padj), color = significant)) +
  geom_point(alpha = 0.6, size = 1.5) +
  theme_bw() +
  labs(
    title = "DESeq2 results: DAF vs CMT",
    x = "log2 fold change",
    y = "-log10 adjusted p-value",
    color = "Result"
  )


```

x-axis = log2 fold change

y-axis = -log10 adjusted p-value

Genes higher up are more statistically significant. Genes farther left/right have stronger fold changes.

11. Variance stabilizing transformation for PCA

```{r}
vsd <- vst(dds, blind = FALSE)

vsd
```

```{r}
vst_matrix <- assay(vsd)

dim(vst_matrix)
```

12. PCA using DESeq2 built in plot

```{r}
plotPCA(vsd, intgroup = c("location", "month"))

```

13. The biggest visible pattern is that **many DAF samples cluster more toward the right side**, while **several CMT samples are more spread out and include left-side outliers**.

    That suggests location may be contributing to gene expression differences. This supports the idea that environmental/location-related factors could affect oyster gene expression.

    PC1 explains \~19% of variance

    PC2 explains \~10% of variance

14. Custom PCA plot

There is not a perfectly clean separation by month. Some June, July, and August samples overlap. That means month/timepoint matters somewhat, but from this view, it does not look like month alone explains everything.

```{r}
pca_data <- plotPCA(
  vsd,
  intgroup = c("location", "month", "oyster_id"),
  returnData = TRUE
)

percent_var <- round(100 * attr(pca_data, "percentVar"))

ggplot(pca_data, aes(x = PC1, y = PC2, color = location, shape = month)) +
  geom_point(size = 4, alpha = 0.9) +
  theme_bw() +
  labs(
    title = "PCA of variance-stabilized gene expression",
    x = paste0("PC1: ", percent_var[1], "% variance"),
    y = paste0("PC2: ", percent_var[2], "% variance"),
    color = "Location",
    shape = "Month"
  )
```

14. PCA with oyster ID labels

```{r}
ggplot(pca_data, aes(x = PC1, y = PC2, color = location, shape = month, label = oyster_id)) +
  geom_point(size = 4, alpha = 0.9) +
  geom_text(vjust = -1, size = 3) +
  theme_bw() +
  labs(
    title = "PCA of gene expression with oyster replicate labels",
    x = paste0("PC1: ", percent_var[1], "% variance"),
    y = paste0("PC2: ", percent_var[2], "% variance"),
    color = "Location",
    shape = "Month"
  )
```

15. PCA with top variable genes manually

```{r}

# Step 16: PCA on top variable genes manually

# Calculate variance for each gene
gene_variances <- apply(vst_matrix, 1, var)

# Pick top 500 most variable genes
top_variable_genes <- names(sort(gene_variances, decreasing = TRUE))[1:500]

# Run PCA on samples
pca <- prcomp(t(vst_matrix[top_variable_genes, ]), scale. = FALSE)

# Make PCA dataframe
pca_df <- as.data.frame(pca$x)
pca_df$sample_id <- rownames(pca_df)

# Add metadata WITHOUT duplicating sample_id
metadata_for_pca <- as.data.frame(coldata)
metadata_for_pca$sample_id <- rownames(metadata_for_pca)

pca_df <- merge(
  pca_df,
  metadata_for_pca,
  by = "sample_id"
)

# Check for duplicate column names
anyDuplicated(names(pca_df))
names(pca_df)

# Percent variance explained
percent_var_manual <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)))

# Plot
ggplot(pca_df, aes(x = PC1, y = PC2, color = location, shape = month)) +
  geom_point(size = 4, alpha = 0.9) +
  theme_bw() +
  labs(
    title = "PCA using top 500 variable genes",
    x = paste0("PC1: ", percent_var_manual[1], "% variance"),
    y = paste0("PC2: ", percent_var_manual[2], "% variance"),
    color = "Location",
    shape = "Month"
  )
```

16. Save plots

```{r}

saveRDS(dds, "oyster_deseq2_dds.rds")
saveRDS(vsd, "oyster_deseq2_vst.rds")
saveRDS(res_annotated, "DESeq2_location_DAF_vs_CMT_results.rds")
```

Heatmap

```{r}
if (!requireNamespace("pheatmap", quietly = TRUE)) {
  install.packages("pheatmap")
}

library(pheatmap)
library(dplyr)

```

```{r}
class(dds)
class(vsd)
head(res_annotated)

vst_matrix <- assay(vsd)

dim(vst_matrix)
```

1.  Choose deferentially expresses genes for the heatmap

```{r}

sig_genes <- res_annotated |>
  filter(
    !is.na(padj),
    padj < 0.05,
    abs(log2FoldChange) >= 1
  ) |>
  arrange(padj)

nrow(sig_genes)
head(sig_genes)

sig_genes_strict <- res_annotated |>
  filter(
    !is.na(padj),
    padj < 0.01,
    abs(log2FoldChange) >= 2
  ) |>
  arrange(padj)

nrow(sig_genes_strict)

top_heatmap_genes <- sig_genes_strict |>
  slice_head(n = 50) |>
  pull(gene_id)

heatmap_matrix <- vst_matrix[top_heatmap_genes, ]
```

```{r}
top_heatmap_genes <- sig_genes |>
  slice_head(n = 50) |>
  pull(gene_id)

length(top_heatmap_genes)
head(top_heatmap_genes)
```

2.  Pull expression values for those genes

```{r}
heatmap_matrix <- vst_matrix[top_heatmap_genes, ]

dim(heatmap_matrix)
heatmap_matrix[1:5, 1:5]
```

3.  Add sample annotations to the heatmap

```{r}
annotation_col <- as.data.frame(coldata[, c("location", "month", "oyster_id")])

rownames(annotation_col) <- rownames(coldata)

annotation_col
```

```{r}
#Quick check
all(rownames(annotation_col) == colnames(heatmap_matrix))
```

4.  Make heatmap

```{r}
pheatmap(
  heatmap_matrix,
  scale = "row",
  annotation_col = annotation_col,
  show_rownames = TRUE,
  show_colnames = TRUE,
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  fontsize_row = 6,
  fontsize_col = 7,
  main = "Top differentially expressed genes: DAF vs CMT"
)
```

```{r}
# Make sure annotation rows match the heatmap columns
annotation_col <- annotation_col[colnames(heatmap_matrix), ]

# Set factor order so pheatmap sorts the way we want
annotation_col$location <- factor(annotation_col$location, levels = c("CMT", "DAF"))
annotation_col$month <- factor(annotation_col$month, levels = c("Jun", "Jul", "Aug"))
annotation_col$oyster_id <- factor(annotation_col$oyster_id, levels = c("SJ1", "SJ2", "SJ3"))

# Order samples by location, then month, then oyster
sample_order <- rownames(annotation_col)[order(
  annotation_col$month,
  annotation_col$location,
  annotation_col$oyster_id
)]

# Reorder matrix and annotation
heatmap_matrix_ordered <- heatmap_matrix[, sample_order]
annotation_col_ordered <- annotation_col[sample_order, ]

# Plot cleaner heatmap
pheatmap(
  heatmap_matrix_ordered,
  scale = "row",
  annotation_col = annotation_col_ordered,
  show_rownames = FALSE,
  show_colnames = TRUE,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  clustering_distance_rows = "euclidean",
  clustering_method = "complete",
  main = "Top differentially expressed genes: DAF vs CMT"
)
```

red/orange = higher-than-average expression for that gene

pale/white = around average

blue/teal = lower-than-average expression

The heatmap shows variance-stabilized expression values for the top differentially expressed genes between DAF and CMT, scaled by gene. Each row represents a gene and each column represents a sample. The column annotations show location, month, and oyster replicate. While some genes show higher relative expression in subsets of DAF or CMT samples, the samples do not cluster perfectly by location. Instead, the clustering suggests that location-associated expression differences exist, but month and biological replicate also contribute to variation in gene expression. This supports the idea that environmental context may influence gene expression, while also highlighting the importance of accounting for timepoint and biological replicate effects in downstream analysis.

Save

```{r}
pheatmap(
  heatmap_matrix,
  scale = "row",
  annotation_col = annotation_col,
  show_rownames = FALSE,
  show_colnames = TRUE,
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  main = "Top differentially expressed genes: DAF vs CMT",
  filename = "DESeq2_DAF_vs_CMT_heatmap.png",
  width = 10,
  height = 8
)

list.files(pattern = "heatmap")
```

Check

```{r}
# Metadata alignment
all(rownames(coldata) == colnames(count_matrix))
coldata

# Experimental design balance
table(coldata$location)
table(coldata$month)
table(coldata$oyster_id)
table(coldata$location, coldata$month)
table(coldata$location, coldata$oyster_id)

# Count matrix checks
dim(count_matrix)
anyDuplicated(rownames(count_matrix))
anyDuplicated(colnames(count_matrix))
all(count_matrix == round(count_matrix))
summary(colSums(count_matrix))

# Low-count gene check
summary(rowSums(count_matrix))
sum(rowSums(count_matrix) >= 10)
sum(rowSums(count_matrix >= 10) >= 3)
```

```{r}
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = coldata,
  design = ~ month + location
)

keep <- rowSums(counts(dds) >= 10) >= 3
dds <- dds[keep, ]

dds <- DESeq(dds)

plotDispEsts(dds, main = "Dispersion estimates after low-count filtering")
```

```{r}
rm(list = ls())
```

FIX

```{r}
library(DESeq2)
library(SummarizedExperiment)
library(dplyr)
```

```{r}
count_file <- "/work/clh162/Data+/Arushi/2025-2026-team-code/Data+_team/Arushi/featureCounts_counts/combined_featureCounts_gene_counts.txt"

file.exists(count_file)

fc <- read.delim(
  count_file,
  comment.char = "#",
  check.names = FALSE
)

dim(fc)
names(fc)
head(fc)
```

```{r}
counts <- fc[, 7:ncol(fc)]

sample_names <- basename(colnames(counts))
sample_names <- sub("_sorted\\.bam$", "", sample_names)

colnames(counts) <- sample_names
rownames(counts) <- fc$Geneid

count_matrix <- as.matrix(counts)
storage.mode(count_matrix) <- "integer"

dim(count_matrix)
colnames(count_matrix)
summary(colSums(count_matrix))
all(count_matrix == round(count_matrix))
```

```{r}
sample_names <- colnames(count_matrix)

sample_parts <- do.call(rbind, strsplit(sample_names, "_"))

coldata <- data.frame(
  sample_id = sample_names,
  location = sample_parts[, 1],
  oyster_id = sample_parts[, 2],
  timepoint_raw = sample_parts[, 3],
  tissue = sample_parts[, 4],
  sequencing_sample = sample_parts[, 5],
  lane = sample_parts[, 6],
  stringsAsFactors = FALSE
)

coldata$collection_date <- as.Date(
  coldata$timepoint_raw,
  format = "%d%b%y"
)

coldata$month <- factor(
  format(coldata$collection_date, "%b"),
  levels = c("Jun", "Jul", "Aug")
)

coldata$location <- factor(coldata$location)
coldata$location <- relevel(coldata$location, ref = "CMT")

coldata$oyster_id <- factor(coldata$oyster_id)
coldata$tissue <- factor(coldata$tissue)

rownames(coldata) <- coldata$sample_id

all(rownames(coldata) == colnames(count_matrix))
table(coldata$location, coldata$month)
table(coldata$location, coldata$oyster_id)
```

```{r}
summary_file <- "/work/clh162/Data+/Arushi/2025-2026-team-code/Data+_team/Arushi/featureCounts_counts/combined_featureCounts_gene_counts.txt.summary"

fc_summary <- read.delim(summary_file, check.names = FALSE)

fc_summary
```

```{r}
assigned <- fc_summary[fc_summary$Status == "Assigned", -1]
total <- colSums(fc_summary[, -1])

assigned_percent <- as.numeric(assigned) / total * 100

assigned_check <- data.frame(
  sample = names(assigned),
  assigned_reads = as.numeric(assigned),
  total_reads = as.numeric(total),
  assigned_percent = assigned_percent
)

assigned_check
summary(assigned_check$assigned_percent)
```

```{r}
dds_location <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = coldata,
  design = ~ location
)

dds_location <- dds_location[rowSums(counts(dds_location) >= 10) >= 3, ]
dds_location <- DESeq(dds_location)

plotDispEsts(dds_location, main = "Design: ~ location")
```

```{r}
#This is the model we will be going with

dds_interaction <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = coldata,
  design = ~ month * location
)

keep <- rowSums(counts(dds_interaction) >= 10) >= 3
dds_interaction <- dds_interaction[keep, ]

dds_interaction <- DESeq(dds_interaction)

plotDispEsts(
  dds_interaction,
  main = "Design: ~ month * location"
)

resultsNames(dds_interaction)
```

```{r}
dds_month <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = coldata,
  design = ~ month
)

dds_month <- dds_month[rowSums(counts(dds_month) >= 10) >= 3, ]
dds_month <- DESeq(dds_month)

plotDispEsts(dds_month, main = "Design: ~ month")
```

Pair Wise Comparisons

Main model design: design = \~ month \* location

1.  Run interaction model

```{r}
library(DESeq2)
library(dplyr)

dds_interaction <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = coldata,
  design = ~ month * location
)

keep <- rowSums(counts(dds_interaction) >= 10) >= 3
dds_interaction <- dds_interaction[keep, ]

dds_interaction <- DESeq(dds_interaction)

plotDispEsts(
  dds_interaction,
  main = "Design: ~ month * location"
)

resultsNames(dds_interaction)
```

2.  DAF vs CMT in June

```{r}
res_DAF_vs_CMT_Jun <- results(
  dds_interaction,
  name = "location_DAF_vs_CMT"
)

summary(res_DAF_vs_CMT_Jun)
```

3.  DAF vs CMT in July

```{r}
res_DAF_vs_CMT_Jul <- results(
  dds_interaction,
  contrast = list(
    c("location_DAF_vs_CMT", "monthJul.locationDAF")
  )
)

summary(res_DAF_vs_CMT_Jul)
```

4.  DAF vs CMT in August

```{r}
res_DAF_vs_CMT_Aug <- results(
  dds_interaction,
  contrast = list(
    c("location_DAF_vs_CMT", "monthAug.locationDAF")
  )
)

summary(res_DAF_vs_CMT_Aug)
```

5.  Convert all to a dataframe

```{r}

res_Jun_df <- as.data.frame(res_DAF_vs_CMT_Jun) |>
  tibble::rownames_to_column("gene_id") |>
  arrange(padj)

res_Jul_df <- as.data.frame(res_DAF_vs_CMT_Jul) |>
  tibble::rownames_to_column("gene_id") |>
  arrange(padj)

res_Aug_df <- as.data.frame(res_DAF_vs_CMT_Aug) |>
  tibble::rownames_to_column("gene_id") |>
  arrange(padj)
```

7.  Count significant genes for each month

```{r}
sig_summary <- data.frame(
  comparison = c("DAF vs CMT in Jun", "DAF vs CMT in Jul", "DAF vs CMT in Aug"),
  padj_0.05 = c(
    sum(!is.na(res_Jun_df$padj) & res_Jun_df$padj < 0.05),
    sum(!is.na(res_Jul_df$padj) & res_Jul_df$padj < 0.05),
    sum(!is.na(res_Aug_df$padj) & res_Aug_df$padj < 0.05)
  ),
  padj_0.05_lfc_1 = c(
    sum(!is.na(res_Jun_df$padj) & res_Jun_df$padj < 0.05 & abs(res_Jun_df$log2FoldChange) >= 1),
    sum(!is.na(res_Jul_df$padj) & res_Jul_df$padj < 0.05 & abs(res_Jul_df$log2FoldChange) >= 1),
    sum(!is.na(res_Aug_df$padj) & res_Aug_df$padj < 0.05 & abs(res_Aug_df$log2FoldChange) >= 1)
  )
)

sig_summary
```

8.  Save results

```{r}
write.csv(res_Jun_df, "DESeq2_DAF_vs_CMT_Jun_results.csv", row.names = FALSE)
write.csv(res_Jul_df, "DESeq2_DAF_vs_CMT_Jul_results.csv", row.names = FALSE)
write.csv(res_Aug_df, "DESeq2_DAF_vs_CMT_Aug_results.csv", row.names = FALSE)

write.csv(sig_summary, "DESeq2_pairwise_monthly_summary.csv", row.names = FALSE)
```

9.  Comparison Plot

```{r}
library(ggplot2)

ggplot(sig_summary, aes(x = comparison, y = padj_0.05_lfc_1)) +
  geom_col() +
  theme_bw() +
  coord_flip() +
  labs(
    title = "Significant DAF vs CMT genes by month",
    x = "Comparison",
    y = "Number of significant genes\npadj < 0.05 and |log2FC| >= 1"
  )
```

Outlier Check

```{r}
assays(dds_interaction)[["cooks"]][1:5, 1:5]

cooks <- assays(dds_interaction)[["cooks"]]

sample_cooks_summary <- data.frame(
  sample = colnames(cooks),
  median_cooks = apply(cooks, 2, median, na.rm = TRUE),
  mean_cooks = apply(cooks, 2, mean, na.rm = TRUE),
  max_cooks = apply(cooks, 2, max, na.rm = TRUE)
)

sample_cooks_summary[order(sample_cooks_summary$mean_cooks, decreasing = TRUE), ]
```

```{r}
library(ggplot2)

ggplot(sample_cooks_summary, aes(x = reorder(sample, mean_cooks), y = mean_cooks)) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  labs(
    title = "Average Cook's distance by sample",
    x = "Sample",
    y = "Mean Cook's distance"
  )
```

```{r}
res_DAF_vs_CMT_Jun_noCook <- results(
  dds_interaction,
  name = "location_DAF_vs_CMT",
  cooksCutoff = FALSE
)

res_DAF_vs_CMT_Jul_noCook <- results(
  dds_interaction,
  contrast = list(c("location_DAF_vs_CMT", "monthJul.locationDAF")),
  cooksCutoff = FALSE
)

res_DAF_vs_CMT_Aug_noCook <- results(
  dds_interaction,
  contrast = list(c("location_DAF_vs_CMT", "monthAug.locationDAF")),
  cooksCutoff = FALSE
)
```

```{r}
data.frame(
  comparison = c("Jun", "Jul", "Aug"),
  sig_with_cooks_filter = c(
    sum(!is.na(res_DAF_vs_CMT_Jun$padj) & res_DAF_vs_CMT_Jun$padj < 0.05),
    sum(!is.na(res_DAF_vs_CMT_Jul$padj) & res_DAF_vs_CMT_Jul$padj < 0.05),
    sum(!is.na(res_DAF_vs_CMT_Aug$padj) & res_DAF_vs_CMT_Aug$padj < 0.05)
  ),
  sig_without_cooks_filter = c(
    sum(!is.na(res_DAF_vs_CMT_Jun_noCook$padj) & res_DAF_vs_CMT_Jun_noCook$padj < 0.05),
    sum(!is.na(res_DAF_vs_CMT_Jul_noCook$padj) & res_DAF_vs_CMT_Jul_noCook$padj < 0.05),
    sum(!is.na(res_DAF_vs_CMT_Aug_noCook$padj) & res_DAF_vs_CMT_Aug_noCook$padj < 0.05)
  )
)
```

For each sample, how many genes have a very high Cooks Distance?\

```{r}
cooks <- assays(dds_interaction)[["cooks"]]

cooks_by_sample <- data.frame(
  sample = colnames(cooks),
  mean_cooks = apply(cooks, 2, mean, na.rm = TRUE),
  median_cooks = apply(cooks, 2, median, na.rm = TRUE),
  max_cooks = apply(cooks, 2, max, na.rm = TRUE),
  genes_cooks_gt_1 = colSums(cooks > 1, na.rm = TRUE),
  genes_cooks_gt_5 = colSums(cooks > 5, na.rm = TRUE),
  genes_cooks_gt_10 = colSums(cooks > 10, na.rm = TRUE)
)

cooks_by_sample[order(cooks_by_sample$genes_cooks_gt_1, decreasing = TRUE), ]
```

```{r}
library(ggplot2)

ggplot(cooks_by_sample, aes(x = reorder(sample, genes_cooks_gt_1), y = genes_cooks_gt_1)) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  labs(
    title = "Number of genes with Cook's distance > 1 by sample",
    x = "Sample",
    y = "Genes with Cook's distance > 1"
  )
```
