# Importing Count Data

```{r}
# import libraries
suppressPackageStartupMessages({
  library(AnnotationDbi)
  library(org.Mm.eg.db)
  library(hgu95av2.db)
  library(SummarizedExperiment)
  library(GenomicRanges)
  library(DESeq2)
  library(ggplot2)
  library(ComplexHeatmap)
  library(lubridate)
  library(readxl)
  library(dplyr)
})
```

```{r}
# reads csv data into a variable called counts and skips the first header line
counts <- read.csv("/work/clh162/Data+/Sophie/2025-2026-team-code-sac165/Data+_team/Sophie/gene_counts/gene_counts.csv", skip = 1, check.names = FALSE, row.names = 1)

# remove the full path, .bam extension, and job id/array task id in the column names to just extract the sample names
sample_names <- basename(colnames(counts)[6:ncol(counts)])
sample_names <- sub("_sorted\\.bam", "", sample_names)
sample_names <- sub("^(.*_Gill).*", "\\1", sample_names)
colnames(counts)[6:ncol(counts)] <- sample_names

# set the column names to just the sample names
count_matrix <- counts[, 6:ncol(counts)]
colnames(count_matrix)

# gives number of rows and number of columns
dim(count_matrix)

# save count_matrix
write.table(
  count_matrix,
  file = "/work/clh162/Data+/Sophie/2025-2026-team-code-sac165/Data+_team/Sophie/gene_counts_clean.txt",
  sep = "\t",
  quote = FALSE,
  row.names = TRUE,
  col.names = NA
)
```

```{r}
# read in sample annotations and match the column names of the count matrix to the row names of the column data (make sure the sample names are the same and in the same positions)
col_data <- read.csv("/work/clh162/Data+/Sophie/2025-2026-team-code-sac165/Data+_team/Sophie/xRNA_clean.csv", row.names = 2) 
col_data <- col_data[match(colnames(count_matrix), rownames(col_data)), ]

# read in gene annotations
gene_annotation_gtf <- read.delim("/work/clh162/OysterRNA24/hisat2_align/Cv_genome_RU_2025_shared/Cvi_RU25.gtf", comment.char = "#", header = FALSE, sep = "\t", quote = "")

# name the columns and get gene_id from attribute column
names(gene_annotation_gtf) <- c("seqname", "source", "feature", "start", "end", "score", "strand", "frame", "attribute")
table(gene_annotation_gtf$feature)
transcripts <- gene_annotation_gtf[gene_annotation_gtf$feature == "transcript",]
transcripts$gene_id <- sub('.*gene_id "([^"]+)".*', '\\1', transcripts$attribute)
head(transcripts$gene_id)

# create gene annotation object
row_ranges <- transcripts[, c("gene_id", "seqname", "start", "end", "strand", "feature", "attribute")]
rownames(row_ranges) <- row_ranges$gene_id

# match the row names of the count matrix to the row names of the row ranges
row_ranges <- row_ranges[match(rownames(count_matrix), rownames(row_ranges)), ]
```

```{r}
# make sure samples are in the same order
all.equal(colnames(count_matrix), rownames(col_data))

# make sure genes are in the same order
all.equal(rownames(count_matrix), rownames(row_ranges))
```

```{r}
# convert row_ranges data frame to GRanges object
row_ranges <- GRanges(seqnames = row_ranges$seqname, 
                      ranges = IRanges(
                        start = row_ranges$start,
                        end = row_ranges$end),
                      strand = row_ranges$strand,
                      gene_id = row_ranges$gene_id)
```

# Create SummarizedExperiment Object

```{r}
# create SummarizedExperiment object
se <- SummarizedExperiment(assays = list(count_matrix = as.matrix(count_matrix)), 
                            rowRanges = row_ranges,
                            colData = col_data)
```

# Create DESeqDataSet Object

```{r}
# convert SummarizedExperiment object to DESeqDataSet object
dds = DESeqDataSet(se, design = ~ Location)

# remove rows with low gene counts (keep rows that have at least 10 reads total)
keep <- rowSums(counts(dds) >= 10) >=3
dds <- dds[keep,]

# set the factor level, let DAF (Duke Aquafarm) be reference
dds$Location = relevel(dds$Location, ref = "DAF")
```

# Run DESeq2

```{r}
dds <- DESeq(dds)
results <- results(dds)

# summary results
summary(results)

results0.01 <- results(dds, alpha = 0.01) # with smaller p-value
summary(results0.01)

# contrasts (find info about comparisons made)
resultsNames(dds)
```

# MA Plot

```{r}
plotMA(results)
```

# PCA Plot

```{r}
# transform dds into varianced-stabilized dataset (vsd)
vsd <- vst(dds, blind = FALSE)

# plot
pcaData <- plotPCA(vsd, intgroup = "Location", returnData = TRUE)
ggplot(pcaData, 
       aes(x = PC1, y = PC2, color = Location, label = name)) + geom_point(size = 3) + geom_text(vjust = -1, size = 2)
```

# Gene Expression Heatmap

```{r}
results_ordered <- results[order(results$padj), ]
top_genes <- rownames(results_ordered)[1:50] # order results and get the top genes based on adjusted p-value (top 50 lowest p-values = top 50 most statistically significant genes)
mat <- assay(vsd)[top_genes, ] # get expression values
mat <- t(scale(t(mat))) # scale each gene
heatmap_annotation <- HeatmapAnnotation(Location = colData(vsd)$Location)
Heatmap(mat,
        name = "Expression",
        top_annotation = heatmap_annotation,
        show_row_names = FALSE, # can change if want to see gene ids
        cluster_rows = TRUE,
        cluster_columns = TRUE)
```

# PCA plot also taking month into account
```{r}
# changes characters into dates
col_data$Date_Collected <- dmy(col_data$Date_Collected)

# creates month variable and makes it a column in col_data
col_data$Month <- factor(
  as.character(month(col_data$Date_Collected,
                     label = TRUE,
                     abbr = FALSE)),
  levels = c("June", "July", "August"),
  ordered = FALSE
)

colData(se)$Month <- col_data$Month

colData(vsd)$Month <- col_data$Month

# create PCA plot with month and location
pcaData_month <- plotPCA(
    vsd,
    intgroup = c("Location", "Month"),
    returnData = TRUE
)

ggplot(pcaData_month, 
       aes(x = PC1, y = PC2, color = Month, shape = Location)) + geom_point(size = 3)
```
# Gene Expression Heatmap (each month per location)
```{r}
sample_order <- order(colData(vsd)$Location, colData(vsd)$Month)

heatmap_annotation_month <- HeatmapAnnotation(
  Month = colData(vsd)$Month[sample_order],
  Location = colData(vsd)$Location[sample_order]
)

Heatmap(
  mat[, sample_order],
  name = "Expression",
  top_annotation = heatmap_annotation_month,
  show_row_names = FALSE,
  cluster_rows = TRUE,
  cluster_columns = FALSE
)
```
# Gene Expression Heatmap (each location per month aka CMT vs. DAF for june, july, august)
```{r}
sample_order <- order(colData(vsd)$Month, colData(vsd)$Location)

heatmap_annotation_month <- HeatmapAnnotation(
  Month = colData(vsd)$Month[sample_order],
  Location = colData(vsd)$Location[sample_order]
)

Heatmap(
  mat[, sample_order],
  name = "Expression",
  top_annotation = heatmap_annotation_month,
  show_row_names = FALSE,
  cluster_rows = TRUE,
  cluster_columns = FALSE
)
```

# Create new DESeqDataSet object with month
```{r}
dds_time <- DESeqDataSet(
    se,
    design = ~ Location * Month
)

# create new object with likelihood ratio test (LRT) between simpler model (just location) and more complex model (location and month)
#  dds_time <- DESeq(
#     dds_time,
#     test = "LRT",
#     reduced = ~ Location
# )

keep <- rowSums(counts(dds_time) >= 10) >=3
dds_time <- dds_time[keep,]

results_time <- results(dds_time)

top_month_genes <- rownames(results_time)[order(results_time$padj)][1:12]

vsd_time <- vst(dds_time, blind = FALSE)

# fix months
colData(vsd_time)$CollectionDate <- env_data$CollectionDate
colData(vsd_time)$Month <- factor(
  format(colData(vsd_time)$CollectionDate, "%B"),
  levels = c("June", "July", "August")
)

plot_df <- data.frame(
  Gene = rep(top_month_genes, times = ncol(vsd_time)),
  Expression = as.vector(assay(vsd_time)[top_month_genes, ]),
  Month = rep(colData(vsd_time)$Month, each = length(top_month_genes)),
  Location = rep(colData(vsd_time)$Location, each = length(top_month_genes)),
  Sample = rep(colnames(vsd_time), each = length(top_month_genes))
)

# faceted line plot for top twelve genes ordered by month/time
ggplot(plot_df,
       aes(x = Month,
           y = Expression,
           color = Location,
           group = Location)) +
  geom_point(alpha = 0.7, size = 2) +
  stat_summary(fun = mean,
               geom = "line",
               linewidth = 1) +
  stat_summary(fun = mean,
               geom = "point",
               size = 3) +
  facet_wrap(~ Gene, scales = "free_y") +
  theme_classic() +
  labs(y = "VST-normalized expression",
       x = "Month")
```
# Run DESeq2 Subset by Month
```{r}
# june
dds_june <- dds_time[, colData(dds_time)$Month == "June"]
design(dds_june) <- ~ Location
dds_june <- DESeq(dds_june)
results_june <- results(dds_june)

# july
dds_july <- dds_time[, colData(dds_time)$Month == "July"]
design(dds_july) <- ~ Location
dds_july <- DESeq(dds_july)
results_july <- results(dds_july)

# august
dds_august <- dds_time[, colData(dds_time)$Month == "August"]
design(dds_august) <- ~ Location
dds_august <- DESeq(dds_august)
results_august <- results(dds_august)
```

# Pairwise Comparison MA Plots (DAF vs. CMT for each month)
```{r}
# june
plotMA(results_june, main = "CMT vs. DAF - June", ylim = c(-5, 5))

# july
plotMA(results_july, main = "CMT vs. DAF - July", ylim = c(-5, 5))

# august
plotMA(results_august, main = "CMT vs. DAF - August", ylim = c(-5, 5))

# count significant genes per month (shows how much divergence exists between locations each month)
sum(results_june$padj < 0.05, na.rm = TRUE)
sum(results_july$padj < 0.05, na.rm = TRUE)
sum(results_august$padj < 0.05, na.rm = TRUE)

# determine if the different significant genes are consistent each month
# get significant genes
sig_june <- rownames(results_june)[which(results_june$padj < 0.05)]
sig_july <- rownames(results_july)[which(results_july$padj < 0.05)]
sig_august <- rownames(results_august)[which(results_august$padj < 0.05)]
# count genes that are consistently significant across all three months
length(Reduce(intersect,
              list(sig_june, sig_july, sig_august)))

Reduce(intersect,
              list(sig_june, sig_july, sig_august))

length(intersect(sig_june, sig_july))
length(intersect(sig_june, sig_august))
length(intersect(sig_july, sig_august))

```

# Pairwise Comparison Volcano Plots (DAF vs. CMT for each month)
```{r}
library(tibble)

# volcano plot function
plot_volcano <- function(res, title) {

  df <- as.data.frame(res) %>%
    rownames_to_column("gene_id")

  df$Significant <- with(
    df,
    padj < 0.05 & abs(log2FoldChange) > 1
  )
  
  label_genes <- df %>%
  filter(
    padj < 1e-7,
    abs(log2FoldChange) > 2
  )

  ggplot(
    df,
    aes(
      log2FoldChange,
      -log10(padj),
      color = Significant
    )
  ) +
    geom_point(alpha = 0.7) +
    geom_vline(
      xintercept = c(-1, 1),
      linetype = "dashed"
    ) +
    geom_hline(
      yintercept = -log10(0.05),
      linetype = "dashed"
    ) +
    scale_color_manual(
      values = c("grey70", "red")
    ) +
    theme_classic() +
    ggtitle(title) #+
    #geom_text_repel(
      #data = label_genes,
      #aes(label = gene_id)
    #)
}

# june
plot_volcano(
  results_june,
  "CMT vs. DAF - June"
)

# july
plot_volcano(
  results_july,
  "CMT vs. DAF - July"
)

# august
plot_volcano(
  results_august,
  "CMT vs. DAF - August"
)
```

# Run DESeq2 Subset by Location
```{r}
# DAF
dds_daf <- dds_time[, colData(dds_time)$Location == "DAF"]
dds_daf$Month <- relevel(dds_daf$Month, ref = "June")
design(dds_daf) <- ~ Month
dds_daf <- DESeq(dds_daf)

# DAF July vs. June
results_daf_july_june <- results(
  dds_daf, contrast = c("Month", "July", "June")
)

# DAF August vs. June
results_daf_august_june <- results(
  dds_daf, contrast = c("Month", "August", "June")
)

# DAF August vs. July
results_daf_august_july <- results(
  dds_daf, contrast = c("Month", "August", "July")
)

# CMT
dds_cmt <- dds_time[, colData(dds_time)$Location == "CMT"]
dds_cmt$Month <- relevel(dds_cmt$Month, ref = "June")
design(dds_cmt) <- ~ Month
dds_cmt <- DESeq(dds_cmt)

# CMT July vs. June
results_cmt_july_june <- results(
  dds_cmt,
  contrast = c("Month", "July", "June")
)

# CMT August vs. June
results_cmt_august_june <- results(
  dds_cmt,
  contrast = c("Month", "August", "June")
)

# CMT August vs. July
results_cmt_august_july <- results(
  dds_cmt,
  contrast = c("Month", "August", "July")
)
```

# Pairwise Comparison Volcano Plots (between each month per location)
```{r}
# DAF July vs. June
plot_volcano(
  results_daf_july_june,
  "DAF - July vs. June"
)

# DAF August vs. July
plot_volcano(
  results_daf_august_july,
  "DAF - August vs. July"
)

# DAF August vs. June
plot_volcano(
  results_daf_august_june,
  "DAF - August vs. June"
)
```

```{r}
# CMT July vs. June
plot_volcano(
  results_cmt_july_june,
  "CMT - July vs. June"
)

# CMT August vs. July
plot_volcano(
  results_cmt_august_july,
  "CMT - August vs. July"
)

# CMT August vs. June
plot_volcano(
  results_cmt_august_june,
  "CMT - August vs. June"
)


```


# Generate lists of total significant upregulated and downregulated gene ids
```{r}
# NOTE: I am defining "significant" as padj < 0.05 and absolute value of log2FoldChange > 0

results_df <- as.data.frame(results)

# get list of upregulated gene ids
upregulated_genes <- subset(results_df, padj < 0.05 & log2FoldChange > 0)
upregulated_gene_ids <- rownames(upregulated_genes)
upregulated_gene_ids

# get list of downregulated gene ids
downregulated_genes <- subset(results_df, padj < 0.05 & log2FoldChange < 0)
downregulated_gene_ids <- rownames(downregulated_genes)
downregulated_gene_ids
```

# Counting upregulated and downregulated genes for each location and month
```{r}
# NOTE: upregulated means more expressed in CMT than DAF and downregulated means less expressed in CMT than DAF (aka more expressed in DAF than CMT)
# june
upregulated_june_counts <- sum(
  results_june$padj < 0.05 &
    results_june$log2FoldChange > 0,
  na.rm = TRUE
)

downregulated_june_counts <- sum(
  results_june$padj < 0.05 &
    results_june$log2FoldChange < 0,
  na.rm = TRUE
)

# july
upregulated_july_counts <- sum(
  results_july$padj < 0.05 &
    results_july$log2FoldChange > 0,
  na.rm = TRUE
)

downregulated_july_counts <- sum(
  results_july$padj < 0.05 &
    results_july$log2FoldChange < 0,
  na.rm = TRUE
)

# august
upregulated_august_counts <- sum(
  results_august$padj < 0.05 &
    results_august$log2FoldChange > 0,
  na.rm = TRUE
)

downregulated_august_counts <- sum(
  results_august$padj < 0.05 &
    results_august$log2FoldChange < 0,
  na.rm = TRUE
)

# put counts into table
table <- data.frame(
  Month = c("June", "July", "August"),
  Up_in_CMT = c(upregulated_june_counts, upregulated_july_counts, upregulated_august_counts),
  Up_in_DAF = c(downregulated_june_counts, downregulated_july_counts, downregulated_august_counts)
)
table
```

# Generate lists of significant upregulated and downregulated gene ids per month
```{r}
# june
upregulated_june_gene_ids <- rownames(results_june)[
  !is.na(results_june$padj) & # excludes the genes with NA padj values
    results_june$padj < 0.05 & 
    results_june$log2FoldChange > 0
]

downregulated_june_gene_ids <- rownames(results_june)[
  !is.na(results_june$padj) & # excludes the genes with NA padj values
    results_june$padj < 0.05 & 
    results_june$log2FoldChange < 0
]

upregulated_june_gene_ids
downregulated_june_gene_ids

# july
upregulated_july_gene_ids <- rownames(results_july)[
  !is.na(results_july$padj) & # excludes the genes with NA padj values
    results_july$padj < 0.05 & 
    results_july$log2FoldChange > 0
]

downregulated_july_gene_ids <- rownames(results_july)[
  !is.na(results_july$padj) & # excludes the genes with NA padj values
    results_july$padj < 0.05 & 
    results_july$log2FoldChange < 0
]

upregulated_july_gene_ids
downregulated_july_gene_ids

# august
upregulated_august_gene_ids <- rownames(results_august)[
  !is.na(results_august$padj) & # excludes the genes with NA padj values
    results_august$padj < 0.05 & 
    results_august$log2FoldChange > 0
]

downregulated_august_gene_ids <- rownames(results_august)[
  !is.na(results_august$padj) & # excludes the genes with NA padj values
    results_august$padj < 0.05 & 
    results_august$log2FoldChange < 0
]

upregulated_august_gene_ids
downregulated_august_gene_ids
```

# Environmental Data
```{r}
# load in all environmental data files for summer 2024
ph <- read_excel("/work/clh162/Data+/Sophie/2025-2026-team-code-sac165/pH_2024.xlsx")
salinity <- read_excel("/work/clh162/Data+/Sophie/2025-2026-team-code-sac165/Salinity_2024.xlsx")
do <- read_excel("/work/clh162/Data+/Sophie/2025-2026-team-code-sac165/DOsensor_2024.xlsx")
precip <- read_excel("/work/clh162/Data+/Sophie/2025-2026-team-code-sac165/Precip_2024.xlsx")

# standardize dates
ph$DateTime <- as.Date(ph$DateTime)
salinity$DateTime <- as.Date(salinity$DateTime)
do$DateTime <- as.Date(do$DateTime)
precip$DateTime <- as.Date(precip$DateTime)

# create variable for sample collection dates
collection_dates <- as.Date(c("2024-06-20", "2024-07-22", "2024-08-19"))
col_data$CollectionDate <- collection_dates
```

# Salinity
```{r}
# calculate 7-day average for the 7 days before the collection dates
salinity_avg <- salinity %>%
  group_by(Farm) %>%
  group_modify(~{
    data.frame(
      CollectionDate = collection_dates,
      Salinity_7d_Avg = sapply(
        collection_dates, 
        function(date){
          mean(.x$Sal_ppt[.x$DateTime >= date - 7 & .x$DateTime <= date], na.rm = TRUE)
        }
      )
    )
  }) %>%
  ungroup()

salinity_avg$Location <- recode(
  salinity_avg$Farm, "DUML" = "DAF", "CMAST" = "CMT"
)

salinity_avg

# add to environmental data table
env_data <- left_join(
  col_data,
  salinity_avg,
  by = c("Location", "CollectionDate"))

# calculate correlation coefficients
salinity_correlation <- sapply(
  top_month_genes,
  function(gene)
    cor(
      assay(vsd_time)[gene, ],
      env_data$Salinity_7d_Avg,
      method = "spearman"
    )
)
salinity_correlation

# create PCA plot for salinity and location
pcaData$Salinity <- env_data$Salinity_7d_Avg
pcaData$Location <- env_data$Location

ggplot(
  pcaData,
  aes(
    PC1,
    PC2,
    color = Salinity,
    shape = Location
  )
) +
  geom_point(size = 4) +
  theme_classic()
```

# Temperature
```{r}
# calculate 7-day average for the 7 days before the collection dates
temp_avg <- salinity %>%
  group_by(Farm) %>%
  group_modify(~{
    data.frame(
      CollectionDate = collection_dates,
      Temp_7d_Avg = sapply(
        collection_dates, 
        function(date){
          mean(.x$Temp_C[.x$DateTime >= date - 7 & .x$DateTime <= date], na.rm = TRUE)
        }
      )
    )
  }) %>%
  ungroup()

temp_avg$Location <- recode(
  temp_avg$Farm, "DUML" = "DAF", "CMAST" = "CMT"
)

# create environmental data table
env_data <- left_join(
  env_data,
  temp_avg,
  by = c("Location", "CollectionDate"))

# calculate correlation coefficients
temp_correlation <- sapply(
  top_month_genes,
  function(gene)
    cor(
      assay(vsd_time)[gene, ],
      env_data$Temp_7d_Avg,
      method = "spearman"
    )
)
temp_correlation

# create PCA plot for temperature and location
pcaData$Temp <- env_data$Temp_7d_Avg
pcaData$Location <- env_data$Location

ggplot(
  pcaData,
  aes(
    PC1,
    PC2,
    color = Temp,
    shape = Location
  )
) +
  geom_point(size = 4) +
  theme_classic()
```

# pH
```{r}
# calculate 7-day average for the 7 days before the collection dates
ph_avg <- ph %>%
  group_by(Farm) %>%
  group_modify(~{
    data.frame(
      CollectionDate = collection_dates,
      pH_7d_Avg = sapply(
        collection_dates, 
        function(date){
          mean(.x$pH[.x$DateTime >= date - 7 & .x$DateTime <= date], na.rm = TRUE)
        }
      )
    )
  }) %>%
  ungroup()

ph_avg$Location <- recode(
  ph_avg$Farm, "DUML" = "DAF", "CMAST" = "CMT"
)

ph_avg

# add to environmental data table
env_data <- left_join(
  env_data,
  ph_avg,
  by = c("Location", "CollectionDate"))

# calculate correlation coefficients
ph_correlation <- sapply(
  top_month_genes,
  function(gene)
    cor(
      assay(vsd_time)[gene, ],
      env_data$pH_7d_Avg,
      method = "spearman",
      use = "complete.obs"
    )
)
ph_correlation

# create PCA plot for ph and location
pcaData$pH <- env_data$pH_7d_Avg
pcaData$Location <- env_data$Location

ggplot(
  pcaData,
  aes(
    PC1,
    PC2,
    color = pH,
    shape = Location
  )
) +
  geom_point(size = 4) +
  theme_classic()
```
# Dissolved Oxygen
```{r}
# calculate 7-day average for the 7 days before the collection dates
do_avg <- do %>%
  group_by(Farm) %>%
  group_modify(~{
    data.frame(
      CollectionDate = collection_dates,
      DO_7d_Avg = sapply(
        collection_dates, 
        function(date){
          mean(.x$DomgL[.x$DateTime >= date - 7 & .x$DateTime <= date], na.rm = TRUE)
        }
      )
    )
  }) %>%
  ungroup()

do_avg$Location <- recode(
  do_avg$Farm, "DUML" = "DAF", "CMAST" = "CMT"
)

do_avg

# add to environmental data table
env_data <- left_join(
  env_data,
  do_avg,
  by = c("Location", "CollectionDate"))

# calculate correlation coefficients
do_correlation <- sapply(
  top_month_genes,
  function(gene)
    cor(
      assay(vsd_time)[gene, ],
      env_data$DO_7d_Avg,
      method = "spearman",
      use = "complete.obs"
    )
)
all(colnames(vsd_time) == env_data$Sample)
do_correlation

# create PCA plot for dissolved oxygen and location
pcaData$DO <- env_data$DO_7d_Avg
pcaData$Location <- env_data$Location

ggplot(
  pcaData,
  aes(
    PC1,
    PC2,
    color = DO,
    shape = Location
  )
) +
  geom_point(size = 4) +
  theme_classic()
```

# Environmental Data Correlations
```{r}
# table of correlation coefficients for environmental variables
correlation_table <- data.frame(
  Gene = names(temp_correlation),
  Temperature = temp_correlation,
  Salinity = salinity_correlation,
  DO = do_correlation,
  pH = ph_correlation,
  row.names = NULL
)
correlation_table
```
# Further Environmental Data Analysis - Temperature 
```{r}
# I am choosing to graph gene expression and temperature vs month on the same plot for the gene with the highest correlation coefficient (evm.TU.Chr04.3052)
gene <- "evm.TU.Chr04.3052"

plot_df <- data.frame(
  Date = env_data$CollectionDate,
  Location = env_data$Location,
  Expression = assay(vsd_time)[gene, ],
  Temperature = env_data$Temp_7d_Avg
)

ggplot(plot_df, aes(Date, color = Location)) +
  geom_point(aes(y = Expression), size = 3) +
  stat_summary(
    aes(y = Expression, group = Location, linetype = "Gene Expression"),
    fun = mean,
    geom = "line",
    linewidth = 1
  ) +
  geom_line(
    aes(y = Temperature, group = interaction(Location, "Temperature"), linetype = "Temperature"),
    linewidth = 1
  ) +
  geom_point(
    aes(y = Temperature),
    size = 3
  ) +
  scale_linetype_manual(
    name = "Variable",
    values = c(
      "Gene Expression" = "solid",
      "Temperature" = "twodash"
    )
  ) +
  theme_classic() +
  labs(
    y = "Value",
    title = gene
  )
```

# Further Environmental Data Analysis - Dissolved Oxygen
```{r}
# I am choosing to graph gene expression and temperature vs month on the same plot for the gene with the highest correlation coefficient (evm.TU.Chr04.3052)
gene <- "evm.TU.Chr04.3052"

plot_df <- data.frame(
  Date = env_data$CollectionDate,
  Location = env_data$Location,
  Expression = assay(vsd_time)[gene, ],
  DO = env_data$DO_7d_Avg
)

ggplot(plot_df, aes(Date, color = Location)) +
  geom_point(aes(y = Expression), size = 3) +
  stat_summary(
    aes(y = Expression, group = Location, linetype = "Gene Expression"),
    fun = mean,
    geom = "line",
    linewidth = 1
  ) +
  geom_line(
    aes(y = DO, group = interaction(Location, "DO"), linetype = "DO"),
    linewidth = 1
  ) +
  geom_point(
    aes(y = DO),
    size = 3
  ) +
  scale_linetype_manual(
    name = "Variable",
    values = c(
      "Gene Expression" = "solid",
      "DO" = "twodash"
    )
  ) +
  theme_classic() +
  labs(
    y = "Value",
    title = gene
  )
```

# Environmental Variables Correlation Matrix
```{r}
# create matrix of just environmental data
env_mat <- env_data[, c("Temp_7d_Avg", "Salinity_7d_Avg", "DO_7d_Avg", "pH_7d_Avg")]
rownames(env_mat) <- rownames(col_data)

correlation_matrix <- cor(env_mat, method = "spearman", use = "complete.obs")
correlation_matrix
```

# Yale 2025 Genome
```{r}
# combine all sample counts into one matrix with all 18 samples
count_25_dir <- "/work/clh162/henry/htseq_counts_CvYale2025"

# get all count files in the directory
count_files <- list.files(
  count_25_dir,
  pattern = "_counts\\.txt$",
  full.names = TRUE
)

# create count matrix 
count_25_matrix <- count_files %>%
  map(function(file) {
    sample_name <- sub(
      "_counts$",
      "",
      tools::file_path_sans_ext(basename(file))
    )
    read_tsv(file, col_names = c("gene_id", "count")) %>%
      rename(!!sample_name := count) # replace count with sample name
  }) %>%
  reduce(full_join, by = "gene_id") %>%
  replace(is.na(.), 0)

```


# Saving the Objects
```{r}
# save SummarizedExperiment
saveRDS(se, "/work/clh162/Data+/Sophie/2025-2026-team-code-sac165/Data+_team/Sophie/summarized_experiment.rds")

# save DESeqDataSet object
saveRDS(dds, "/work/clh162/Data+/Sophie/2025-2026-team-code-sac165/Data+_team/Sophie/dds_object.rds")

# save column data
saveRDS(col_data, "/work/clh162/Data+/Sophie/2025-2026-team-code-sac165/Data+_team/Sophie/col_data.rds")

# save environmental data matrix
saveRDS(env_mat, "/work/clh162/Data+/Sophie/2025-2026-team-code-sac165/Data+_team/Sophie/env_mat.rds")

# save pairwise comparisons (between each location per month)
saveRDS(dds_june, "/work/clh162/Data+/Sophie/2025-2026-team-code-sac165/Data+_team/Sophie/dds_june.rds")
saveRDS(dds_july, "/work/clh162/Data+/Sophie/2025-2026-team-code-sac165/Data+_team/Sophie/dds_july.rds")
saveRDS(dds_august, "/work/clh162/Data+/Sophie/2025-2026-team-code-sac165/Data+_team/Sophie/dds_august.rds")

# save pairwise comparisons (between each month per location)
saveRDS(dds_daf, "/work/clh162/Data+/Sophie/2025-2026-team-code-sac165/Data+_team/Sophie/dds_daf.rds")
saveRDS(dds_cmt, "/work/clh162/Data+/Sophie/2025-2026-team-code-sac165/Data+_team/Sophie/dds_cmt.rds")

```
